---
title: "Take-home Exercise 02: Geospatial Analytics for Public Good - Urban mobility"
author: "Yiqiong PAN"
date: "Oct 1, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 4

editor: visual
---

## Overview

### Background and Motivation

Urban mobility refers to the movement of people and goods across all transport modes within and between cities. With population growth and climate pressures, Cities are developing smart and sustainable transport systems to enhance accessibility and reduce carbon emissions.

During the progress, vast amounts of data are generated from technologies that track people using public transport such as buses, MRT, and roads from smart cards, GPS, and RFID systems. For instance, tap-on and tap-off data from smart cards capture trip flows from origins to destinations, and discounts are often offered for off-peak travel at the tap-off point.

Yet, such data are often used only for simple GIS mapping, rather than for deeper spatial analyses that could reveal richer insights into mobility patterns.

### Methods & Objective

Using **Local Measures of Spatial Autocorrelation (LMSA)** and **Emerging Hot Spot Analysis (EHSA)** , this exercise aims to analyse bus trips generated by origin at the hexagon level during **three** peak periods: 

- weekday morning (6 am – 9 am), 

- weekday afternoon (5 pm – 8 pm), 

- weekend/holiday (11 am – 8 pm).

Through the progress, our goals are to identify specific locations where values are significantly clustered or dispersed, and to reveal how hot and cold spots evolve over time. We hope to discover emerging travel trends for more responsive transport planning.

## Loading Packages in to R

There are **six** packages used in this exercise. `add more package`

| Package | Description |
|------------------------|------------------------------------------------|
| [sf](https://r-spatial.github.io/sf/) | Handles importing, managing, and processing vector-based geospatial data |
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data wrangling and visualisation, including **readr**, **dplyr**, and **ggplot2**. |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static and interactive thematic (choropleth) maps. |
| [knitr](https://cran.r-project.org/web/packages/knitr/index.html/) | Provides an elegant, flexible, and fast static table generation with R |
| [kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) | An extension of knitr for creating elegant html table with R |
| [DT](https://rstudio.github.io/DT/) | Provides an R interface to the JavaScript library DataTables for create interactive htnl tables. |
| [rvest](https://rvest.tidyverse.org/) | Scrapes or harvests data from web pages. |

: {tbl-colwidths="\[15,85\]"}

The code chunk below loads the required package into R.

```{r}
pacman::p_load(tidyverse, sf, sfdep, tmap, knitr, kableExtra, DT, rvest)
```

## Getting the Data into R environment

1.  Singapore boundary data, that we explored in Hands-on Ex 1-2

2.  LTADataMall: bus stop locations in SHP

3.  LTADataMall: the bus trip data for August 2025 were retrieved using GetPostman to download the CSV file.

### Importing the Geospatial Ploygon Data: Master Plan 2019 Subzone

The code chunk below imports Master Plan 2019 Subzone Boundary (No Sea) from Singapore’s open data portal.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

```{r}
extract_kml_field <- function(html_text, field_name) {
  # return NA if missing or empty
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)          #parse the HTML string 
  rows <- page %>% html_elements("tr")  #get all table rows <tr>
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>% #keep row where <th> table header text = field_name
    html_element("td") %>%                                       #take its <td> table data
    html_text2()                                                 #extract clean text
  
  if (length(value) == 0) NA_character_ else value        #return NA if no match, otherwise the value
}
```

```{r}
mpsz <- mpsz %>%
  mutate(
     #for each description string, call the function, It returns a length-1 character per row (or NA)
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>% #drop the raw KML attributes
  relocate(geometry, .after = last_col()) #move the geometry column to the end
```

```{r}
class(mpsz)
mpsz <- st_transform(mpsz, 3414)
st_crs(mpsz)
```

```{r}
write_rds(mpsz, "data/rds/mpsz.rds")

```

### Importing the Geospatial Point Data: Bus Stop Locations

The code chunk below imports *Bus Stop Location* shapefile downloaded from LTA DataMall into R environment.

```{r}
BusStop <- st_read(dsn = "data/LTADataMall",
                   layer = "BusStop") %>%
  st_transform(crs = 3414)
```

### Importing the Attribute Data: Bus Trips

The code chunk below imports *Passenger Volume by Origin Destination Bus Stops* downloaded from LTA DataMall into R environment.

```{r}
odbus <- read_csv("data/LTADataMall/origin_destination_bus_202508.csv")
summary(odbus)
```

### Visualisation of the Bus Stop Distribution in Singapore

Here we plot the bus stops in Singapore to see the distribution of their locations.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(BusStop), add = TRUE, cex = 0.5, col ="blue")
```

```{r}
#tmap_mode("view")
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop) +
  tm_dots()
```

```{r}
tmap_mode("plot")
```

Since there are a few points located outside the Singapore boundary, these need to be excluded. If we encounter an error message such as st_crs(x) != st_crs(y), it indicates that the coordinate reference systems (CRS) of the two objects are not the same. In such cases, we should assign the correct and consistent CRS (i.e., EPSG:3414). Both st_transform() and st_set_crs() can be used to apply the appropriate CRS.

### Extracting Bus Stops Located within Singapore

The code chunk below uses [`st_join()`](https://r-spatial.github.io/sf/reference/st_join.html) to select all bus stops within Singapore.

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz,
  join = st_within,
  left = FALSE)
```

```{r}
sum(duplicated(BusStop_in_SG$BUS_STOP_N))
```

```{r}
nrow(BusStop_in_SG)                 # total rows
length(unique(BusStop_in_SG$BUS_STOP_N))  # unique bus stops
```

```{r}
dp_busstops <-BusStop_in_SG %>%
  group_by(BUS_STOP_N) %>%
  filter(n()> 1) %>%
  arrange(BUS_STOP_N)
kable(dp_busstops)
```

There are **5** bus stops outside of Singapore.



```{r}
BusStop_SG <- st_join(
  BusStop, mpsz,
  join = st_within,
  left = TRUE)
```

```{r}
length(BusStop_SG$BUS_STOP_N) - length(BusStop_in_SG$BUS_STOP_N)
```

Next, we double-check that the points have been correctly filtered to include only those within Singapore.

```{r}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop_in_SG) +
  tm_dots()
```

### Creating the analytical hexagons (Singapore with Honeycomb Grid)

-   cellsize: A numeric value (length 1 or 2) specifying the target cell size. For square/rectangular cells, this defines width and height. For hexagonal cells, it defines the distance between opposite edges. So for the actual analysis, we should use `cellsize = 750`.

```{r}
hexagon <- st_make_grid(mpsz,
                        cellsize = 750,  ## attention 375m required for take-home ex.
                        what = "polygon",
                        square = FALSE) %>%
  st_sf()                                ## this st_* function doesn't return sf data.frame, we need to specifically coerce the file to sf object.

```

### Checking the hexagon layer visually

The code chunk below creates the hexagon grid with bus stop in Singapore.

```{r}
#| echo: false
tm_shape(hexagon) +
  tm_fill(col = "white", title = "Hexagons") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Analytical hexagon layer overplot on Singapore",
            main.title.position = "center",
            main.title.size = 1.0,
            legend.height = 0.35, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scale_bar(bg.color = "white", bg.alpha = 0.5) +
  tm_shape(mpsz) +
  tm_fill("green", title = "Singapore Boundary", alpha = 0.5) +
  tm_shape(BusStop_in_SG) +
  tm_dots(col = "red", size = 0.005, title = "Bus Stops")

```

We identified two tasks from the hexagon:

1.  In mpsz, the unique IDs SUBZONE_N or SUBZONE_C should be attached to the hexagon geometries.

2.  Keep only hexagons that contain at least one bus stop.

## Selecting hexagons with bus stops

### Filtering hexagons with bus stops

To implement this, we first filter the hexagons using a spatial intersection with the bus-stop points.

Recall: st_intersection() returns the geometric overlap (catchment), while st_intersects() returns a logical relation (TRUE/FALSE). After filtering, the hexagon count is reduced from 4,524 to 924.

```{r}
hexagon$busstop_count = 
  lengths(st_intersects(hexagon, BusStop_in_SG))
```

```{r}
hexagon <- filter(hexagon, busstop_count >0)
```

### Checking the hexagon layer visually

The plot confirms only the hexagon with bus stops retained.

```{r}
#| eval: false
#| echo: false
#| fig-width: 10
#| fig-height: 5
tm_shape(mpsz) +
  tm_fill("green", title = "Singapore Boundary", alpha = 0.5) +
  tm_shape(hexagon) +
  tm_fill(col = "white", title = "Hexagons", alpha = 1) +
  tm_borders(alpha = 0.2) +
  tm_layout(main.title = "Analytical hexagon layer overplot on Singapore",
            main.title.position = "center",
            main.title.size = 1.0,
            legend.height = 0.35, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scale_bar(bg.color = "white", bg.alpha = 0.5) +
  tm_shape(BusStop_in_SG) +
  tm_dots(col = "red", size = 0.001, title = "Bus Stops") +
  tm_grid(alpha = 0.2)
```

```{r}
tm_shape(mpsz) + tm_fill("green",fill_alpha =0.6) +
tm_shape(hexagon) + tm_polygons("grey90") + tm_borders("gray60") +
tm_shape(BusStop_in_SG) + tm_dots("black",  size = 0.1,shape = 19) + 
tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & LTADataMall",
            position = c("right", "bottom")) +
tm_title("Analytical Hexagon Grid with Bus Stops in Singapore", size = 1)
```

### Assigning ids to each hexagon

Here we have a quick look the hexagon dataset.

```{r}
kable(head(hexagon)) %>%
  kable_styling(font_size = 24)
```

The code chunk below removes the unwanted column and generates a set of 4-digit HEX_ID strings along with leading zero, then converts them from character to factor (ID) type using as.factor().

```{r}
#hexagon <- hexagon %>%
  #select(, -busstop_count)  #- to drop the column

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(nrow(hexagon))) %>%  #4d is to accommodate the hexagon number
  as.factor()  #ordinal scale

kable(head(hexagon))
```

```{r}
#tmap_mode("view")
tm_shape(hexagon) +
  tm_fill("HEX_ID")
tmap_mode("plot")
```



## Preparing Trip Generation Data

### Cleaning the data

Year_Month and destination_pt_code can be excluded from this analysis.


```{r}
# select the variables
trips <- odbus %>%
  select (c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%  # easier when performing join later
  rename(HOUR_OF_DAY= TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS) #easy interpretation
```




The BUS_STOP_N (post code of bus stop origination) is converted to factor since they are unique identifier and easier grouping group_by later.

```{r}
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N)
kable(head(trips))
```

### Populating Hexagon IDs into BusStop data frame

Next, we combine the hexagon, bus stop, and trip datasets.

This code creates a lookup table linking each bus stop BUS_STOP_N to the hexagon it belongs to HEX_ID

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID))

kable(head(bs_hex))
```



```{r}
bs_hex_unq <- bs_hex %>%
  distinct(BUS_STOP_N, .keep_all = TRUE)
```

```{r}
trips_unq<- inner_join(trips, bs_hex_unq)
kable(head(trips_unq))
```


### Adding HEX_ID into bus trips data

We merge the HEX_ID and Bus Stop into trip data

dups

```{r}
trips<- inner_join(trips, bs_hex)
kable(head(trips))
```






### Aggregating TRIPS based on HEX_ID

The code chunk below groups trip volume by **HEX_ID, DATE_TYPE, and HOUR_OF_DAY**.

#Duplicated trips
```{r}
#| eval: false
trips <- trips %>%
  group_by(
    HEX_ID,
    DAY_TYPE,
    HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS))





trips_wide_dups <- trips %>%
  group_by(
    HEX_ID,
    DAY_TYPE,
    HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS)) %>%
  ungroup() %>%
  pivot_wider(names_from = HOUR_OF_DAY,
              values_from = TRIPS)
  
colnames(trips_wide_dups)


peak_trips_dups <- trips_wide_dups %>%
  mutate(
    WEEKDAY_AM = if_else(
      DAY_TYPE == "WEEKDAY",
      rowSums(.[3:5], na.rm = TRUE),
      NA_real_
    ),
    WEEKDAY_PM = if_else(
      DAY_TYPE == "WEEKDAY",
      rowSums(.[14:16], na.rm = TRUE),
      NA_real_
    ),
    WEEKEND = if_else(
      DAY_TYPE %in% c("WEEKEND/HOLIDAY", "WEEKENDS/HOLIDAY"),
      rowSums(.[8:16], na.rm = TRUE),
      NA_real_)) %>%
  select(HEX_ID, WEEKDAY_AM, WEEKDAY_PM, WEEKEND) %>%
  group_by(HEX_ID) %>%
  summarise(
    WEEKDAY_AM = sum(WEEKDAY_AM, na.rm = TRUE),
    WEEKDAY_PM = sum(WEEKDAY_PM, na.rm = TRUE),
    WEEKEND = sum(WEEKEND, na.rm = TRUE))

```


## Filter Peak Periods

**three** peak periods: 

- weekday morning (6 am – 9 am), 

- weekday afternoon (5 pm – 8 pm), 

- weekend/holiday (11 am – 8 pm).

```{r}
trips_wide <- trips_unq %>%
  group_by(
    HEX_ID,
    DAY_TYPE,
    HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS)) %>%
  ungroup() %>%
  pivot_wider(names_from = HOUR_OF_DAY,
              values_from = TRIPS)
  
colnames(trips_wide)
```


```{r}
peak_trips <- trips_wide %>%
  mutate(
    WEEKDAY_AM = if_else(
      DAY_TYPE == "WEEKDAY",
      rowSums(.[3:5], na.rm = TRUE),
      NA_real_
    ),
    WEEKDAY_PM = if_else(
      DAY_TYPE == "WEEKDAY",
      rowSums(.[14:16], na.rm = TRUE),
      NA_real_
    ),
    WEEKEND = if_else(
      DAY_TYPE %in% c("WEEKEND/HOLIDAY", "WEEKENDS/HOLIDAY"),
      rowSums(.[8:16], na.rm = TRUE),
      NA_real_)) %>%
  select(HEX_ID, WEEKDAY_AM, WEEKDAY_PM, WEEKEND) %>%
  group_by(HEX_ID) %>%
  summarise(
    WEEKDAY_AM = sum(WEEKDAY_AM, na.rm = TRUE),
    WEEKDAY_PM = sum(WEEKDAY_PM, na.rm = TRUE),
    WEEKEND = sum(WEEKEND, na.rm = TRUE))

```



```{r}
kable(head(peak_trips, 20))
```

Bus stops 0000 (Changi Airport) and 00000 (Changi Bay) from the Bus Stop Locations dataset have no passenger tap-on data; the corresponding hex cell HEX_ID 834 was excluded after the join.
Similarly, 31029 (Western Water Catchment) and its corresponding hex cell HEX_ID 082 were excluded.
In total, the number of hex cells was reduced from 923???? to 824.


### The interactive trips data frame

```{r}
htmltools::div(              # creates a html wrapper to style the table
                             # sets font-size, 80% of page width, auto margin to center the table
  style = "font-size: 18px; width: 80%; margin: auto;",
                             # makes an interactive table ((search, sort, paginate))
  DT::datatable(             
    peak_trips,
    options = list(          # JavaScript options
      pageLength = 10,       # shows 10 rows per page
      scrollY = 300,         # gives the table a 500px-tall scrollable view port
      scrollX = TRUE         # enables horizontal scrolling
    ),
    class = "stripe hover compact"  #tightens spacing
  )
)
```



```{r}
peak_hex <- left_join(peak_trips,hexagon)
```

```{r}
write_rds(peak_hex, "data/rds/peak_hex.rds")
```


```{r}
peak_hex <-read_rds("data/rds/peak_hex.rds")
```

```{r}
peak_hex <- st_as_sf(peak_hex)
```


```{r}
sum(is.na(peak_hex[["WEEKDAY_AM"]]))
sum(is.na(peak_hex[["WEEKDAY_PM"]]))
sum(is.na(peak_hex[["WEEKEND"]]))
```
```{r}
peak_hex$WEEKDAY_AM <- as.numeric(as.factor(peak_hex$WEEKDAY_AM))
```

```{r}
ggplot(data = peak_hex,
       aes(x = "",
           y = WEEKEND)) +
  geom_boxplot()
```
```{r}
#| code-fold: true
#| code-summary: "View code"
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}


get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}

boxmap <- function(vnam, df, 
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  var <- as.vector(unlist(var))
  bb <- boxbreaks(var)
  tm_shape(mpsz) +
  tm_fill("gray",fill_alpha =0.5) +
  tm_shape(df) +
        tm_fill(vnam,
        fill.scale = tm_scale(
        breaks = bb,
        values = "-RdBu",         
        value.na = NA,            
        labels = c("lower outlier", 
                     "< 25%", 
                     "25% - 50%", 
                     "50% - 75%",
                     "> 75%", 
                     "upper outlier")),
      fill.legend = tm_legend(
        title = legtitle,        
        na.show = FALSE)) +
    tm_borders() +
    tm_layout(main.title = mtitle) +
    tm_compass(type="8star", size = 2) +
    tm_scale_bar() +
    tm_grid(alpha =0.2) +
    tm_credits("Source: data.gov.sg & LTADataMall",
            position = c("right", "bottom"))
}
```


```{r}
boxmap("WEEKDAY_AM", peak_hex,mtitle = "Box Map(Weekday Morning Peak)")
boxmap("WEEKDAY_PM", peak_hex,mtitle = "Box Map(Weekday Evening Peak)")
boxmap("WEEKEND", peak_hex,mtitle = "Box Map(Weekend/holiday Peak)")
```

```{r}
#| code-fold: true
#| code-summary: "View code"
get.var <- function(vname, df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}

percentmap <- function(vnam, df, legtitle = NA, mtitle = "Percentile Map") {
  percent <- c(0, .01, .1, .5, .9, .99, 1) # defines the breaks
  var <- get.var(vnam, df) # retrieve the variable 
  
  # Ensure var is a vector, not a data frame
  var <- as.vector(unlist(var))
  
  bperc <- quantile(var, percent, na.rm = TRUE) # calculates the percentile values
  
  tm_shape(mpsz) +
    tm_fill("gray",fill_alpha =0.5) +
    tm_shape(df) +
    tm_fill(vnam,
            fill.scale = tm_scale(
              breaks = bperc,
              values = "-RdBu",         
              value.na = NA,            
              labels = c("<1%", "1–10%", "10–50%", "50–90%", "90–99%", ">99%")),
            fill.legend = tm_legend(
              title = legtitle,        
              na.show = FALSE)) +
    tm_borders() +
    tm_layout(main.title = mtitle)+
    tm_compass(type="8star", size = 2) +
    tm_scale_bar() +
    tm_grid(alpha =0.2) +
    tm_credits("Source: data.gov.sg & LTADataMall",
            position = c("right", "bottom"))

}
```


```{r}
percentmap("WEEKDAY_AM", peak_hex,mtitle = "Percentiles Map(Weekday Morning Peak)")
percentmap("WEEKDAY_PM", peak_hex,mtitle = "Percentiles Map(Weekday Evening Peak)")
percentmap("WEEKEND", peak_hex,mtitle = "Percentiles Map(Weekend/holiday Peak)")

```

```{r}
percentmap("WEEKEND", peak_hex)

```


Two more steps are required before we begin the analysis:

Convert the data into a space–time cube to run EHSA on that cube.
