---
title: "04: Spatial Weights and Applications"
author: "Yiqiong PAN"
date: "Sep 15, 2025"
date-modified: "last-modified"
execute:
  echo: true #display the code
  eval: true #run the code
  message: false
  warning: false
  freeze: true # not render if nothing edited
editor: visual
format: 
  html:
    number-sections: true
    number-depth: 4
---

## Overview

In this exercise, we apply spatial weights in a practical implementation in R.

Spatial data are rarely independent which often violates the independence assumptions of classical statistics (e.g., regression or ANOVA).Properly defining and weighting neighbours is crucial to account for spatial autocorrelation and avoid biased outcomes. Through it, we can understand who influences whom and how strong the influence is.

## The Study Area and Data

The following two data sets are used here:

1.  Geospatial data: Hunan county boundary layer in `shapefile` format

2.  Aspatial data: `Hunan_2012.csv` for Hunan’s local development indicators in 2012.

### Getting started

A total of **five** R packages will be used in this exercise.

| Package | Description |
|------------------------------------|------------------------------------|
| [sf](https://r-spatial.github.io/sf/) | Simple Features, a new R package which handles importing, managing, and processing vector-based geospatial data. |
| [spdep](https://cran.r-project.org/web/packages/spdep/index.html) | Tools for computing spatial weights and calculating spatially lagged variables. |
| [knitr](https://cran.r-project.org/web/packages/knitr/index.html) | Provides useful functions for dynamic document generation. |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static or interactive choropleth maps. |
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data import, cleaning, transformation, and visualisation (e.g., readr, dplyr, tidyr, ggplot2). |

: {tbl-colwidths="\[15,85\]"}

After installation via `install.packages()`, we load them into R environment using the code below.

```{r}
pacman:: p_load(sf, spdep, knitr, tmap, tidyverse)
```

## Importing Data into R

In this section, we import above-mentioned two data sets into the R environment by using `st_read()` from `sf` package for geospatial data and `read_csv()` from `readr` of `tiyeverse` family.

### Importing `sharefile` File

The code chunk below uses `st_read()` to import the Hunan county layer in `sharefile` format to R as an `sf` object for spatial analysis.

```{r}
hunan <- st_read(dsn = "data/geospatial",
                layer = "Hunan")

glimpse(hunan)

tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.4)
```

### Importing `csv` File

The code chunk uses read_csv() of `readr` to import the attribute table of local development indicators in 2012.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
glimpse(hunan2012)
class(hunan2012)
```

### Performing Relational Join

The code chunk uses `left_join()` of `dplyr` package from `tidyverse` to append the attribute table of `hunan` the spatial polygon data frame with the attribute fields of `hunan2012` data frame `by = join_by(County)` and selects required fields.

```{r}
hunan <- left_join(hunan, hunan2012) %>%
   select(1:4, 7, 15)
```

## Visualising Reginal Development Indicator

Next we use `tm_polygons()` to create a base map and use `qtm()` also from `tmap` to create a choropleth map to show the distribution of GDP per capita GDPPC 2012.

```{r, fig.height=12, fig.width=16}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp = 1, ncol =2)
```

## Computing Contiguity Spatial Weights

We define the neighbours (who influences whom) using: Contiguity(shared border), Distance(within k kilometres), Trajectory: order or flow.

We weight the neighbours (how strong the influence is) using: Binary (1/0), Distance decay (closer neighbours carry more weight), Strength-based (combined measures, such as shared boundary length, population, or flows).

## Computing Distance Based Neighbours

## Weights Based on IDW (Inversed Distance Method)

## Row-standardised Weights Matrix

## Application of Spatial Weight Matrix

## Reference

Kam, T. S. 8 Spatial Weights and Applications. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap08
