---
title: "04: Spatial Weights and Applications"
author: "Yiqiong PAN"
date: "Sep 15, 2025"
date-modified: "last-modified"
execute:
  echo: true #display the code
  eval: true #run the code
  message: false
  warning: false
  freeze: true # not render if nothing edited
editor: visual
format: 
  html:
    number-sections: true
    number-depth: 4
    code-link: true
    toc: true
---

## Overview

In this exercise, we apply spatial weights in a practical implementation in R.

Spatial data are rarely independent which often violates the independence assumptions of classical statistics (e.g., regression or ANOVA).Properly defining and weighting neighbours is crucial to account for spatial autocorrelation and avoid biased outcomes. Through it, we can understand who influences whom and how strong the influence is.

## The Study Area and Data

The following two data sets are used here:

1.  Geospatial data: Hunan county boundary layer in `shapefile` format

2.  Aspatial data: `Hunan_2012.csv` for Hunan’s local development indicators in 2012.

### Getting started

A total of **five** R packages will be used in this exercise.

| Package | Description |
|------------------------------------|------------------------------------|
| [sf](https://r-spatial.github.io/sf/) | Simple Features, a new R package which handles importing, managing, and processing vector-based geospatial data. |
| [spdep](https://cran.r-project.org/web/packages/spdep/index.html) | Tools for computing spatial weights and calculating spatially lagged variables. |
| [knitr](https://cran.r-project.org/web/packages/knitr/index.html) | Provides useful functions for dynamic document generation. |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static or interactive choropleth maps. |
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data import, cleaning, transformation, and visualisation (e.g., readr, dplyr, tidyr, ggplot2). |

: {tbl-colwidths="\[15,85\]"}

After installation via `install.packages()`, we load them into R environment using the code below.

```{r}
pacman:: p_load(sf, spdep, knitr, tmap, tidyverse)
```

## Importing Data into R

In this section, we import above-mentioned two data sets into the R environment by using `st_read()` from `sf` package for geospatial data and `read_csv()` from `readr` of `tiyeverse` family.

### Importing `sharefile` File

The code chunk below uses `st_read()` to import the Hunan county layer in `sharefile` format to R as an `sf` object for spatial analysis.

```{r}
hunan <- st_read(dsn = "data/geospatial",
                layer = "Hunan")

glimpse(hunan)

tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.3)
```

### Importing `csv` File

The code chunk uses read_csv() of `readr` to import the attribute table of local development indicators in 2012.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
glimpse(hunan2012)
class(hunan2012)
```

### Performing Relational Join

The code chunk uses `left_join()` of `dplyr` package from `tidyverse` to append the attribute table of `hunan` the spatial polygon data frame with the attribute fields of `hunan2012` data frame `by = join_by(County)` and selects required fields.

```{r}
colnames(hunan)
```

```{r}
colnames(hunan2012)
```

```{r}
hunan_jn <- left_join(hunan, hunan2012) 
colnames(hunan_jn)

```

```{r}
hunan<- hunan_jn %>%
   select(1:4, 7, 15)

glimpse(hunan)
hunan
```

## Visualising Reginal Development Indicator

Next we use `tm_polygons()` to create a base map and use `qtm()` also from `tmap` to create a choropleth map to show the distribution of GDP per capita GDPPC 2012.

```{r, fig.height=7, fig.width=14}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp = 1, ncol =2)
```

::: callout-note
Quick summary:

The side-by-side maps show that the distribution of GDP per capita (GDPPC) across Hunan Province is not evenly spread. The majority of counties record values below 60K. The darkest-shaded area is Changsha, which has the highest GDPPC, ranging from 80K to 100K, with income gradually decreasing as one moves outward from the capital circle. Outside of this core, two other counties stand out with GDPPC between 60K and 80K: Lengshuikeng, located in central Hunan, and Zixing in the southeast.
:::

## Computing Contiguity Spatial Weights

Neighbours could be defined by using: Contiguity (shared border).

(Insee & Eurostat, 2018) Contiguity-defined neighbourhoods are widely used in demographic and social studies. Several types of contiguity can be distinguished. In **Rook** contiguity, two neighbourhoods are considered adjacent if they share at least two common boundary points (i.e., a full edge). On the other hand, **Queen** contiguity requires only one common boundary point (e.g., an edge or a vertex). As a result, the number of neighbours defined under Queen contiguity is always greater than or equal to those defined under Rook contiguity.

Next, we use `poly2nb()` of **spdep** package to calculate contiguity weights metrics for the study area `hunan`.The function allows us to specify the type of contiguity by setting the `queen` argument to either `TRUE` (Queen contiguity) or `FALSE` (Rook contiguity).


::: panel-tabset

### (QUEEN) Contiguity Based Neighbours

The code chunk below returns **QUEEN** contiguity weight matrix.

From the **summary statistics**, we learn that the dataset contains 88 counties, with a total of 448 neighbour connections across all regions. Among these, percentage of actual neighbour links relative to all possible links is 5.79%. On average, each county has about five neighbours.

Looking at the link number distribution, we see that region ID 85 is the most connected, with 11 neighbours, while region ID 30 and 65 are the least connected, each having only one neighbour. Most counties have between four and six neighbours, which is consistent with the average of five links per region.

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

By using the index code chunk below, we can print the list of neighbour ploygons for most-connected region 85.The numbers represent the polygon IDs as stored in hunan `SpatialPolygonsDataFrame` class.

Further, we can retrieve the `county name` of region 85 by looking it up in the hunan dataset, specifically using the syntax `dataset$column[i]`, which returns the corresponding value in that column.

We can also retrieve 11 neighbouring county names by using the`c()` function to concatenate the required region IDs inside the indexing operation.

```{r}
wm_q[[85]] 
class(wm_q[[85]])
hunan$County[85]
hunan$NAME_3[c(1,  2,  3,  5,  6, 32, 56, 57, 69, 75, 78)]
```

We can also wrap both steps together once we get familiar with the data frames.

```{r}
hunan$NAME_3[wm_q[[85]]]
```

To simplify the code and reduce the risk of syntax errors, we can first assign the neighbour indices of a given region to a helper variable then retieve the corresponding GDPPC values.

```{r}
nb1 <- wm_q[[1]]
hunan$GDPPC[nb1]
```

Additionally, we can also use the structure function `str()` to display the complete matrix.

```{r}
str(wm_q)
```



### (ROOK) Contiguity Based Neighbours

As mentioned earlier, the measurement type can be easily switched to **Rook** mode by disabling the `queen` argument. The code chunk below uses `poly2nb()` to compute the Rook contiguity weight matrix.

```{r}
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```
The summary shows that there are 88 regions in the dataset, with a total of 440 possible neighbour connections. Among these, percentage of actual neighbour links relative to all possible links is 5.68%. On average, each region has around five neighbours. Similarly, one region is the most connected, with 10 Rook neighbours, while regions 30 and 65 are the least connected, each having only one link.

The code chunk below retreives the region ID 85's neighbour names and GDPPC under ROOK.

We used the `setdiff(a, b)` function to identify the neighbours included under Queen contiguity but not under Rook. The result returned region ID **57**, named **Nan**. It is unclear whether this indicates a missing value. However, the county is located at the northeast vertex of region 85 (Taoyuan) according to the baseline map.

```{r}
wm_r[[85]]
hunan$County[85]
hunan$NAME_3[wm_r[[85]]]
hunan$GDPPC[wm_r[[85]]]

setdiff(hunan$NAME_3[wm_q[[85]]], hunan$NAME_3[wm_r[[85]]]) 
setdiff(wm_q[[85]], wm_r[[85]]) 

```

### Prepreration for Visualisation


A connectivity graph shows polygons as points and draws lines to their neighbours. To build this, we need a point for each polygon, usually the centroid.

Using `sf`, we calculate centroids, then extract longitude and latitude into two separate `numeric` objects. This is done by mapping `st_centroid` over the geometry column (`us.bound`) with `map_dbl` from purrr.

- Longitude: access with `[[1]]`

- Latitude: access with `[[2]]`

The above gives numeric vectors of coordinates for use in graph construction.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
class(longitude)
latitude  <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```


Next we use cbind() to combine longitude and latitude and create a `matrix` object.

```{r}
coords <- cbind(longitude, latitude)

head(coords)
```



::: callout-note
**Syntax rules on \[\]**

-   **Single** brackets **\[\]** return a sublist (still of type list).

```{r}
wm_q[1]
class(wm_q[1])
```

-   **Double** brackets **\[\[ \]\]** return the contents of that element.

```{r}
wm_q[[1]]
class(wm_q[[1]])
```
:::



### Mapping QUEEN & ROOK & Comparsion

The code chunk below use plot() from base R to create two maps one for **QUEEN** contiguity based and one for **ROOK** contiguity based. It also overlays two sets of maps for **comparison**.

```{r, fig.width=10, fig.height=6}

par(mfrow = c(1,2))

plot(hunan$geometry, border = "lightgrey",
     main = "QUEEN Contiguity") # countries base map
plot(wm_q, coords, # centroid coords of each polygon
     pch = 19, # solid circles
     cex = 0.6, # shrinks point size to 60%
     add = TRUE, # overlays this coords with base map
     col = "red") # connecting line in red



plot(hunan$geometry, border = "lightgrey",
     main = "ROOK Contiguity") # countries base map
plot(wm_r, coords, # centroid coords of each polygon
     pch = 19, # solid circles
     cex = 0.6, # shrinks point size to 60%
     add = TRUE, # overlays this coords with base map
     col = "red") # connecting line in red

```

```{r, fig.width=10, fig.height=6}
#| echo: false
plot(hunan$geometry, border = "lightgrey",
     main = "Contiguity Comparison") # countries base map
plot(wm_q, coords, # centroid coords of each polygon
     pch = 19, # solid circles
     cex = 0.6, # shrinks point size to 60%
     add = TRUE, # overlays this coords with base map
     col = "BLUE") # connecting # countries base map
plot(wm_r, coords, # centroid coords of each polygon
     pch = 19, # solid circles
     cex = 0.6, # shrinks point size to 60%
     add = TRUE, # overlays this coords with base map
     col = "red") # connecting line in red

legend("bottomleft",
       legend = c("Rook & Queen links", "Queen-only links"),
       col    = c("red", "blue"),
       lty    = 1, lwd = 2, bty = "n")
```

Here we visualise `setdiff()` differences in connectivity between Queen and Rook contiguity.

The comparison map above highlights the additional **4** neighbour links (equally unidirectional 8 nonzero entries), captured by Queen in blue.

The most connected region, Taoyuan (in the north), has 11 neighbours under Queen and 10 under Rook. One corner-touching neighbour is excluded in the Rook method due to its definition. Similar differences can also be observed in the lower northeast part of the province, where Queen contiguity identifies additional vertex-based neighbours that Rook does not.



:::

## Computing distance based neighbours

## Weights Based on IDW (Inversed Distance Method)

## Row-standardised Weights Matrix

## Application of Spatial Weight Matrix


We define the neighbours (who influences whom) using: Contiguity(shared border), Distance(within k kilometres), Trajectory: order or flow.

We weight the neighbours (how strong the influence is) using: Binary (1/0), Distance decay (closer neighbours carry more weight), Strength-based (combined measures, such as shared boundary length, population, or flows).

## Reference

Insee, & Eurostat. (2018, October). Handbook of spatial analysis: Theory and application with R (V. Loonis, Dir.; M.-P. de Bellefon, Coord.; Insee Méthodes, No. 131). Insee. https://www.insee.fr/en/information/3635545

Kam, T. S. 8 Spatial Weights and Applications. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap08
