---
title: "In-class Ex06 "
author: "Yiqiong PAN"
date: "Oct 3, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 4

editor: visual
---

## Overview
side note: use knitr to general reports.

## Getting started

### Loading Packages in to R

```{r}
pacman::p_load(sf,tmap,tidyverse,rvest, knitr, sfdep)
```

### Getting the Data into R environment

1. Singapore boundary data, that we explored in Hands-on Ex 1-2

2. LTADataMall: bus stop locations in SHP

3. LTADataMall: bus trips during Aug 2025. We untilise the getpostman to download the csv file.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

```{r}
extract_kml_field <- function(html_text, field_name) {
  # return NA if missing or empty
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)          #parse the HTML string 
  rows <- page %>% html_elements("tr")  #get all table rows <tr>
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>% #keep row where <th> table header text = field_name
    html_element("td") %>%                                       #take its <td> table data
    html_text2()                                                 #extract clean text
  
  if (length(value) == 0) NA_character_ else value        #return NA if no match, otherwise the value
}
```


```{r}
mpsz <- mpsz %>%
  mutate(
     #for each description string, call the function, It returns a length-1 character per row (or NA)
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>% #drop the raw KML attributes
  relocate(geometry, .after = last_col()) #move the geometry column to the end
```


```{r}
write_rds(mpsz, "data/rds/mpsz.rds")

```

```{r}
class(mpsz)
mpsz <- st_transform(mpsz, 3414)
st_crs(mpsz)
```

```{r}
BusStop <- st_read(dsn = "data/LTADataMall",
                   layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
odbus <- read_csv("data/LTADataMall/origin_destination_bus_202508.csv")
summary(odbus)
```


Here we plot the bus stops in Singapore to see the distribution of their locations.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(BusStop), add = TRUE, cex = 0.5, col ="blue")
```
```{r}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop) +
  tm_dots()
```
Since there are a few points outside the Singapore boundaries, we need to exclude these. If we see the error messages that st_crs(x) != st_crs(y), it reminds us to assign the correct/same projection crs (i.e. 3414). (t_transform() or st_set_crs both work the same way to pass on the crs.

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz,
  join = st_within,
  left = FALSE)
```

Next we double check we have filtered the points in Singapore.

```{r}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop_in_SG) +
  tm_dots()
```
### Creating the analytical hexagons (Singapore with Honeycomb Grid)

- cellsize: A numeric value (length 1 or 2) specifying the target cell size. For square/rectangular cells, this defines width and height. For hexagonal cells, it defines the distance between opposite edges. 

```{r}
hexagon <- st_make_grid(mpsz,
                        cellsize = 700,  ## attention 375m required for take-home ex.
                        what = "polygon",
                        square = FALSE) %>%
  st_sf()                                ## this st_* function doesn't return sf data.frame, we need to specifically coerce the file to sf object.

```

Two Tasks:
- In mpsz, there are two unique IDs that are SUBZONE_N and SUBZONE_C. We need to add the reference to the geometry values.
- Als owe only want hexagons with bus stops.


We filter the hexgaons with bus stops.

Recall: 

- St_intersections() returns the geographic catchment of the intersection. 

- St_intersects() returns true/false (1/0) of the intersection. 

We reduce the hexagon number from 4524 to 924. 

```{r}
hexagon$n_collisions = 
  lengths(st_intersects(hexagon, BusStop_in_SG))
```

```{r}
hexagon <- filter(hexagon, n_collisions >0)
```


```{r}
hexagon <- hexagon %>%
  select(, -n_collisions)  #- to drop the column

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(nrow(hexagon))) %>%  #04 is to accommodate the hexagon number
  as.factor()  #ordinal scale

kable(head(hexagon))
```


### Tidying the trip data

Year_Month and destination_pt_code could be excluded for this analysis. 

```{r}
# select the variables
trips <- odbus %>%
  select (c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%  # easier when performing join later
  rename(HOUR_OF_DAY= TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS) #easy interpration
``` 

```{r}
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N)
kable(head(trips))
```


Next we combine the hexagon with bus stops and trips together.

Hexagon and bus stops

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID))

kable(head(bs_hex))
```

We add HEX_ID into bus trips.

```{r}
trips<- inner_join(trips, bs_hex)
kable(head(trips))
```

Note: two more steps to go before we can start the analysis.
convert time-space cube..... for EHSA. before doing any analysis.



