---
title: "In-class Exercise 6a: Take-Home Exercise 2 Kick Starter"
author: "Yiqiong PAN"
date: "Oct 3, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 4

editor: visual
---

## Overview

In the exercise, we learn how to

-   download dynamic data by using LTA DataMall API and [postman](https://www.postman.com/);

-    import and tidy geospatial data using sf and tidyverse;

-   import and tidy aspatial data using tidyverse;

-   create analytic hexagon data using sf

## Loading Packages in to R

There are **six** packages used in this exercise.

| Package | Description |
|----------------------|--------------------------------------------------|
| [sf](https://r-spatial.github.io/sf/) | Handles importing, managing, and processing vector-based geospatial data |
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data wrangling and visualisation, including **readr**, **dplyr**, and **ggplot2**. |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static and interactive thematic (choropleth) maps. |
| [knitr](https://cran.r-project.org/web/packages/knitr/index.html/) | Provides an elegant, flexible, and fast static table generation with R |
| [kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) | An extension of knitr for creating elegant html table with R |
| [DT](https://rstudio.github.io/DT/) | Provides an R interface to the JavaScript library DataTables for create interactive htnl tables. |
| [rvest](https://rvest.tidyverse.org/) | Scrapes or harvests data from web pages. |

: {tbl-colwidths="\[15,85\]"}

The code chunk below loads the required package into R.

```{r}
pacman::p_load(tidyverse, sf, sfdep, tmap, knitr, kableExtra, DT, rvest)
```

## Getting the Data into R environment

1.  Singapore boundary data, that we explored in Hands-on Ex 1-2

2.  LTADataMall: bus stop locations in SHP

3.  LTADataMall: the bus trip data for August 2025 were retrieved using GetPostman to download the CSV file.

### Importing the Geospatial Ploygon Data: Master Plan 2019 Subzone

The code chunk below imports Master Plan 2019 Subzone Boundary (No Sea) from Singapore’s open data portal.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

```{r}
extract_kml_field <- function(html_text, field_name) {
  # return NA if missing or empty
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)          #parse the HTML string 
  rows <- page %>% html_elements("tr")  #get all table rows <tr>
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>% #keep row where <th> table header text = field_name
    html_element("td") %>%                                       #take its <td> table data
    html_text2()                                                 #extract clean text
  
  if (length(value) == 0) NA_character_ else value        #return NA if no match, otherwise the value
}
```

```{r}
mpsz <- mpsz %>%
  mutate(
     #for each description string, call the function, It returns a length-1 character per row (or NA)
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>% #drop the raw KML attributes
  relocate(geometry, .after = last_col()) #move the geometry column to the end
```


```{r}
class(mpsz)
mpsz <- st_transform(mpsz, 3414)
st_crs(mpsz)
```

```{r}
write_rds(mpsz, "data/rds/mpsz.rds")

```

### Importing the Geospatial Point Data: Bus Stop Locations

The code chunk below imports *Bus Stop Location* shapefile downloaded from LTA DataMall into R environment.

```{r}
BusStop <- st_read(dsn = "data/LTADataMall",
                   layer = "BusStop") %>%
  st_transform(crs = 3414)
```

### Importing the Attribute Data: Bus Trips

The code chunk below imports *Passenger Volume by Origin Destination Bus Stops* downloaded from LTA DataMall into R environment.

```{r}
odbus <- read_csv("data/LTADataMall/origin_destination_bus_202508.csv")
summary(odbus)
```

### Visualisation of the Bus Stop Distribution in Singapore

Here we plot the bus stops in Singapore to see the distribution of their locations.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(BusStop), add = TRUE, cex = 0.5, col ="blue")
```

```{r}
tmap_mode("view")
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop) +
  tm_dots()
```
```{r}
tmap_mode("plot")
```


Since there are a few points located outside the Singapore boundary, these need to be excluded. If we encounter an error message such as st_crs(x) != st_crs(y), it indicates that the coordinate reference systems (CRS) of the two objects are not the same. In such cases, we should assign the correct and consistent CRS (i.e., EPSG:3414). Both st_transform() and st_set_crs() can be used to apply the appropriate CRS.

### Extracting Bus Stops Located within Singapore

The code chunk below uses [`st_join()`](https://r-spatial.github.io/sf/reference/st_join.html) to select all bus stops within Singapore.

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz,
  join = st_within,
  left = FALSE)
```

Next, we double-check that the points have been correctly filtered to include only those within Singapore.

```{r}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop_in_SG) +
  tm_dots()
```

**Question**:

Do you know how many bus stops are located outside of Singapore main island? Describe how the answer was derives.

```{r}
BusStop_SG <- st_join(
  BusStop, mpsz,
  join = st_within,
  left = TRUE)
```

```{r}
length(BusStop_SG$BUS_STOP_N) - length(BusStop_in_SG$BUS_STOP_N)
```

There are **5** bus stops outside of Singapore.


### Creating the analytical hexagons (Singapore with Honeycomb Grid)

-   cellsize: A numeric value (length 1 or 2) specifying the target cell size. For square/rectangular cells, this defines width and height. For hexagonal cells, it defines the distance between opposite edges. So for the actual analysis, we should use `cellsize = 750`.

```{r}
hexagon <- st_make_grid(mpsz,
                        cellsize = 750,  ## attention 375m required for take-home ex.
                        what = "polygon",
                        square = FALSE) %>%
  st_sf()                                ## this st_* function doesn't return sf data.frame, we need to specifically coerce the file to sf object.

```


### Checking the hexagon layer visually

The code chunk below creates the hexagon grid with bus stop in Singapore.

```{r}
#| echo: false
tm_shape(hexagon) +
  tm_fill(col = "white", title = "Hexagons") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Analytical hexagon layer overplot on Singapore",
            main.title.position = "center",
            main.title.size = 1.0,
            legend.height = 0.35, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scale_bar(bg.color = "white", bg.alpha = 0.5) +
  tm_shape(mpsz) +
  tm_fill("green", title = "Singapore Boundary", alpha = 0.5) +
  tm_shape(BusStop_in_SG) +
  tm_dots(col = "red", size = 0.005, title = "Bus Stops")

```

We identified two tasks from the hexagon:

1.  In mpsz, the unique IDs SUBZONE_N or SUBZONE_C should be attached to the hexagon geometries.

2.  Keep only hexagons that contain at least one bus stop.


## Selecting hexagons with bus stops

### Filtering hexagons with bus stops

To implement this, we first filter the hexagons using a spatial intersection with the bus-stop points.

Recall: st_intersection() returns the geometric overlap (catchment), while st_intersects() returns a logical relation (TRUE/FALSE). After filtering, the hexagon count is reduced from 4,524 to 924.

```{r}
hexagon$busstop_count = 
  lengths(st_intersects(hexagon, BusStop_in_SG))
```

```{r}
hexagon <- filter(hexagon, busstop_count >0)
```


### Checking the hexagon layer visually

The plot confirms only the hexagon with bus stops retained. 

```{r}
#| fig-width: 10
#| fig-height: 5
tm_shape(mpsz) +
  tm_fill("green", title = "Singapore Boundary", alpha = 0.5) +
  tm_shape(hexagon) +
  tm_fill(col = "white", title = "Hexagons", alpha = 1) +
  tm_borders(alpha = 0.2) +
  tm_layout(main.title = "Analytical hexagon layer overplot on Singapore",
            main.title.position = "center",
            main.title.size = 1.0,
            legend.height = 0.35, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scale_bar(bg.color = "white", bg.alpha = 0.5) +
  tm_shape(BusStop_in_SG) +
  tm_dots(col = "red", size = 0.001, title = "Bus Stops") +
  tm_grid(alpha = 0.2)
```


Plot the hexagons.

```{r}
tm_shape(mpsz) + tm_polygons("green",fill_alpha =0.9) +
tm_shape(hexagon) + tm_polygons("grey90") + tm_borders("gray60") +
tm_shape(BusStop_in_SG) + tm_dots("black",  size = 0.1,shape = 19) + 
tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & LTADataMall",
            position = c("right", "bottom")) +
tm_title("Honeycomb Grid with Bus Stops in Singapore", size = 1)
```


### Assigning ids to each hexagon

Here we have a quick look the hexagon dataset.

```{r}
kable(head(hexagon)) %>%
  kable_styling(font_size = 24)
```



The code chunk below removes the unwanted column and generates a set of 4-digit HEX_ID strings along with leading zero, then converts them from character to factor (ID) type using as.factor().

```{r}
hexagon <- hexagon %>%
  select(, -busstop_count)  #- to drop the column

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(nrow(hexagon))) %>%  #4d is to accommodate the hexagon number
  as.factor()  #ordinal scale

kable(head(hexagon))
```

## Preparing Trip Generation Data

### Cleaning the data

Year_Month and destination_pt_code can be excluded from this analysis.

```{r}
# select the variables
trips <- odbus %>%
  select (c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%  # easier when performing join later
  rename(HOUR_OF_DAY= TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS) #easy interpration
```

The BUS_STOP_N (post code of bus stop origination) is converted to factor since they are unique identifier and easier grouping group_by later.

```{r}
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N)
kable(head(trips))
```

### Populating Hexagon IDs into BusStop data frame

Next, we combine the hexagon, bus stop, and trip datasets.

This code creates a lookup table linking each bus stop BUS_STOP_N to the hexagon it belongs to HEX_ID

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID))

kable(head(bs_hex))
```

### Adding HEX_ID into bus trips data

We merge the HEX_ID and Bus Stop into trip data.

```{r}
trips<- inner_join(trips, bs_hex)
kable(head(trips))
```

### Aggregating TRIPS based on HEX_ID

The code chunk below groups trip volume by **HEX_ID, DATE_TYPE, and HOUR_OF_DAY**.

```{r}
trips <- trips %>%
  group_by(
    HEX_ID,
    DAY_TYPE,
    HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS))

kable(head(trips))
```
### The interactive trips data frame

```{r}
htmltools::div(              # creates a html wrapper to style the table
                             # sets font-size, 80% of page width, auto margin to center the table
  style = "font-size: 18px; width: 80%; margin: auto;",
                             # makes an interactive table ((search, sort, paginate))
  DT::datatable(             
    trips,
    options = list(          # JavaScript options
      pageLength = 10,       # shows 10 rows per page
      scrollY = 500,         # gives the table a 500px-tall scrollable view port
      scrollX = TRUE         # enables horizontal scrolling
    ),
    class = "stripe hover compact"  #tightens spacing
  )
)
```



Two more steps are required before we begin the analysis:

Convert the data into a space–time cube to run EHSA on that cube.
