---
title: "In-class Ex06 "
author: "Yiqiong PAN"
date: "Oct 3, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 4

editor: visual
---

## Overview

## Getting started

### Loading Packages in to R

```{r}
pacman::p_load(sf,tmap,tidyverse,rvest, knitr, sfdep)
```

Side note: use knitr to general reports.


### Getting the Data into R environment

1. Singapore boundary data, that we explored in Hands-on Ex 1-2

2. LTADataMall: bus stop locations in SHP

3. LTADataMall: the bus trip data for August 2025 were retrieved using GetPostman to download the CSV file.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

```{r}
extract_kml_field <- function(html_text, field_name) {
  # return NA if missing or empty
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)          #parse the HTML string 
  rows <- page %>% html_elements("tr")  #get all table rows <tr>
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>% #keep row where <th> table header text = field_name
    html_element("td") %>%                                       #take its <td> table data
    html_text2()                                                 #extract clean text
  
  if (length(value) == 0) NA_character_ else value        #return NA if no match, otherwise the value
}
```


```{r}
mpsz <- mpsz %>%
  mutate(
     #for each description string, call the function, It returns a length-1 character per row (or NA)
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>% #drop the raw KML attributes
  relocate(geometry, .after = last_col()) #move the geometry column to the end
```


```{r}
write_rds(mpsz, "data/rds/mpsz.rds")

```

```{r}
class(mpsz)
mpsz <- st_transform(mpsz, 3414)
st_crs(mpsz)
```

```{r}
BusStop <- st_read(dsn = "data/LTADataMall",
                   layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
odbus <- read_csv("data/LTADataMall/origin_destination_bus_202508.csv")
summary(odbus)
```


Here we plot the bus stops in Singapore to see the distribution of their locations.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(BusStop), add = TRUE, cex = 0.5, col ="blue")
```
```{r}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop) +
  tm_dots()
```
Since there are a few points located outside the Singapore boundary, these need to be excluded. If we encounter an error message such as st_crs(x) != st_crs(y), it indicates that the coordinate reference systems (CRS) of the two objects are not the same. In such cases, we should assign the correct and consistent CRS (i.e., EPSG:3414). Both st_transform() and st_set_crs() can be used to apply the appropriate CRS.

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz,
  join = st_within,
  left = FALSE)
```

Next, we double-check that the points have been correctly filtered to include only those within Singapore.
```{r}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop_in_SG) +
  tm_dots()
```
### Creating the analytical hexagons (Singapore with Honeycomb Grid)

- cellsize: A numeric value (length 1 or 2) specifying the target cell size. For square/rectangular cells, this defines width and height. For hexagonal cells, it defines the distance between opposite edges. So for the actual analysis, we should use `cellsize = 750`.

```{r}
hexagon <- st_make_grid(mpsz,
                        cellsize = 750,  ## attention 375m required for take-home ex.
                        what = "polygon",
                        square = FALSE) %>%
  st_sf()                                ## this st_* function doesn't return sf data.frame, we need to specifically coerce the file to sf object.

```

We identified two tasks from the hexagon:

1. In mpsz, the unique IDs SUBZONE_N or SUBZONE_C should be attached to the hexagon geometries.

2. Keep only hexagons that contain at least one bus stop.

To implement this, we first filter the hexagons using a spatial intersection with the bus-stop points. 

Recall: st_intersection() returns the geometric overlap (catchment), while st_intersects() returns a logical relation (TRUE/FALSE). After filtering, the hexagon count is reduced from 4,524 to 924.

```{r}
hexagon$n_collisions = 
  lengths(st_intersects(hexagon, BusStop_in_SG))
```

```{r}
hexagon <- filter(hexagon, n_collisions >0)
```

The code chunk below removes the unwanted column and generates a set of 4-digit HEX_ID strings with leading zeros, then converts them from character to factor (ID) type using as.factor().

```{r}
hexagon <- hexagon %>%
  select(, -n_collisions)  #- to drop the column

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(nrow(hexagon))) %>%  #4d is to accommodate the hexagon number
  as.factor()  #ordinal scale

kable(head(hexagon))
```

### Tidying the trip data

Year_Month and destination_pt_code can be excluded from this analysis.

```{r}
# select the variables
trips <- odbus %>%
  select (c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%  # easier when performing join later
  rename(HOUR_OF_DAY= TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS) #easy interpration
``` 

The BUS_STOP_N (post code of bus stop origination) is converted to factor since they are unique identifier and easier grouping group_by later.
```{r}
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N)
kable(head(trips))
```


Next, we combine the hexagon, bus stop, and trip datasets.

This code creates a lookup table linking each bus stop BUS_STOP_N to the hexagon it belongs to HEX_ID

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID))

kable(head(bs_hex))
```

We merge the HEX_ID and Bus Stop into trip data.
```{r}
trips<- inner_join(trips, bs_hex)
kable(head(trips))
```


Plot the hexagons.

```{r}
tm_shape(hexagon) + tm_borders(col = "grey60") +
tm_shape(mpsz) + tm_borders(col = "red", lwd = 2) +
tm_layout(title = "Hexagons within Singapore Boundary")
```


Two more steps are required before we begin the analysis:


Convert the data into a space–time cube to run EHSA on that cube.





