---
title: "06: Geographical Segmentation with Spatially Constrained Clustering Techniques"
author: "Yiqiong PAN"
date: "Sep 30, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 4

editor: visual
---

## Overview

In this exercise, we use methods like **Hierarchical Cluster Analysis** (HCA) and Spatially Constrained Cluster Analysis (SCCA) including (**SKATER and ClusGeo**) to group neighbouring spatial units into homogeneous regions (areas that are similar to each other in terms of variables, i.e. geographically referenced **multivariate**).

### Learning Outcome

The tasks to be completed are to:

-   convert simple feature data.frame into R's SpatialPolygonDataFrame object using `as_Spatial()` of **sf** package

-   perform cluster analysis with `hclust()` of Base R

-   perform spatially constrained cluster analysis using `skater()` of Base R

-   visualise the analysis using **ggplot2** and **tmap** package.

## Getting Started

### The Analytical Question

In real world application likes geobusiness and geospatail related policy making, regions are defined by statistical similarity in the multivariate data. In the case study of [Shan State](https://en.wikipedia.org/wiki/Shan_State) [Myanmar](https://en.wikipedia.org/wiki/Myanmar), we find and outline the clusters of areas that share similar characteristics, through multiple **Information and Communication technology** (ICT) measures, namely: Radio, Television, Land line phone, Mobile phone, Computer, and Internet at home.

## The Data

| Dataset | Description | Format |
|-------------------------|--------------------------|----------------------|
| myanmar_township_boundaries | GIS Myanmar Township Boundary Data represented as polygon features | ESRI shapefile |
| Shan-ICT.csv | 2014 Myanmar Population and Housing Census at the township level | CSV |

The data can be downloaded from [Myanmar Information Management Unit (MIMU)](http://themimu.info/).

### Installing and Loading R Packages

There are **14** packages used in this exercise.

-   Spatial data (**sf, spdep, tmap**)

-   Cluster analysis (**ClustGeo, cluster, NbClust, factoextra**)

-   Multivariate exploration (**corrplot, heatmaply, psych, GGally**)

-   Visualisation / general workflow (**tidyverse, ggpubr**)

| Package | Description |
|-------------------|-----------------------------------------------------|
| [sf](https://r-spatial.github.io/sf/) | Handles importing, managing, and processing vector-based geospatial data |
| [rgdal](https://cran.r-project.org/package=rgdal) | Provides bindings to the GDAL/OGR library for reading/writing spatial data formats. |
| [spdep](https://cran.r-project.org/package=spdep) | Tools for spatial dependence: building neighbour lists, spatial weights, and spatial autocorrelation analysis. |
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data wrangling and visualization, including **readr**, **dplyr**, and **ggplot2**. |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static and interactive thematic (choropleth) maps. |
| [corrplot](https://cran.r-project.org/package=corrplot) | Provides graphical displays of correlation matrices for multivariate data exploration. |
| [ggpubr](https://cran.r-project.org/package=ggpubr) | Simplifies publication-ready plots based on **ggplot2** with statistical annotations. |
| [heatmaply](https://cran.r-project.org/package=heatmaply) | Generates interactive heatmaps for exploring multivariate data relationships. |
| [cluster](https://cran.r-project.org/package=cluster) | Functions for cluster analysis, including hierarchical, partitioning, and silhouette methods. |
| [ClustGeo](https://cran.r-project.org/package=ClustGeo) | Performs spatially constrained clustering by balancing attribute similarity with geographical contiguity. |
| [NbClust](https://cran.r-project.org/package=NbClust) | Determines the optimal number of clusters using multiple indices and methods. |
| [facoextra](https://cran.r-project.org/package=factoextra) | Visualisation tools for clustering and multivariate analyses (e.g., PCA, CA, HCPC). |
| [GGally](https://ggobi.github.io/ggally/) | Extension of ggplot2 with functions for pair plots, network plots, and correlation plots. |
| [psych](https://cran.r-project.org/package=psych) | rocedures for psychometric and multivariate analysis, including factor analysis. |

The code chunk below loads the packages into R environment.

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally, knitr)
```

## Data Import and Preparation

### Importing Geospatial Data into R Environment

The code chunk below uses[`st_read()`](https://www.rdocumentation.org/packages/sf/versions/0.7-2/topics/st_read) from **sf** package to import Mayanmar Township Boundary (`myanmar_township_boundaries`) in shapefile format into R as a simple feature data.frame.

```{r}
shan_sf <- st_read(dsn = "data/geospatial",
                   layer = "myanmar_township_boundaries") %>%
  filter (ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>% # filter Shan State
  select(c(2:7))    # select columns on ST, DT, and TS
```

```{r}
shan_sf
```

The `sf` object follows Hardy Wickham’s [tidyverse](https://edzer.github.io/rstudio_conf/#1) principles, so we can handle it like a tidy tibble (rows = observations, cols = variables), with an extra geometry column.

```{r}
glimpse(shan_sf)
```

### Importing Aspatial Data into R Environment

The chunk below imports ICT variables via `read_csv()` function of **readr** package as a **tibble** data.frame. The data are from **The 2014 Myanmar Population and Housing Census Myanmar**.

```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```

The code chunk below summrises the statistics from attribute table. There are a total of eleven fields and 55 observation in the tibble data.frame.

```{r}
summary(ict)
```

### Derive Bew Variables Using dplyr Package

The code chunk below derives the **penetration rate of each ICT variable** based on total household number in township. We should always transform absolute values to **relative measures** before doing spatial analysis or clustering, otherwise the results will be dominated by township size.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

Let's have a quick check on the new talbe.

```{r}
summary(ict_derived)
```

New columns, RADIO_PR, TV_PR, LLPHONE_PR, MPHONE_PR, COMPUTER_PR, and INTERNET_PR,are created as planned.

## Exploratory Data Analysis (EDA)

### EDA Using Statistical Graphics

To examine the variables such as the number of households with radio, we can use graphical methods that highlight different aspects of the data distribution:

**Histogram**

-   Shows the overall shape of the distribution.

-   Helps identify whether the data is left-skewed, right-skewed, or approximately normal.

-   Useful for understanding the frequency of values and detecting clustering or gaps in the data.

```{r,fig.width= 4, fig.height=3}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

**Boxplot**:

-   Summarises the distribution using the five-number summary (minimum, Q1, median, Q3, maximum).

-   Makes it easy to detect outliers and understand the spread and central tendency of the data.

-   Allows quick comparison across multiple variables or groups (e.g., by township).

```{r,fig.width= 4, fig.height=3}
ggplot(data=ict_derived, 
       aes(x= RADIO )) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Let's also visualise the distribution of the newly derived variables (i.e. Radio penetration rate).

```{r,fig.width= 4, fig.height=3}
ggplot(data = ict_derived,
       aes(x = RADIO_PR)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
```

```{r,fig.width= 4, fig.height=3}
ggplot(data = ict_derived,
       aes(x = RADIO_PR)) +
  geom_boxplot(color = "black", fill = "light blue")
```

::: callout-note
-   What can you observed from the distributions reveal in the histogram and boxplot?

The penetration rate data are more normally distributed and show fewer outliers than the raw ownership counts, since the effect of township size has been normalised.
:::

In the figure below, **multiple** histograms are plotted to reveal the distribution of the selected variables in the `ict_derived` data.frame.

In **tmap**, `tmap_arrange()` (or `tmarange`) is used to arrange multiple map plots side by side or in a grid layout.

In **ggplot2**, the equivalent is [`ggarrange()`](https://rpkgs.datanovia.com/ggpubr/reference/ggarrange.html) from the [**ggpubr**](https://rpkgs.datanovia.com/ggpubr/index.html) package.

```{r, fig.width=10}
#| code-fold: true
#| code-summary: "View code"

radio <- ggplot(data= ict_derived,
                aes(x = RADIO_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")

tv <- ggplot(data = ict_derived, aes(x= TV_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")

llphone <- ggplot(data = ict_derived, aes(x = LLPHONE_PR)) +
  geom_histogram(bins = 20, color = "black",fill = "light blue")

mphone <- ggplot(data = ict_derived, aes(x = MPHONE_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")

computer <- ggplot(data = ict_derived, aes(x = COMPUTER_PR)) +
  geom_histogram(bins = 20, color = "black", fill ="light blue") 

internet <- ggplot(data = ict_derived, aes(x = INTERNET_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")

ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3,
          nrow = 2)
```

### EDA Using Choropleth Map

#### Joining Geospatial Data with Aspatial Data

Previously we use the **ggplot2** to visualise the distribution and outliers of the ICT variables in the tibble data.frame `ict_derived`. To incorporate the statistical data with corresponding township locations, we now use [`left_join()`](https://dplyr.tidyverse.org/reference/mutate-joins.html) function from the **dplyr** package to combine the two data sets ( `shan_sf` as an `sf` data.frame and `ict_derived`) based on the primary key **'TS_PCODE'**. We then plot a choropleth map of the selected variable distribution using the `qtm()` function from the **tmap** package. The code chunk below performs the above-mentioned first step.

```{r}
shan_sf <- left_join(shan_sf, ict_derived, by = c("TS_PCODE" = "TS_PCODE"))
write_rds(shan_sf, "data/rds/shan_sf.rds") #save for further analysis
```

```{r}
shan_sf  <- read_rds("data/rds/shan_sf.rds")
```

#### Preparing a Choropleth Map

The code chunk below uses the `qtm()` function from the **tmap** package to create a choropleth map that visualises the spatial distribution of radio penetration rates across Shan State townships.

```{r}
qtm(shan_sf, "RADIO_PR")
```

Next let's visualise why it is important to transform absolute values to **relative measures** before performing spatial analysis. Otherwise the results could be biased. The difference becomes clear when we comparing the distribution of the total number households across townships (TT_HOUSEHOLDS.map) with the distribution of households owning Radio (Radio.map).

The code chunk plot two choropleth maps and arrange them side by side.

```{r, fig.height=6, fig.width=10}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill("TT_HOUSEHOLDS", # colour or numeric or categorical variable in `shp` 
          n = 5,
          style = "jenks",  # preferred number of classes for numeric 'col' 
          title = "Total households", textNA = NA) +
   tm_layout(legend.position = c("right", "top")) +
  tm_borders(alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill( "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ",textNA = NA) + 
  tm_layout(legend.position = c("right", "top")) +
  tm_borders(alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map, 
             asp=1, ncol=2)  #Aspect ratio of each map
```

```{r}
any(is.na(shan_sf))
```

```{r, fig.height=6, fig.width=10}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks", textNA = NA) +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_layout(outer.margins=0, asp=0)
```

::: callout-note
-   Can you identify the differences?

Although the penetration rate map partly mirrors the distribution of total households, radio ownership does not scale one-to-one with township population. Other factors (e.g., income, market access etc) influence adoption, so using **relative measures** such as penetration rate reduces size bias and provides a more accurate basis for clustering and spatial comparisons than raw counts.
:::

::: callout-tip
-   Jenks: Reveals natural clusters in skewed data by minimising within-class variance.

Tip: Great for one-off maps; for comparisons, compute once and use fixed breaks.

-   Equal interval: Splits the value range into equal-width bins. simple and consistent.

Tip: Best when data are fairly uniform or when you need fixed policy thresholds.

-   Quantile: Puts an equal number of features in each class, highlighting ranks/percentiles.

Tip: Ensures balanced colours; watch for tiny value differences flipping classes.

-   Pretty: Uses rounded, human-friendly breakpoints for clean legends.

Tip: Ideal for communication; double-check that the bins still reflect the data distribution.
:::

## Correlation Analysis

Before running cluster analysis, we should check correlations between variables. If a pair is highly correlated (e.g., \|r\| \>= 0.8), one of the pair should be dropped to avoid redundancy. The code chunk below uses [`corrplot.mixed()`](https://cran.r-project.org/web/packages/corrplot/corrplot.pdf) function from [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) package to visualise the correlations.

```{r}
cluster_vars.cor = cor(ict_derived[,12:17]) #corr matrix
corrplot.mixed(cluster_vars.cor,  # use mixed method to visualise the matrix
               lower = "ellipse", # visualisation methods for lower triangular 
               upper = "number", ## visualisation methods for upper triangular 
               tl.pos = "lt",  #text label position: lt means left and top, d diagonal, n means no.
               diag ="l", #glyph in diagonal: n/l/u means no/lower/upper 
               tl.col = "black")
```

We observe that `INTERNET_PR` is highly correlated with `COMPUTER_PR` (r = 0.87), so we will drop one of them from the cluster analysis to avoid redundancy.

## Hiearachy Cluster Anlaysis

The general workflow of hierarchy cluster analysis consists of:

1.  Extracting clustering variables (screen for NAs & high correlations). `st_set_geometry(NULL)`; `select()`; `row.names()`

2.  Standardising the variables (min-max/ z-scores). `normalize()`; `scale()`

3.  Computing a proximity (distance) matrix (often Euclidean). `dist()`

4.  Running hierarchical clustering (choose linkage, e.g., ward.D2). `hclust()`;`agnes()`

5.  Determining the optimal number of clusters (dendrogram + silhouette/gap/NbClust). `clusGap(FUN = hcut)`

6.  Interpreting the dendrogram & profile clusters (cluster means, maps, tables).`rect.hclust()`

7.  Visualisation (dendrogram, heatmap, spatial choropleths). `heatmaply()`, `cutree()`, `cbind()`, `qtm()`

### Extracting Clustering Variables

The code chunk below extracts the **5** clustering variables (excluding INTERNET_PR) from `shan_sf` sf object to a new data.frame.

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>% # or st_drop_geometry() removes the geometry
select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR") # INTERNET_PR  is omitted
head(cluster_vars,10)
summary(cluster_vars)
class(cluster_vars)
```

The code chunk sets the row names to **township names** so those labels automatically flow through to new tables and new maps.

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars, 10)
```

The code chunk below deletes the TS.x field.

```{r}
shan_ict <-select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

### Data Standardisation

In cluster analysis, multiple variables often have different scales. To prevent variables with larger magnitudes from dominating the results, we should standardise the inputs before clustering.

#### Min-Max Standardisation

In the code chunk below, `normalize()` from the [**heatmaply**](https://cran.r-project.org/web/packages/heatmaply/index.html) package is used to normalise the clustering variables via the min–max method. We then use summary() to display summary statistics for the normalised variables.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

The values of the min-max standardised clustering variables range from 0-1.

#### Z-score Standardisation

Z-score standardisation can be done with base R’s scale(). The code below standardises the clustering variables.

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

After transformation, each variable has mean 0 and standard deviation 1. We use [`describe()`](https://www.rdocumentation.org/packages/Hmisc/versions/4.4-0/topics/describe) of [**pysch**](https://cran.r-project.org/web/packages/psych/index.html) package rather than base `summary()` because it reports the standard deviation.

::: callout-warning
**Note on assumptions**:

We standardise variables using z-scores to place them on a common scale. This does not assume normality; however, for highly skewed or outlier-prone variables, consider a transformation (log/Yeo–Johnson) or robust scaling (median/MAD) before or instead of z-scaling.
:::

#### Visualising the Standardised Clustering Varaibles

The code chunk below visualises the distribution of the clustering variable **RADIO_PR** before and after standardisation.

```{r, fig.width=12, fig.height=4}
r <- ggplot(data=ict_derived, aes(x= `RADIO_PR`)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue") +
  ggtitle("RAW values")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s<- ggplot(data = shan_ict_s_df, aes(x= `RADIO_PR`)) +
    geom_histogram(bins = 20, color = "black", fill = "light blue") +
  ggtitle("Min-max standardisation")


shan_ict_z_df <- as.data.frame(shan_ict.z)
z<- ggplot(data = shan_ict_z_df, aes(x= `RADIO_PR`)) +
    geom_histogram(bins = 20, color = "black", fill = "light blue") +
  ggtitle("Z-score standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

```{r}
names(shan_ict[1:5])
```

The code chunk below generates the before-and-after distribution for each ICT clustering variables.

```{r, fig.width=12, fig.height=20}
#| code-fold: true
#| code-summary: "View code"

shan_ict_s_df <- as.data.frame(shan_ict.std)
shan_ict_z_df <- as.data.frame(shan_ict.z)

plot_histograms <- function(col) {
  r <- ggplot(ict_derived, aes(x = .data[[col]])) +
    geom_histogram(bins = 20, color = "black", fill = "lightblue") +
    ggtitle(paste("Raw:", col))
  
  s <- ggplot(shan_ict_s_df, aes(x = .data[[col]])) +
    geom_histogram(bins = 20, color = "black", fill = "lightblue") +
    ggtitle(paste("Min-max:", col))
  
  z <- ggplot(shan_ict_z_df, aes(x = .data[[col]])) +
    geom_histogram(bins = 20, color = "black", fill = "lightblue") +
    ggtitle(paste("Z-score:", col))

  ggarrange(r, s, z, ncol = 3, nrow = 1)}


# for (col in names(shan_ict) {print(plot_histograms(col))}

plots <- lapply(names(shan_ict), plot_histograms) # list
ggarrange(plotlist = plots, ncol = 1, nrow = length(plots))
```

::: callout-note
**What statistical conclusion can you draw from the histograms above?**

The histograms compare raw, min–max scaled, and z-score scaled values for each variable. standardisation only changes the measurement units, not the distribution shapes.
:::

```{r, fig.width=12, fig.height=4}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

### Computing Proximity Matrix

In R, several packages provide functions to calculate distance matrices. Here, we will use the [`dist()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/dist.html) function of base R to compute a proximity matrix.

The function does not calculate geographic distance (like latitude–longitude great-circle distance).

Instead, **it calculates mathematical distances in feature space**, based on the values of the clustering variables (columns).

::: callout-note
dist(x, method = "euclidean", diag = FALSE, upper = FALSE, p = 2)

-   x is a numeric matrix, data frame or dist object.
-   p is the power of the Minkowski distance.
-   6 methods include **euclidean, maximum, manhattan, canberra, binary, and minkowski**, with euclidean being the default.
:::

The code chunk demonstrates how to generate a proximity matrix using the euclidean method.

```{r}
proxmat <- dist(shan_ict,method="euclidean")
```

```{r}
summary(proxmat)
```

```{r echo=TRUE, eval=FALSE}
proxmat
```

### Computing Hierachical Clustering

In R, several packages provide functions for hierarchical clustering. We use the [`hclust()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/hclust.html) function from base R to build clusters.

Thefunction applies an agglomerative approach (bottom up).

::: callout-note
hclust(d, method = "complete", members = NULL)

-   d is a **dist** object

-   method supports eight clustering methods: ward.D, ward.D2, single, complete, average (UPGMA), mcquitty (WPGMA), median (WPGMC), and centroid (UPGMC).
:::

The code chunk below computes hierarchical clustering using the **ward.D** method. The results are stored in an object of class **hclust**, which contains the tree structure (**dendrogram**) generated by the clustering process.

```{r}
hclust_ward <- hclust(proxmat, method = "ward.D")
hclust_ward$labels[1]
hclust_ward$height[1]
```

The code chunk creates the tree by using *plot()* of R Graphics.

```{r fig.width=10, fig.height=6}
plot(hclust_ward, cex = 0.9) #shrink by 10%
```

### Selecting the Optimal Clustering Algorithm

One challenge in hierarchical clustering is determining the stronger cluster structure. Unlike hclust(), the [`agnes()`](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/agnes) function from the [**cluster**](https://cran.r-project.org/web/packages/cluster/) package directly returns the agglomerative coefficient (AC). When AC closer to 1, it indicates a stronger clustering structure.

::: callout-note
agnes(x, diss = inherits(x, "dist"), metric = "euclidean", stand = FALSE, method = "average", par.method, keep.diss = n \< 100, keep.data = !diss, trace.lev = 0) agnes: short for **Agglomerative Nesting.**
:::

Below, we compute the agglomerative coefficient for several linkage methods and compare them.

```{r}
m <- c( "average", "single", "complete", "ward")        # creates a vector of methods
names(m) <- c( "average", "single", "complete", "ward") # gives each element a name

ac <- function(x) {
  agnes(shan_ict, method = x)$ac  #$ac extracts ac
}

map_dbl(m, ac) # applies the function ac() to each element of m.
```

### Determining Optimal Clusters

Another technical challenge is deciding on the optimal number of clusters to retain. Choosing too few clusters may oversimplify the data, while too many clusters may lead to overfitting and poor interpretability.

Three commonly used methods to determine the appropriate number of clusters are:

[**Elbow Method**](https://en.wikipedia.org/wiki/Elbow_method_(clustering))

-   Plots the total within-cluster sum of squares (WSS) against the number of clusters.

-   The “elbow” point, where the rate of decrease sharply slows, indicates a suitable number of clusters.

[**Average Silhouette Method**](https://www.sciencedirect.com/science/article/pii/0377042787901257?via%3Dihub)

-   Computes the silhouette coefficient, which measures how similar an observation is to its own cluster compared to other clusters.

-   The number of clusters with the highest average silhouette value is considered optimal.

[**Gap Statistic Method**](https://hastie.su.domains/Papers/gap.pdf)

-   Compares the total within-cluster variation for k different numbers of clusters with the expected values under a null reference distribution.

-   The optimal number of clusters is the one that maximises the gap statistic.

#### Gap Statistic Method

The code chunk below uses [`clusGap()`](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/clusGap) of **cluster**(https://cran.r-project.org/web/packages/cluster/index.html) package to compute the gap statistics. The [`hcut()`](https://rpkgs.datanovia.com/factoextra/reference/hcut.html) function used is from [**factoextra**](https://rpkgs.datanovia.com/factoextra/index.html) package.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, #hcut or kmeans
                    nstart = 25, #random starts for kmeans
                    K.max = 10, 
                    B = 50)  #bootstrap
# Print the result
print(gap_stat, method = "firstmax") # common ones: first(local)max, globalmax, tibs2001SEmax(first peak)
```

The code chunk below visualises the outcome by using [`fviz_gap_stat()`](https://rpkgs.datanovia.com/factoextra/reference/fviz_nbclust.html) of [**factoextra**](https://rpkgs.datanovia.com/factoextra/index.html) package.

```{r}
fviz_gap_stat(gap_stat)
```

::: callout-important
**Interpreting the Gap Statistic**

-   Although the highest Gap value appears at k = 1, retaining only one cluster is not meaningful, as clustering is intended to partition data into multiple groups. A more practical choice is k = 6 where the Gap value reaches its next highest point.

**Additional Methods**

Beyond the Elbow, Silhouette, and Gap Statistic methods, another powerful tool is the [**NbClust**](https://cran.r-project.org/web/packages/NbClust/) package (Charrad et al., 2014). This package provides access to 30 different indices for determining the optimal number of clusters.
:::

```{r}
set.seed(12345)

nb_hclust <- NbClust(
  data = shan_ict,        # data all numeric, no categorical columns)
  distance = "euclidean", # same as for hcut
  min.nc = 2,             # minimum number of clusters to consider
  max.nc = 10,            # maximum number of clusters to consider
  method = "ward.D2",     # hierarchical clustering linkage
  index = "all"           # or pick e.g. "silhouette", "gap", etc.
)

nb_hclust

```

```{r}
nb_hclust$Best.nc
```

While NbClust favoured 2 clusters, the Gap Statistic pointed to 6 clusters. Considering interpretability and the need for meaningful subgroup distinctions, the 6-cluster solution was selected for subsequent analysis.

### Interpreting the Dendrograms

In the dendrogram, each leaf represents an observation. Moving upward, similar observations are joined into branches, which then merge at higher levels. The vertical axis shows the fusion height, reflecting dissimilarity: lower values indicate greater similarity, while higher values mean less similarity. Similarity can only be inferred from the height at which branches first merge, not from their horizontal positions.

```{r fig.width=10, fig.height=6}
plot(hclust_ward, cex = 0.9) #shrink by 10%
```

We use [`rect.hclust()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/rect.hclust.html) of R stats to draw the border boxes. `border = 2:5` means: to use **red, green, blue, and cyan** as the rectangle borders.

```{r fig.width=10, fig.height=6}
plot(hclust_ward, cex = 0.9)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

### Visually-driven Hierarchical Clustering Analysis

In this section, we use the [**heatmaply**](https://cran.r-project.org/web/packages/heatmaply/index.html) package to create visually driven hierarchical clustering (both interactive and static).

#### Transforming the Data Frame into a Matrix

The code chunk below transforms the `shan_ict` data from into a data matrix.

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

#### Plotting Interactive Cluster Heatmap using heatmaply()

The code chunk below uses [`heatmaply()`](https://talgalili.github.io/heatmaply/reference/heatmaply.html) [**heatmaply**](https://cran.r-project.org/web/packages/heatmaply/index.html) package to create interactive cluster heatmap.

```{r, fig.width=8, fig.height=6}
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

### Mapping the Clusters Formed

As six clusters appear appropriate, we use [*cutree()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/cutree.html) of Base R to derive a 6-cluster model.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
groups
```

To visualise the clustering results, this list must be appended to the `shan_sf` simple feature object. The process can be carried out in three steps:

-   Convert the groups list into a matrix.

-   Use `cbind()` to attach this matrix to `shan_sf`, producing a new simple feature object called `shan_sf_cluster`.

-   Apply the `rename()` function from the **dplyr** package to change the column name from `as.matrix.groups` to `CLUSTER`.

Subsequently, the `qtm()` function from the **tmap** package is used to generate a choropleth map that displays the cluster results. The map shows that the clusters are highly fragmented, which highlights a key limitation of non-spatial clustering methods such as hierarchical cluster analysis.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

```{r}
qtm(shan_sf_cluster, "CLUSTER")
```

## Spatially Constrained Clustering: SKATER approach

Next we use [`skater(https://r-spatial.github.io/spdep/reference/skater.html)`]() from [**spdep**](https://r-spatial.github.io/spdep/index.html) to find spatially constrained cluster.

The `skater()` stands for Spatial 'K'luster Analysis by Tree Edge Removal.

### Converting into SpatialPolygonsDataFrame

The code chunk below uses [`as_Spatial()'](https://r-spatial.github.io/sf/reference/coerce-methods.html) to coerce`shan_sf`as`sf`into Spatial*Polygons*DataFrame which can be processed by`skater()\`function later.

```{r}
shan_sp <- as_Spatial(shan_sf)
```

### Computing Neighbour List

The code chunk below uses [`ploy2nb()`](https://r-spatial.github.io/spdep/reference/poly2nb.html) of **spdep** to compute the neighbours list from polygon list.

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

The code chunk below first extracts the geometry component from `shan_sf` using `st_geometry()`. It then calculates the centroids of the polygons `st_centroid()` and coverts them to their coordinates `st_coordinates()`. Two plots are produced: the first shows the outlines of shan_sf Shan State townships, and the second overlays the neighbour linkages on top of the map. The neigbhour network is drawn between centroids of the polygons.

```{r}
coords <- st_coordinates(
  st_centroid(st_geometry(shan_sf))
)
```


```{r}
plot(st_geometry(shan_sf),
     border = gray(0.5))
plot(shan.nb,
     coords,
     col = "blue",
     add = TRUE)
```

### Calculating Edge Costs

The code chunk below uses [`nbcosts()`](https://r-spatial.github.io/spdep/reference/nbcosts.html) of **spdep** to computes attribute-based costs, e.g. the absolute difference in attribute values (or Euclidean distance in attribute space) between neighbours, as passing in a data.frame of attributes.


```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
head(lcosts, 3)
```
For each observation, this computes the pairwise dissimilarity on the five variables with its neighbours, serving as a generalised spatial weight. We then convert the neighbour list into a weights object, using the calculated lcosts as the weights, similar to how inverse distance weights were derived.

In order to achieve this, [*nb2listw()*](https://r-spatial.github.io/spdep/reference/nb2listw.html) of **spdep** package is used as shown in the code chunk below.


```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```
### Computing minimum spanning tree

The code chunk below calculates the minimum spanning tree by using [`mtree()`](https://r-spatial.github.io/spdep/reference/mstree.html) of spdep.

```{r}
shan.mst <- mstree(shan.w)
summary(shan.mst)
class(shan.mst)
dim(shan.mst)
head(shan.mst,3)
```

The minimum spanning tree has n-1 edges(links). 

:::callout-note

MST is about connecting everything cheaply (like designing a road or utility network). V1, V2, V3 are self row index, neighbour's row index, and edge weight (distance or dissimilarity).

Hamiltonian Path and Travelling Salesman Problem are about visiting everything in a sequence (like a salesman’s route).
:::

The code chunk plots a minimum spanning tree that connects all townships using edges weighted by attribute-based similarity.
```{r}
plot(st_geometry(shan_sf), border = gray(0.5))
plot.mst(shan.mst, coords,
         col = "blue", cex.lab = 0.7,
         cex.circles = 0.005, add = TRUE)
```

### Computing Spatially Constrained Clusters Ysing SKATER Method

The code chunk uses the [`skater()`](https://r-spatial.github.io/spdep/reference/skater.html) of **spdep**

```{r}
clust6 <- skater(edges = shan.mst[,1:2],
                 data = shan_ict,
                 method = "euclidean",
                 ncuts = 5)
```


```{r}
str(clust6)
```



:::callout-note
skater(edges, data, ncuts, crit, vec.crit, method = c("euclidean", 
    "maximum", "manhattan", "canberra", "binary", "minkowski", 
    "mahalanobis"), p = 2, cov, inverted = FALSE)
    
    
- Three Mandatory Arguments: 

the first two columns of the MST matrix (i.e. not the cost), 

the data matrix (to update the costs as units are being grouped), 

the number of cuts. Note: **It is set to one less than the number of clusters.** 

:::

The output of `skater` object consists of `groups` vector which labels cluster for each observation.

The edges.groups list provides a summary for each cluster.

Sum-of-squares measures e.g. ssto (total) and ssw (within) shows how each cut affects the overall criterion.

The chunk blow stores the `groups`vector from `skater` object `clust6`  which assigns each row(township) to a cluster. If we want the number of neighbours for each township, we could use nb list `shan.nb` to check the cadinality.

```{r}
ccs6 <- clust6$groups
ccs6
```


```{r}
table(ccs6)
```
```{r}
clust6$edges.groups[[1]]$node
```

Lastly, we can also plot the pruned tree that shows the six clusters on top of the township area.

```{r}
plot(st_geometry(shan_sf), border = gray(0.5))
plot(clust6, coords, cex.lab = .7, 
     groups.colors = c("red","green", "blue", "brown", "pink"),
     cex.circles = 0.005,
     add = TRUE)
```
### Visualising the Clusters in Choropleth Map

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename (`SP_CLUSTER` = `as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

The code chunk below plots the hierarchical clustering ad spatially constrained hierarchical clustering maps side by side.


```{r, fig.width=12, fig.height=6}

hclust.map <- qtm(shan_sf_cluster, "CLUSTER") + 
  tm_borders() +
  tm_layout(legend.position = c("right","top")) +
  tm_title("Hierarchical Clustering")


shclust.map <- qtm(shan_sf_spatialcluster, "SP_CLUSTER") + 
  tm_borders() +
  tm_layout(legend.position = c("right","top")) +
  tm_title("Spatially Constrained Hierarchical Clustering")

tmap_arrange(hclust.map, shclust.map,
             asp = NA, ncol = 2)
```

## Spatially Constrained Clustering: ClustGeo Method

In this section, we perform non-spatially constrained hierarchical cluster analysis and spatially constrained cluster analysis with **ClustGeo** package.


### A short note about ClustGeo package

[ClustGeo](https://cran.r-project.org/web/packages/ClustGeo/) is an R package for spatially constrained clustering. 

It implements a Ward-like (variance-minimising) hierarchical method, `hclustgeo()`, that blends two dissimilarities:

- D0: a `dist` object represents dissimilarity in the **attribute space** (can be non-Euclidean; observations may be weighted).

- D1: a `dist` object represents dissimilarity in the **spatial/constraint space**.

A mixing parameter alpha (0,1) controls the trade-off via a convex combination of homogeneity criteria:

criterion= (1-a) * crit(D0) + a * crit*(D1)

The goal is to choose alpha to increase spatial contiguity without unduly harming cluster quality on the variables of interest. The function `choicealpha()` estimates a suitable alpha by comparing solutions across alpha values.


:::callout-note
hclustgeo(D0, D1 = NULL, alpha = 0, scale = TRUE, wt = NULL)

- hclustgeo(D0) = non-spatial (standard Ward).

- hclustgeo(D0, D1, alpha) = spatially constrained clustering.

- Scale is used when both D0 and D1 are included, and scaled by i.e. divided by their max.

:::

### Ward-like hierarchical clustering: ClustGeo

`hclustgeo()` extends the base R function hclust(). Under the hood, it constructs a dendrogram object with the same structure as stats::hclust. 

 The code chunk below use `hclustgeo()` to perform hierarchical clustering. 

```{r}
nongeo_cluster <- hclustgeo(proxmat)
nongeo_cluster
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, k = 6, border = 2:5) 
#`border = 2:5` means: to use **red, green, blue, and cyan** as the rectangle borders.

```
### Mapping the clusters formed

The code chunk below plots the non-spatial constrained hierachical clustering.

```{r}
groups <- as.factor(cutree(nongeo_cluster, k = 6))
```

```{r}
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```
```{r, fig.width=12, fig.height=6}

hclust.map <- qtm(shan_sf_cluster, "CLUSTER") + 
  tm_borders() +
  tm_layout(legend.position = c("right","top")) +
  tm_title("Hierarchical Clustering (hclust)")

ngeohclust.map <- qtm(shan_sf_ngeo_cluster, "CLUSTER") + 
  tm_borders() +
  tm_layout(legend.position = c("right","top")) +
  tm_title("Non-Spatially Constrained Hierarchical Clustering (hclustgeo)")


tmap_arrange(hclust.map, ngeohclust.map,
             asp = NA, ncol = 2)
```

### Spatially Constrained Hierarchical Clustering

To prepare for spatially constrained hierachical clustering, we use [`st_distance()`](https://r-spatial.github.io/sf/reference/geos_measures.html) of **sf** package to compute a spatial distance matrix.

```{r}
dist <- st_distance(shan_sf, shan_sf)
head(dist,1)
distmat <- as.dist(dist)
summary(distmat)
```
```{r}
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0,1,0.1), K = 6, graph = TRUE)
```

The two diagnostic plots show that at an **alpha** level of **0.2–0.3**, the trade-off between attribute quality and spatial quality is well balanced. Here, Q denotes the **partition quality index**, analogous to an R2 measure, indicating how much of the variability is explained by the clustering partition.


The code chunk below uses alpha = 0.2 to derive the clusters.

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
```

```{r}
groups <- as.factor(cutree(clustG, k = 6))
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename (`CLUSTER` = `as.matrix.groups.`)
qtm(shan_sf_Gcluster, "CLUSTER")
```

```{r, fig.width=12, fig.height=6}

hclust.map <- qtm(shan_sf_Gcluster, "CLUSTER") + 
  tm_borders() +
  tm_layout(legend.position = c("right","top")) +
  tm_title("Spatially Constrained (hclustgeo)")


shclust.map <- qtm(shan_sf_spatialcluster, "SP_CLUSTER") + 
  tm_borders() +
  tm_layout(legend.position = c("right","top")) +
  tm_title("Spatially Constrained (skater)")

tmap_arrange(hclust.map, shclust.map,
             asp = NA, ncol = 2)
```


## Visual Interpretation of Clusters

### Visualising Individual Clustering Variable

The code chunk below visualises the distribution of a clustering variable (i.e. RADIO_PR) by cluster.

```{r}
ggplot(data = shan_sf_ngeo_cluster,
       aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```
The boxplot shows Cluster 3 has the highest Radio Ownership per thousand household on average and followed by Cluster 2, 1, 4, 5 and 6.


```{r}
plot_boxplots <- function(var) {
  ggplot(data = shan_sf_ngeo_cluster, aes(x = CLUSTER, y = .data[[var]])) +
    geom_boxplot() +
    labs( y = var)
  }
vars <- names(shan_ict)
for (v in vars) {print(plot_boxplots(v))}
```
### Multivariate Visualisation

The code chunk creates parallel coordinate plots to reveal clustering variables by cluster using  [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](https://ggobi.github.io/ggally/index.html) package.


```{r, fig.width=12}
ggparcoord(data = shan_sf_ngeo_cluster,
           columns = c(17:21),
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Multiple Paralle Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) +
  theme(axis.text.x = element_text(angle = 30))

```
The parallel coordinate plot shows that households in **Cluster 4** townships tend to own the **highest** numbers of **TVs and mobile phones**, whereas households in **Cluster 5** townships record the **lowest** ownership across all five ICT assets.


Below is a summary of the **`scale` options in `ggparcoord()`**:

* **std**: standardise by mean and standard deviation.
* **robust**: standardise by median and median absolute deviation.
* **uniminmax**: rescale each variable to [0, 1] range.
* **globalminmax**: no scaling; all variables share the same global min–max.
* **center**: uniminmax scaling, then centre at a chosen summary value.
* **centerObs**: uniminmax scaling, then centre at a chosen observation.

There is no universally best method. 

Finally, the code chunk below calculates summary statistics such as mean, median, sd, etc.


```{r}
shan_sf_ngeo_cluster %>%
  st_set_geometry(NULL) %>% #also when we extract clustering variable
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TVO_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```

```{r}
vars <- c("RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")

cluster_summary <- function(data, cluster_var, vars) {
  data %>%
    st_set_geometry(NULL) %>%             # drop geometry if sf
    group_by(.data[[cluster_var]]) %>%
    summarise(
      across(all_of(vars),
             list(mean = mean, sd = sd, median = median),
             .names = "{.col}_{.fn}"),
      .groups = "drop"
    )
}

summary_table_t <- cluster_summary(shan_sf_ngeo_cluster, "CLUSTER", vars) %>%
  t() %>%
  as.data.frame()

kable(summary_table_t, caption = "Cluster Summary (Mean, SD, Median)")
  
```





## Reference

Kam, T. S. 8 Spatial Weights and Applications. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap12
