---
title: "06: Geographical Segmentation with Spatially Constrained Clustering Techniques"
author: "Yiqiong PAN"
date: "Sep 30, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 4

editor: visual
---

## Overview

In this exercise, we use methods like **Hierarchical Cluster Analysis** (HCA) and Spatially Constrained Cluster Analysis (SCCA) including (**SKATER and ClusGeo**) to group neighbouring spatial units into homogeneous regions (areas that are similar to each other in terms of variables, i.e. geographically referenced **multivariate**).

### Learning Outcome

The tasks to be completed are to:

-   convert simple feature data.frame into R's SpatialPolygonDataFrame object using `as_Spatial()` of **sf** package

-   perform cluster analysis with `hclust()` of Base R

-   perform spatially constrained cluster analysis using `skater()` of Base R

-   visualise the analysis using **ggplot2** and **tmap** package.

## Getting Started

### The Analytical Question

In real world application likes geobusiness and geospatail related policy making, regions are defined by statistical similarity in the multivariate data. In the case study of [Shan State](https://en.wikipedia.org/wiki/Shan_State) [Myanmar](https://en.wikipedia.org/wiki/Myanmar), we find and outline the clusters of areas that share similar characteristics, through multiple **Information and Communication technology** (ICT) measures, namely: Radio, Television, Land line phone, Mobile phone, Computer, and Internet at home.

## The Data

| Dataset | Description | Format |
|-------------------------|--------------------------|----------------------|
| myanmar_township_boundaries | GIS Myanmar Township Boundary Data represented as polygon features | ESRI shapefile |
| Shan-ICT.csv | 2014 Myanmar Population and Housing Census at the township level | CSV |

The data can be downloaded from [Myanmar Information Management Unit (MIMU)](http://themimu.info/).



### Installing and Loading R Packages

There are **14** packages used in this exercise.

- Spatial data (**sf, spdep, tmap**)

- Cluster analysis (**ClustGeo, cluster, NbClust, factoextra**)

- Multivariate exploration (**corrplot, heatmaply, psych, GGally**)

- Visualisation / general workflow (**tidyverse, ggpubr**)

| Package | Description |
|------------------|------------------------------------------------------|
| [sf](https://r-spatial.github.io/sf/) | Handles importing, managing, and processing vector-based geospatial data |
| [rgdal](https://cran.r-project.org/package=rgdal) | Provides bindings to the GDAL/OGR library for reading/writing spatial data formats. |
| [spdep](https://cran.r-project.org/package=spdep) | Tools for spatial dependence: building neighbour lists, spatial weights, and spatial autocorrelation analysis. |
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data wrangling and visualization, including **readr**, **dplyr**, and **ggplot2**. |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static and interactive thematic (choropleth) maps. |
| [corrplot](https://cran.r-project.org/package=corrplot) | Provides graphical displays of correlation matrices for multivariate data exploration. |
| [ggpubr](https://cran.r-project.org/package=ggpubr) | Simplifies publication-ready plots based on **ggplot2** with statistical annotations. |
| [heatmaply](https://cran.r-project.org/package=heatmaply) | Generates interactive heatmaps for exploring multivariate data relationships. |
| [cluster](https://cran.r-project.org/package=cluster) | Functions for cluster analysis, including hierarchical, partitioning, and silhouette methods. |
| [ClustGeo](https://cran.r-project.org/package=ClustGeo) | Performs spatially constrained clustering by balancing attribute similarity with geographical contiguity. |
| [NbClust](<https://cran.r-project.org/package=NbClust>) | Determines the optimal number of clusters using multiple indices and methods. |
| [facoextra](<https://cran.r-project.org/package=factoextra>) | Visualisation tools for clustering and multivariate analyses (e.g., PCA, CA, HCPC). |
| [GGally](https://ggobi.github.io/ggally/) | Extension of ggplot2 with functions for pair plots, network plots, and correlation plots. |
| [psych](https://cran.r-project.org/package=psych) | rocedures for psychometric and multivariate analysis, including factor analysis. |


The code chunk below loads the packages into R environment.

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```



## Data Import and Preparation

### Importing Geospatial Data into R Environment

The code chunk below uses[`st_read()`](https://www.rdocumentation.org/packages/sf/versions/0.7-2/topics/st_read) from **sf** package to import Mayanmar Township Boundary (`myanmar_township_boundaries`) in shapefile format into R as a simple feature data.frame.

```{r}
shan_sf <- st_read(dsn = "data/geospatial",
                   layer = "myanmar_township_boundaries") %>%
  filter (ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>% # filter Shan State
  select(c(2:7))    # select columns on ST, DT, and TS
```
```{r}
shan_sf
```
The `sf` object follows Hardy Wickham’s [tidyverse](https://edzer.github.io/rstudio_conf/#1) principles, so we can handle it like a tidy tibble (rows = observations, cols = variables), with an extra geometry column.
```{r}
glimpse(shan_sf)
```
### Importing Aspatial Data into R Environment

The chunk below imports ICT variables via `read_csv()` function of **readr** package as a **tibble** data.frame. The data are from **The 2014 Myanmar Population and Housing Census Myanmar**. 


```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```
The code chunk below summrises the statistics from attribute table. There are a total of eleven fields and 55 observation in the tibble data.frame.

```{r}
summary(ict)
```
### Derive Bew Variables Using dplyr Package


The code chunk below derives the **penetration rate of each ICT variable** based on total household number in township. We should always transform absolute values to **relative measures** before doing spatial analysis or clustering, otherwise the results will be dominated by township size.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

Let's have a quick check on the new talbe.

```{r}
summary(ict_derived)
```
New columns, RADIO_PR, TV_PR, LLPHONE_PR, MPHONE_PR, COMPUTER_PR, and INTERNET_PR,are created as planned.

## Exploratory Data Analysis (EDA)

### EDA Using Statistical Graphics

To examine the variables such as the number of households with radio, we can use graphical methods that highlight different aspects of the data distribution:

**Histogram**

- Shows the overall shape of the distribution.

- Helps identify whether the data is left-skewed, right-skewed, or approximately normal.

- Useful for understanding the frequency of values and detecting clustering or gaps in the data.

```{r,fig.width= 4, fig.height=3}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```



**Boxplot**:

- Summarises the distribution using the five-number summary (minimum, Q1, median, Q3, maximum).

- Makes it easy to detect outliers and understand the spread and central tendency of the data.

- Allows quick comparison across multiple variables or groups (e.g., by township).


```{r,fig.width= 4, fig.height=3}
ggplot(data=ict_derived, 
       aes(x= RADIO )) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Let's also visualise the distribution of the newly derived variables (i.e. Radio penetration rate).

```{r,fig.width= 4, fig.height=3}
ggplot(data = ict_derived,
       aes(x = RADIO_PR)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
```
```{r,fig.width= 4, fig.height=3}
ggplot(data = ict_derived,
       aes(x = RADIO_PR)) +
  geom_boxplot(color = "black", fill = "light blue")
```
:::callout-note
- What can you observed from the distributions reveal in the histogram and boxplot?

The penetration rate data are more normally distributed and show fewer outliers than the raw ownership counts, since the effect of township size has been normalised.
:::

In the figure below, **multiple** histograms are plotted to reveal the distribution of the selected variables in the `ict_derived` data.frame. 

In **tmap**, `tmap_arrange()` (or `tmarange`) is used to arrange multiple map plots side by side or in a grid layout.

In **ggplot2**, the equivalent is [`ggarrange()`](https://rpkgs.datanovia.com/ggpubr/reference/ggarrange.html) from the [**ggpubr**](https://rpkgs.datanovia.com/ggpubr/index.html) package.

```{r, fig.width=10}
#| code-fold: true
#| code-summary: "View code"

radio <- ggplot(data= ict_derived,
                aes(x = RADIO_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")

tv <- ggplot(data = ict_derived, aes(x= TV_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")

llphone <- ggplot(data = ict_derived, aes(x = LLPHONE_PR)) +
  geom_histogram(bins = 20, color = "black",fill = "light blue")

mphone <- ggplot(data = ict_derived, aes(x = MPHONE_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")

computer <- ggplot(data = ict_derived, aes(x = COMPUTER_PR)) +
  geom_histogram(bins = 20, color = "black", fill ="light blue") 

internet <- ggplot(data = ict_derived, aes(x = INTERNET_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")

ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3,
          nrow = 2)
```











## Reference

Kam, T. S. 8 Spatial Weights and Applications. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap12
