---
title: "In-class Exercise 8: Supplement to Hands-on Exercise 8"
author: "Yiqiong PAN"
date: "25 Oct, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 5

editor: visual
---

## The Data

The study integrates both **aspatial** and **geospatial** datasets to model HDB resale prices in Singapore.

| Dataset | Description | Format |
|----|----|----|
| Aspatial dataset | HDB Resale data: a list of HDB resale transacted prices in Singapore from Jan 2017 onwards, downloaded from [Data.gov.sg.](Data.gov.sg) | CSV |
| Geospatial dataset | *P14_SUBZONE_WEB_PL*: a polygon feature data providing information of URA 2014 Master Plan Planning Subzone boundary data, also downloaded from [Data.gov.sg.](Data.gov.sg) | ESRI shapefile |
| Locational factors with geographic coordinates | Eldercare centres, Hawker centres ,Parks ,Supermarkets ,CHAS clinics ,Childcare services ,Kindergartens downloaded from [Data.gov.sg](Data.gov.sg) MRT/LRT stations ,Bus stops downloaded from [Datamall.lta.gov.sg](Datamall.lta.gov.sg), | GeoJSON or Shapefile |
| Locational factors without geographic coordinates | Primary schools, extracted from Data.gov.sg’s school information dataset. CBD coordinates, retrieved from Google. Shopping malls , scraped from [Wikipedia](https://en.wikipedia.org/wiki/List_of_shopping_malls_in_Singapore). Good primary schools , ranked list of popular schools from[Local Salary Forum.](https://www.salary.sg/2021/best-primary-schools-2021-by-popularity/) | CSV |

{tbl-colwidths="\[30,60,10\]"}

## Installing and Loading R packages

There are 8 packages used in this exercise.

|   | Description |
|----|----|
| [GWmodel](https://cran.r-project.org/web/packages/GWmodel/index.html) | Calibrates geographical weighted family of models |
| [rsample](https://rsample.tidymodels.org/) | Provides functions to create different types of resamples and corresponding classes for their analysis. |
| [sf](https://r-spatial.github.io/sf/) | Handles importing, managing, and processing vector-based geospatial data |
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data wrangling and visualisation, including **readr**, **dplyr**, and **ggplot2**. |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static and interactive thematic (choropleth) maps. |
| [SpatialML](https://cran.r-project.org/web/packages/SpatialML/index.html) | Allows for a geographically weighted random forest regression including a function to find the optical bandwidth.  |
| [Metrics](https://cran.r-project.org/web/packages/Metrics/index.html) | Implements metrics for regression, time series, binary classification, classification, and information retrieval problems. |
| [spdep](https://cran.r-project.org/web/packages/spdep/index.html) | Provides tools for spatial dependence: building neighbour lists, spatial weights, and spatial autocorrelation analysis. |

{tbl-colwidths="\[15,85\]"}

We load the following packages into R environment.

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, Metrics, tidyverse, knitr, kableExtra, spatialRF,randomForestExplainer)
```

## Preparing Data

### Reading data file to rds

We first read the input data sets that is saved in `simple feature` data frame.

```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```

### Data Sampling

Calibrating predictive models are computaional intensive, especailly random forest method is used. FOr quick prototyping, 1 10% sample will be selected at random from the data. The code chunk below samples 1500 observations.

```{r}
set.seed(1234)
HDB_sample <-mdata %>%
  sample_n(1500)
```

### Checking of overlapping points

When using GWR model to build explanatory or predictive models, we should first check if there are overlapping points in the HDB locations. In this exercise, we only check the overlapping on the sample data.

```{r}
overlapping_points <- HDB_sample %>%
  mutate(overlap = lengths(st_equals(., .)) >1)
```

### Spatial Jitter

[`st_jitter()`](https://r-spatial.github.io/sf/reference/st_jitter.html) of **sf** package is used to move the overlapped point features away from each other.

```{r}
HDB_sample <- HDB_sample %>%
  st_jitter(amount = 1)
```

Then we could run the overlapping check again to make sure there are no overlapping points in the data set.

```{r}
overlapping_points <- HDB_sample %>%
  mutate(overlap = lengths(st_equals(., .)) >1)
```

### Data Sampling

We split the entire data into 65% training and 35% test by using `initial_split()` of **rsample** package, which belongs to **tigymodels** ecosystem. Then we extract the training set and test set from a data split object created by functions `initial_split()`.

```{r}
set.seed(1234)
resale_split <- initial_split(HDB_sample, prop = 6.67/10,) #1000 & 500
train_data <- training(resale_split)
test_data <- testing(resale_split)
```

We save both files on the disk.

```{r}
write_rds(train_data, "data/model/train_data.rds")
write_rds(test_data, "data/model/test_data.rds")
```

### Computing Correlation Matrix

Before loading predictor into a predictive model, let's plot the correlation matrix to check if there is any sign of multicolinearity.

```{r, fig.width=14, fig.height=10}
HDB_sample_nogeo <- st_drop_geometry(HDB_sample)
corrplot::corrplot(cor(HDB_sample_nogeo[, 2:17]),
                   diag = FALSE, # hide diagonal or TRUE
                   order = "AOE",# angular order of the eigenvectors" /original/AOE/FPC/hclust/alphabet
                   tl.pos = "td", # text label position: top and diagonal/lt/n null
                   tl.cex = 0.5,  # text label size: shrinks 50%
                   method = "number",  # display number/circle/color/shade/pie
                   type = "upper") #show only the upper triangle/full/lower
```

There is no sign of multicolinearity since all correlation values are below **0.8**.

In order to avoid multicollineariy, we use [`ggcorrmat()`](https://indrajeetpatil.github.io/ggstatsplot/articles/web_only/ggcorrmat.html) of **ggstatsplot** is to plot a correlation matrix to check if there are pairs of highly correlated independent variables.

```{r, fig.width=14, fig.height=10}
mdata_nogeo <- mdata %>%
  st_drop_geometry()
ggstatsplot::ggcorrmat(mdata_nogeo[, 2:17])
```

## Retriving the Stored Data

We load the saved data into R.

```{r}
train_data <- read_rds("data/model/train_data.rds")
test_data <- read_rds("data/model/test_data.rds")
```

## Building a Non-spatial Multiple Linear Regression

A multiple regression model is built as a base line model.

```{r}

price_mlr <- lm(resale_price ~ floor_area_sqm +                   # target ~ independent variables
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                data=train_data)
summary(price_mlr)
```

The regression summary report indicates that the model slightly underestimates HDB resale prices, as shown by the residual median of –1,930. However, the R2 value of 0.7373 suggests that the model explains approximately 74% of the variation in HDB resale prices since 2017, which exhibits good predictive performance overall.

### Multicolinearity Check

```{r}
vif <- performance::check_collinearity(price_mlr)
kable(vif, 
      caption = "Variance Inflation Factor (VIF) Results") %>%
  kable_styling(font_size = 18) 
```

```{r}
plot(vif) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| eval: false
write_rds(price_mlr, "data/model/price_mlr.rds" ) 
```

## gwr Predictive Method

### Computing Adaptive Bandwidth

```{r}

bw_adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
    WITHIN_1KM_PRISCH,
  data=train_data,
  approach="CV",
  kernel="gaussian",
  adaptive=TRUE,
  longlat=FALSE)
```

```{r}
bw_adaptive
```

```{r}
#| eval: false
write_rds(bw_adaptive, "data/model/bw_adaptive.rds")
```

```{r}
#| echo: false
bw_adaptive <- read_rds("data/model/bw_adaptive.rds")
```

Then we build the GWR-based hedonic pricing model using adaptive bandwidth and Gaussian kernel. Same reason as the previous step, we save a copy of the `gwrm` file.

```{r}
#| eval: false
gwr_adaptive <- gwr.basic(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  data=train_data,
  bw=bw_adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE,
  longlat = FALSE)
write_rds(gwr_adaptive, 
          "data/model/gwr_adaptive.rds")
```

### Retrieve gwr Output Object

We re-load the saved `gwrm` object and retrieve the report of model output.

```{r}
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
```

```{r}
gwr_adaptive
```

Please refer to [Hands-on_Ex08](https://isss626-ypan.netlify.app/hands-on_ex/hands-on_ex08/hands-on_ex08)

## Predictive Modelling with MLR

### Predicting with test data

### Test data bandwidth

```{r}
#| eval: false
gwr_bw_test_adaptive <- bw.gwr(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                  data=test_data,
                  approach="CV",
                  kernel="gaussian",
                  adaptive=TRUE,
                  longlat=FALSE)

write_rds(gwr_bw_test_adaptive,
          "data/model/gwr_bw_test.rds")
```

```{r}
gwr_bw_test_adaptive <- read_rds(
  "data/model/gwr_bw_test.rds")
```

### Predicting

```{r}
#| eval: false
gwr_pred <- gwr.predict(formula = resale_price ~
                          floor_area_sqm + storey_order +
                          remaining_lease_mths + PROX_CBD + 
                          PROX_ELDERLYCARE + PROX_HAWKER + 
                          PROX_MRT + PROX_PARK + PROX_MALL + 
                          PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                          WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
                          WITHIN_1KM_PRISCH, 
                        data=train_data, 
                        predictdata = test_data, 
                        bw=bw_adaptive, 
                        kernel = 'gaussian', 
                        adaptive=TRUE, 
                        longlat = FALSE)

write_rds(gwr_pred,
          "data/model/gwr_pred.rds")
```

```{r}
gwr_pred<- read_rds("data/model/gwr_pred.rds")
```

```{r}
gwr_pred
```

## Predictive Modelling: RF method

### Data preparation

Firstly, code chunk below is used to extract the coordinates of training and test data sets

```{r}
coords <- st_coordinates(HDB_sample)
coords_train <- st_coordinates(train_data)
coords_test <- st_coordinates(test_data)
```

Next, code chunk below is used to drop the geometry column of both training and test data sets.

```{r}
train_data_nogeom <- train_data %>%
  st_drop_geometry()
```

### Calibrating RF model

```{r}
set.seed(1234)
rf <- ranger(resale_price ~ floor_area_sqm + storey_order + 
               remaining_lease_mths + PROX_CBD + PROX_ELDERLYCARE + 
               PROX_HAWKER + PROX_MRT + PROX_PARK + PROX_MALL + 
               PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
               WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
               WITHIN_1KM_PRISCH,
             data=train_data_nogeom)

```

```{r}
rf
```

## Predictive Modelling: SpatialML method

### Determining bandwidth

```{r}
#| eval: false
set.seed(1234)
gwRF_bw <- grf.bw(formula = resale_price ~ floor_area_sqm + 
                       storey_order + remaining_lease_mths + 
                       PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER + 
                       PROX_MRT + PROX_PARK + PROX_MALL + 
                       PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                       WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                       WITHIN_1KM_PRISCH,
                     dataset=train_data_nogeom, 
                     kernel="adaptive",
                     coords=coords_train,  # to speed up:
                     trees = 100,        # Reduce from 500 to 100-200
                     nthreads = parallel::detectCores() - 1,  # Use multiple cores
                     step = 2            # Larger step size in bandwidth search
)
```

```{r}
#| eval: false
write_rds(gwRF_bw,
          "data/model/gwRF_bw.rds")
```

```{r}
#| eval: false
gwRF_bw <-read_rds("data/model/gwRF_bw.rds")
```

```{r}
#| eval: false
gwRF_bw
```

### Calibrating with grf

```{r}
#| eval: false
set.seed(1234)
gwRF_adaptive <- grf(formula = resale_price ~ floor_area_sqm + 
                       storey_order + remaining_lease_mths + 
                       PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER + 
                       PROX_MRT + PROX_PARK + PROX_MALL + 
                       PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                       WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                       WITHIN_1KM_PRISCH,
                     dframe=train_data_nogeom, 
                     bw=55,
                     kernel="adaptive",
                     coords=coords_train)

write_rds(gwRF_adaptive, "data/model/gwRF_adaptive.rds")
```

```{r}
gwRF_adaptive <- read_rds("data/model/gwRF_adaptive.rds")
```

## Predicting by using the test data

### Preparing the test data

```{r}
test_data_nogeom <- cbind(
  test_data, coords_test) %>%
  st_drop_geometry()
```

### Predicting with the test data

In the code chunk below, `predict.grf()` of **spatialML** for predicting re-sale prices in the **test data** set (i.e. *test_data_nogeom*)

```{r}
gwRF_pred <- predict.grf(gwRF_adaptive, 
                           test_data_nogeom, 
                           x.var.name="X",
                           y.var.name="Y", 
                           local.w=1,
                           global.w=0)
write_rds(gwRF_pred, "data/model/gwRF_pred.rds")
```

```{r}
gwRF_pred <- read_rds("data/model/gwRF_pred.rds")
```

```{r}
summary(gwRF_pred)
```

### Creating DF

Next, the code chunk below is used to convert the output from `predict.grf()` into a data.frame.

```{r}
GRF_pred_df <- as.data.frame(gwRF_pred)
```

Then, `cbind()` is used to append fields in GRF_pred_df data.frame onto test_data.

```{r}
test_data_pred <- cbind(test_data, 
                        GRF_pred_df)
```

## Visualising the predicted values

```{r}
ggplot(data = test_data_pred,
       aes(x = gwRF_pred,
           y = resale_price)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, col = "red", linetype = 1)
```

## Reference

Kam, T. S. 14 Geographically Weighted Predictive Models. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap14.html
