---
title: "Take-home Exercise 3: Modelling HDB Resale Prices with Geographically Weighted Methods"
author: "Yiqiong PAN"
date: "27 Oct, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 6

editor: visual
---

## Overview

Buying a house is considered as one of the most significant investments for individuals. People purchase homes for various reasons, such as building wealth, achieving financial stability, and fulfilling the desire for ownership and privacy. Understanding housing prices is essential, as fluctuations in property values can have profound effects on individuals’ financial wellbeing and the broader economy.

A house price can be influenced by multiple factors, including the overall condition, size, age, and location of the property, as well as the general economy state and broader market sentiment. Traditionally, hedonic pricing models for housing have been estimated using the Ordinary Least Squares (OLS) method. However, this approach does not adequately account for spatial autocorrelation and spatial heterogeneity, both of which are commonly found in housing transaction data. Failure to consider these spatial effects can lead to biased, inconsistent, or inefficient OLS estimates (Anselin, 1998). To address these limitations, Geographically Weighted Models (GWMs) have been developed to better capture spatial variations and improve the explanatory accuracy of housing resale price analysis.

## Objective

In Singapore, the Housing and Development Board (HDB) resale market offers a dynamic context for examining the determinants of housing prices. With about 90 percent of households owning their homes and approximately 74 percent residing in HDB flats, public housing plays a central role in national housing trends (Lim, 2025).

In recent years, strong demand for housing upgrades and a limited supply of relatively new flats have fuelled a steady rise in HDB resale prices, particularly for four room and five room units (Sia, 2025). However, recent data shows that price growth has slowed to its weakest pace in five years, rising by only 0.4 percent in Q3 2025, even as resale activity increased by 1.7 percent (Lim, 2025).

Given these evolving market dynamics, this study aims to develop a hedonic pricing model for HDB resale prices in Singapore using Geographically Weighted Models to more effectively capture spatial variations and identify the key factors influencing housing values across the island.

## Literature Review

Early housing studies relied on global regression models, assuming that the relationship between housing prices and their determinants was spatially uniform. Using this classical approach, Xu (2017) applied a hedonic pricing model to Singapore’s HDB resale market from 2000 to 2016, identifying the importance of structural and locational factors such as flat size, age, and proximity to MRT stations in explaining price variation. Similarly, Chia, Li, and Tang (2017) developed a dynamic general equilibrium model linking Singapore’s public and private housing markets, finding that economic fundamentals, especially demographics and income growth, accounted for nearly 50% of public and over 80% of private housing price increases. These non-spatial approaches, while informative, assumed constant parameter effects across space and could not capture spatial non-stationarity inherent in housing markets.

To address these limitations, Brunsdon, Fotheringham, and Charlton (1996) introduced Geographically Weighted Regression (GWR), a spatially local extension of ordinary least squares that allows regression coefficients to vary by location. This innovation made it possible to identify local variations in housing price determinants across geographic space. More recent research has extended spatial modelling to Singapore’s housing market. Wang et al. (2022) incorporated taxi travel time as a spatial weighting mechanism within a GWR framework, demonstrating superior performance over conventional Euclidean-distance models in capturing spatial price variation. Meanwhile, policy studies such as Phang (2015) contextualised these empirical advances by tracing the evolution of Singapore’s state-led housing policies, linking economic transitions and planning strategies to housing affordability and ownership stability.

Together, these studies illustrate a clear methodological progression, from global, non-spatial hedonic models to local, spatially fixed and adaptive approaches that better capture heterogeneity in housing market behaviour. Building upon this foundation, this study applies a log-transformed GWR model to analyse HDB 4-room resale prices (Jan–Sep 2025), combining structural and locational attributes to explain spatial variations across Singapore’s public housing landscape.

## Independent Variables for the Hedonic Pricing Model

Developing a hedonic pricing model requires careful selection of both intrinsic and environmental factors that influence resale prices. By incorporating these factors, the model aims to better capture spatial variations and identify the key determinants of housing values across Singapore.

The independent variables considered in this study include both structural and locational attributes. Additional relevant factors that may enhance the model’s accuracy are also included.

**Structural Factors**

-   Floor area (square metres): total size of the flat.

-   Floor level: the relative storey height of the unit, typically classified as low, middle, or high floor.

-   Remaining lease (months): tenure balance of the flat at the time of transaction.

-   Age of the unit (months): number of months since completion.

-   Main Upgrading Programme (MUP or HIP): whether major upgrading works have been completed.

**Locational Factors**

-   Proximity to the Central Business District (CBD): measured using City Hall MRT Station as the reference point.

-   Proximity to eldercare facilities.

-   Proximity to hawker centres.

-   Proximity to MRT stations.

-   Proximity to parks.

-   Proximity to good primary schools: school quality determined by popularity scores.

-   Proximity to shopping malls.

-   Proximity to supermarkets.

-   Proximity to hospitals (new).

-   Proximity to community clubs (new).

-   Proximity to sports facilities(new)

-   Number of kindergartens within 350 metres.

-   Number of childcare centres within 350 metres.

-   Number of bus stops within 350 metres.

-   Number to CHAS clinics within 350 metres (new).

-   Number of primary schools within 1 kilometre.

## Getting Started

### The Data

The study integrates both **aspatial** and **geospatial** datasets to model HDB resale prices in Singapore.

+---------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| Dataset                                           | Description                                                                                                                                                                                                                     | Format & Source           |
+===================================================+=================================================================================================================================================================================================================================+===========================+
| Aspatial dataset                                  | HDB Resale data: a list of HDB resale transacted prices in Singapore from Jan 2017 onwards, downloaded from [Data.gov.sg.](Data.gov.sg)                                                                                         | CSV                       |
+---------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| Geospatial dataset                                | MasterPlan2019SubzoneBoundaryNoSeaKML: a polygon feature data providing information of URA 2019 Master Plan Planning Subzone boundary data, also downloaded from [Data.gov.sg.](Data.gov.sg)                                    | KML                       |
+---------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| Structural factor without geographic coordinates  | A list of of completion information for each block was extracted from the [HDB](https://services2.hdb.gov.sg/webapp/BB33RESLSTATUS/BB33SEnquiry) records.                                                                       | CSV                       |
+---------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| Locational factors with geographic coordinates    | The following datasets were downloaded from [data.gov.sg](Data.gov.sg)                                                                                                                                                          | GeoJSON, KML or Shapefile |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Eldercare centres                                                                                                                                                                                                           |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Hawker centres                                                                                                                                                                                                              |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Parks                                                                                                                                                                                                                       |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Supermarkets                                                                                                                                                                                                                |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   CHAS clinics                                                                                                                                                                                                                |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Childcare services                                                                                                                                                                                                          |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Kindergartens                                                                                                                                                                                                               |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Community clubs                                                                                                                                                                                                             |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   MRT and LRT station exits                                                                                                                                                                                                   |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | The following datasets were downloaded from [**datamall.lta.gov.sg**](https://datamall.lta.gov.sg/content/datamall/en/static-data.html):                                                                                        |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Bus stops                                                                                                                                                                                                                   |                           |
+---------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| Locational factors without geographic coordinates | -   A list of Primary schools, extracted from the Ministry of Education ([MOE](https://www.moe.gov.sg/schoolfinder?journey=Primary%20school)).                                                                                  | CSV                       |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   CBD coordinates: retrieved from Google.                                                                                                                                                                                     |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Shopping malls: scraped from [Wikipedia](https://en.wikipedia.org/wiki/List_of_shopping_malls_in_Singapore).                                                                                                                |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Hospitals: scraped from [Wikipeidia](https://en.wikipedia.org/wiki/List_of_hospitals_in_Singapore).                                                                                                                         |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Good primary schools: ranked list of popular schools from [Local Website](https://www.creativecampus.com.sg/best-primary-schools-in-singapore-2024)[.](https://www.salary.sg/2021/best-primary-schools-2021-by-popularity/) |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Sports facilities: scrapped from the [ActiveSG Circle](#0) website.                                                                                                                                                         |                           |
|                                                   |                                                                                                                                                                                                                                 |                           |
|                                                   | -   Demographic distribution by planning zone (June 2024): downloaded from the Department of Statistics ([singstat](#0)).                                                                                                       |                           |
+---------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+

### Installing and Loading R packages

::: panel-tabset
#### Package Description

This analysis utilises **23** R packages in total.

+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Package                                                                                               | Description                                                                                                                                                                                                                                  |
+=======================================================================================================+==============================================================================================================================================================================================================================================+
| [olsrr](https://olsrr.rsquaredacademy.com/)                                                           | Builds OLS and performs diagnostics tests                                                                                                                                                                                                    |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [GWmodel](https://cran.r-project.org/web/packages/GWmodel/index.html)                                 | Calibrates geographical weighted family of models                                                                                                                                                                                            |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [corrplot](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html)            | Performs multivariate data visualisation and analysis                                                                                                                                                                                        |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [sf](https://r-spatial.github.io/sf/)                                                                 | Handles importing, managing, and processing vector-based geospatial data                                                                                                                                                                     |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [tidyverse](https://www.tidyverse.org/)                                                               | A collection of R packages for data wrangling and visualisation, including **readr**, **dplyr**, and **ggplot2**.                                                                                                                            |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [tmap](https://r-tmap.github.io/tmap/)                                                                | Creates cartographic quality static and interactive thematic (choropleth) maps.                                                                                                                                                              |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [ggpubr](https://rpkgs.datanovia.com/ggpubr/)                                                         | Provides some easy-to-use functions for creating and customising ‘ggplot2’- based publication ready plots.                                                                                                                                   |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [gtsummary](https://www.danieldsjoberg.com/gtsummary/)                                                | Provides an elegant and flexible way to create publication-ready analytical and summary tables.                                                                                                                                              |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [performance](https://cran.r-project.org/web/packages/performance/index.html)                         | Provides assessment of regression models performance (e.g. r-squared)                                                                                                                                                                        |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [RColorBrewer](https://cran.r-project.org/web/packages/RColorBrewer/index.html)                       | Provides color schemes for maps (and other graphics) designed by Cynthia Brewer.                                                                                                                                                             |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [sfdep](https://sfdep.josiahparry.com/)                                                               | built on **spdep** provides a tidyverse-friendly interface with list-columns and additional functionality. **spdep** provides tools for spatial dependence: building neighbour lists, spatial weights, and spatial autocorrelation analysis. |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [httr](https://cran.r-project.org/web/packages/httr/index.html)                                       | Tools designed for handling HTTP requests, organised according to HTTP verbs such as GET and POST.                                                                                                                                           |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html)                               | An efficient JSON parser and generator optimised for web use and statistical data processing.                                                                                                                                                |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [knitr](https://cran.r-project.org/web/packages/knitr/index.html)                                     | Provides an elegant, flexible, and fast static table generation with R                                                                                                                                                                       |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) | An extension of knitr for creating elegant html table with R                                                                                                                                                                                 |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [DT](https://rstudio.github.io/DT/)                                                                   | Provides an R interface to the JavaScript library DataTables for create interactive html tables.                                                                                                                                             |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [progress](https://cran.r-project.org/web/packages/progress/index.html)                               | Provides customisable progress bars including percentage and estimated completion time.                                                                                                                                                      |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [rvest](https://cran.r-project.org/web/packages/rvest/index.html)                                     | Scrapes or harvests data from web pages.                                                                                                                                                                                                     |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [plotly](https://plotly.com/r/getting-started/)                                                       | Creates interactive praphs.                                                                                                                                                                                                                  |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [ggstatsplot](https://indrajeetpatil.github.io/ggstatsplot/)                                          | An extension of the ggplot2 package that allows statistical test results to be displayed within the plots, making the visualisations more informative and data-driven                                                                        |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [spatialRF](https://cran.r-project.org/web/packages/spatialRF/index.html)                             | Spatial features selection for Random Forest spatial regression.                                                                                                                                                                             |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [gt](https://gt.rstudio.com/)                                                                         | Creates polished, customisable tables for reporting and visualisation                                                                                                                                                                        |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [randomForestExplainer](https://cran.r-project.org/web/packages/randomForestExplainer/index.html)     | Interpreting Random Forests through variable Importance                                                                                                                                                                                      |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [reactable](https://glin.github.io/reactable/)                                                        | Creates interactive data tables for R with features such as sorting, grouping and customisations.                                                                                                                                            |
+-------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

#### The Code

The following code installs and initialises the necessary R packages for this analysis.

```{r}
pacman::p_load(olsrr, GWmodel, corrplot, sf, tidyverse, tmap, ggpubr, gtsummary, performance, RColorBrewer, sfdep, httr, jsonlite, knitr, kableExtra, DT, progress, rvest, plotly, ggstatsplot,spatialRF, gt, randomForestExplainer, reactable)
```
:::

## Importing Data into R Environment

### HDB Resale Dataset

HDB resale information is publicly available through the Singapore Government’s open data portal. Detailed transaction records are accessible dating back to the 1990s. For this study, we utlise the dataset “Resale flat prices based on registration date from Jan-2017 onwards”, which provides resale transactions from January 2017 to the present.

#### Import and Extraction

We first import the dataset `HDBresale.csv`, downloaded from [data.gov.sg](data.gov.sg) on 25 October 2025. For this study, we focus specifically on **4-room** transactions between **January 2025 and September 2025 (inclusive)**. After filtering, we obtain **8,679** transactions from **4,148** HDB blocks. These 4-room transactions over the last nine months are treated as potential comparable sales for analysis.

The code chunk below extracts transactions within this study period.

```{r}
HDBresale_all <- read_csv(
  "data/rawdata/HDBresale.csv") %>%
  filter(month >= "2025-01" & month < "2025-10")
```

The code chunk below provides an overview of the columns in the dataset.

```{r}
glimpse(HDBresale_all)
```

The dataset consists of 11 columns in total, which are:

-   `month` indicates the transaction date in the format *YYYY/MM*.

-   `town` refers to the planning area.

-   `flat_type` includes categories such as 2-room, 3-room, 4-room, etc., and we will extract only 4-room transactions for this study.

-   `block` and `street_name` describe the property’s location, which can be used to retrieve geometric information (X and Y coordinates) from **OneMap**.

-   `storey_range` represents the floor range in text format (e.g., “04 TO 06”).

-   `floor_area_sqm` is the size of the property in square metres.

-   `flat_model` denotes the layout and design type of the flat. As these evolve over time and have limited explanatory value, we exclude this variable from the model.

-   `lease_commence_date` indicates the year the property obtained its Temporary Occupation Permit (TOP), which is similar in meaning to `remaining_lease`, expressed as text such as “53 years 1 month”.

-   Finally, `resale_price` is our **target variable**.

From the statistic report below, we observe 20,076 transactions across all flat types for the period from January 2025 to September 2025. The median resale price is approximately \$650,000, with a range of \$250,000 to \$1.65 million. The median year of TOP year is 1998, making the flats around 27 years old on average. The median floor area is 93 sqm.

```{r}
summary(HDBresale_all)
```

The following code filters the dataset to include only **4-room** flat transactions.

```{r}
HDBresale <- HDBresale_all %>%
  filter(flat_type == "4 ROOM")
```

In this step, we present a preview table of the filtered 4-room transactions. Since GWR modelling requires spatial coordinates, we must ensure each transaction includes its geographic information. Therefore, we assign latitude and longitude values to each property in the next.

```{r}
kable(head(HDBresale,10), caption = "Preview of 4-Room HDB Resale Transactions", align = "c")
```

#### Geocoding HDB Properties

In this section, The `block` and `streetname` attributes are used to retrieve the geographic coordinates from the OneMap API. Prior to this, abbreviations "ST." are standardised to "SAINT" in the `streetname` field to avoid potential mismatches or errors during the coordinate retrieval process.

The code chunk below uses `gsub(a, b, col)` from base R to replace all of the occurrences of `ST.` to `SAINT` in the column `street_name`.

```{r}
#| eval: false
HDBresale$street_name <- gsub(
  "ST\\.", "SAINT",
  HDBresale$street_name)
```

Next the function called `geocode` is created, which uses the `GET()` function from the **httr** package to send HTTP requests to the OneMap API, and `fromJSON()` from **jsonlite** package to parse the returned JSON data.

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "CODE"
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep= " ")
  query <-list(
    "searchVal" = address,
    "returnGeom" = "Y",
    "getAddrDetails" = "N",  #Y get full address
    "pageNum" = "1")
  res <- GET(base_url, query = query)
  restext <-content(res, as="text")
  
  output <- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)
  
  return(output)
}

```

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "CODE"
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for (i in 1:nrow(HDBresale)) {
  temp_output <- geocode(HDBresale[i,4], HDBresale[i,5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

As the retrieved LATITUDE and LONGITUDE values are in EPSG:4326 (WGS84), we reproject them to EPSG:3414, a Singapore-specific projected coordinate reference system that enables accurate distance calculations. Following the transformation, the `HDBresale` tibble is converted to an `sf` object and stored for further analysis.

```{r}
#| eval: false
HDBresale_sf <- st_as_sf(HDBresale,
                         coords = c("LONGITUDE", "LATITUDE"),
                         crs = 4326) %>%
  st_transform(crs = 3414)
write_rds(HDBresale_sf, "data/geospatial/HDBresale_sf.rds")
```

```{r}
#| echo: false
HDBresale_sf <- read_rds("data/geospatial/HDBresale_sf.rds")
```

The code below conducts a validation check and generates summary statistics for the filtered 4-room dataset. Between January 2025 and September 2025, 8,679 4-room HDB resale transactions were observed. The median resale price is \$630,000 (range: \$300,000–\$1.518 million). The median age of the flats is about 22 years, and the mean floor area is 93 sqm. Four-room flats represent the majority of the HDB resale market in Singapore, which is why they are selected as the focus of this study.

```{r}
unique(HDBresale_sf$month)
```

```{r}
unique(HDBresale_sf$flat_type)
```

```{r}
nrow(HDBresale_sf)
```

```{r}
summary(HDBresale_sf)
```

### Singpore Master Plan Subzone 2019

The Singapore Master Plan Subzone 2019 dataset is required for spatial visualisation and analysis in this study. We reuse the Master Plan 2019 Subzone Boundary (No Sea) layer created in *Take-Home Exercise 02*. The detailed processing steps for importing and preparing this dataset can be found [here](https://isss626-ypan.netlify.app/take-home_ex/take-home_ex02/take-home_ex02#importing-the-geospatial-ploygon-data-master-plan-2019-subzone).

```{r}
#Load pre-processed data
mpsz <- read_rds("data/rds/mpsz.rds")
```

The mpsz sf data frame contains the administrative attributes `REGION_N`, `PLN_AREA_N`, and `SUBZONE_N`, along with the corresponding `geometry` column, which stores **MULTIPOLYGON** features. The dataset is projected in the Singapore SVY21 coordinate reference system (EPSG:3414) using XY coordinates.

```{r}
mpsz
```

We generate a quick visual preview of the imported subzone map to confirm that the spatial layer has been loaded and displayed correctly.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons("PLN_AREA_N", col = "MAP_COLORS", alpha = 0.9, border.col = "white") +
  tm_title("Singapore Master Plan Subzones 2019", frame = FALSE) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_legend(show = FALSE) + 
  tm_layout(frame= FALSE)

```

### Structural Factor MUP/HIP

The internal condition of a flat is one of the key factors influencing its valuation. While detailed renovation histories at the unit level are not publicly available, we incorporate flat condition into the model using HDB’s upgrading programmes, specifically, the Main Upgrading Programme (MUP) and the Home Improvement Programme (HIP), as proxy indicators.

The **MUP**, introduced in the 1990s, focused on upgrading older HDB blocks, flats, and neighbourhood facilities. It was progressively phased out and replaced by the **HIP** in 2007, with all MUP projects completed by 2012. The **HIP** continues to provide essential upgrading works to flats that are approximately 30 years old. Under both schemes, the government subsidises the majority of upgrading costs, with flat owners contributing roughly 5%.

Other initiatives, such as the Neighbourhood Renewal Programme (NRP) and the Lift Upgrading Programme (LUP), have operated alongside HIP. Most blocks that underwent HIP also benefited from either LUP or NRP. However, for simplicity and relevance to internal flat condition, this study focuses only on **MUP** and **HIP**.

Since no consolidated dataset is publicly available, we manually compiled the MUP and HIP status for each block. We began by identifying the HDB blocks present in the nine-month resale transaction dataset, and then verified their corresponding upgrading status using the HDB online [portal](https://services2.hdb.gov.sg/webapp/BB33RESLSTATUS/BB33SEnquiry).

The code chunk below selects the unique HDB blocks that appear in the resale transaction dataset.

```{r}
unique_blocks <- HDBresale_sf %>%
  distinct(block, street_name, town)
nrow(unique_blocks)
```

The following code outputs a preview of the unique HDB blocks extracted from the dataset and saves them to a CSV file for manual processing.

```{r}
kable(head(unique_blocks, 10), caption = "Preview of HDB Blocks")
```

```{r}
#| eval: false
write_csv(unique_blocks, "data/rawdata/unique_blocks.csv")
```

We prepared a HIP/MUP status dataset for 4,148 HDB blocks in our study. In the preview table, the HIP and MUP indicators are coded as 1 for blocks that have undergone the respective upgrading programme and 0 for those that have not.

```{r}
HIP <- read_csv("data/rawdata/HIP_MUP.csv")
kable(head(HIP, 10), caption = "Preivew of HIP/MUP History for HDB Blocks")
nrow(HIP)
```

In total, the resale records include 1,832 blocks that have undergone HIP and 124 blocks that have undergone MUP.

```{r}
sum(HIP$HIP)
sum(HIP$MUP)
```

### Locational Factors with Geographic Coordinates

Accessibility to key services and facilities plays a significant role in residential location decisions. Households typically favour properties with convenient access to schools, public transport, healthcare services, and everyday retail amenities. In the this section, we import a series of locational factor datasets, which already includes geographic coordinates.

::: panel-tabset
#### Bus Stops

We reuse the bus stop dataset created in *Take-home Exercise 02*, which contains the official Land Transport Authority (LTA) bus stop locations (ESRI shapefile, issued August 2025). The detailed preprocessing workflow is documented in this [exercise](https://isss626-ypan.netlify.app/take-home_ex/take-home_ex02/take-home_ex02#importing-the-attribute-data-bus-trips).

During preprocessing, we removed **five** bus stops located outside Singapore (primarily transit points in Johor Bahru, Malaysia) and **eighteen** duplicate bus stops, which were likely associated with temporary operational setups such as road works or infrastructure upgrades.

```{r}
#Load pre-processed data
BusStop_in_SG <- read_rds("data/geospatial/BusStop_in_SG.rds")
BusStop_in_SG
```

The map below visualises the spatial distribution of 5149 bus stops accross Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(BusStop_in_SG) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("Bus Stops in Singapore") +
  tm_layout(frame = FALSE)
```

#### Community Clubs

The Community Clubs dataset, provided in KML format and containing **125** community club locations, is imported into the R environment. As the geometry was encoded in WGS 84 (XYZ), we reproject the data to the Singapore SVY21 system (EPSG: 3414) using `st_transform()` and remove the Z component using `st_zm()`. This ensures suitability for spatial analysis and distance computation.

The same procedure is applied to all other KML and GeoJSON layers to ensure consistent and accurate spatial projection across datasets.

```{r}
communityclubs <- st_read("data/rawdata/CommunityClubs.kml")%>%
  st_transform(crs = 3414)%>%
  st_zm(drop = TRUE, what = "ZM")
```

```{r}
#| eval: false
#| echo: false
write_rds(communityclubs, "data/geospatial/communityclubs.rds")
```

```{r}
#| echo: false
communityclubs <- read_rds("data/geospatial/communityclubs.rds")
```

The map below illustrates the distribution of community clubs across Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(communityclubs) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("Community Clubs in Singapore") +
  tm_layout(frame = FALSE)
```

#### Childcare Centres

Here we import the Childcare kml file and tidy up by standardising the CRS and geometry dimensions.

```{r}
childcares <- st_read("data/rawdata/ChildCareServices.kml")%>%
  st_transform(crs = 3414)%>%
  st_zm(drop = TRUE, what = "ZM")
```

```{r}
#| eval: false
#| echo: false
write_rds(childcares, "data/geospatial/childcares.rds")
```

```{r}
#| echo: false
childcares <- read_rds("data/geospatial/childcares.rds")
```

The map below shows the geographic distribution of 1,925 childcare centres.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha= 0.6) +
tm_shape(childcares) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("Childcare Centres in Singapore") +
  tm_layout(frame = FALSE)
```

#### Clinics

The code chunk below is used to correctly import the clinics KML file. CHA clinics provide subsidised general practitioner (GP) and dental services across Singapore.

```{r}
clinics <- st_read("data/rawdata/CHASClinics.kml")%>%
  st_transform(crs = 3414)%>%
  st_zm(drop = TRUE, what = "ZM")
```

```{r}
#| eval: false
#| echo: false
write_rds(clinics, "data/geospatial/clinics.rds")
```

```{r}
#| echo: false
clinics <- read_rds("data/geospatial/clinics.rds")
```

The map below shows the locations of 1,193 CHAS clinics in Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha= 0.6) +
tm_shape(clinics) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source:data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("CHAS Clinics in Singapore") +
  tm_layout(frame = FALSE)
```

#### Eldercare Centres

The code chunk below imports the elder care services file in ESRI shapefile format.

```{r}
eldercares <- st_read(dsn = "data/rawdata", layer = "ELDERCARE")%>%
  st_transform(crs = 3414)%>%
  st_zm(drop = TRUE, what = "ZM")
```

```{r}
#| eval: false
#| echo: false
write_rds(eldercares, "data/geospatial/eldercares.rds")
```

```{r}
#| echo: false
hawkers <- read_rds("data/geospatial/eldercares.rds")
```

There are 133 elder care centres in Singapore as shown in the map below.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6)+
tm_shape(eldercares) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source:data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("Eldercare Centres in Singapore") +
  tm_layout(frame = FALSE)
```

#### Hawker Centres

The following code imports the Hawker Centres KML dataset and performs the necessary preprocessing steps within the R environment.

```{r}
hawkers <- st_read("data/rawdata/HawkerCentresKML.kml")%>%
  st_transform(crs = 3414)%>%
  st_zm(drop = TRUE, what = "ZM")
```

```{r}
#| eval: false
#| echo: false
write_rds(hawkers, "data/geospatial/hawkers.rds")
```

```{r}
#| echo: false
hawkers <- read_rds("data/geospatial/hawkers.rds")
```

There are 125 hawker centres in Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(hawkers) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("Hawker Centres in Singapore") +
  tm_layout(frame= FALSE)
```

#### Kindergartens

In this step, we import the kindergarten dataset (KML format) into the R environment. While child care centres provide full-day care and early childhood services for infants up to six years of age, kindergartens generally operate shorter-hour programmes focused on early education for children aged three to six.

```{r}
kindergartens <- st_read("data/rawdata/Kindergartens.kml")%>%
  st_transform(crs = 3414)%>%
  st_zm(drop = TRUE, what = "ZM")
```

```{r}
#| eval: false
#| echo: false
write_rds(kindergartens, "data/geospatial/kindergartens.rds")
```

```{r}
#| echo: false
kindergartens <- read_rds("data/geospatial/kindergartens.rds")
```

This map displays where 448 kindergartens are located across Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(kindergartens) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("Kindergartens in Singapore")+ tm_layout(frame = FALSE)
```

#### Parks

Here, we load and process the KML file containing park locations into R.

```{r}
parks <- st_read("data/rawdata/parks.geojson")%>%
  st_transform(crs = 3414)%>%
  st_zm(drop = TRUE, what = "ZM") 
```

```{r}
#| eval: false
#| echo: false
write_rds(parks, "data/geospatial/parks.rds")
```

```{r}
#| echo: false
parks <- read_rds("data/geospatial/parks.rds")
```

The map visualises the geographic locations of 441 parks within Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(parks) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source:data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("Parks in Singapore") +
  tm_layout(frame = FALSE)
```

#### MRT & LRT Station Exits

We read and process the station exit KML file into R as a spatial dataset.

```{r}
mrts <- st_read("data/rawdata/LTAMRTStationExitKML.kml")%>%
  st_transform(crs = 3414)%>%
  st_zm(drop = TRUE, what = "ZM")
```

```{r}
#| eval: false
#| echo: false
write_rds(mrts, "data/geospatial/mrts.rds")
```

```{r}
#| echo: false
mrts <- read_rds("data/geospatial/mrts.rds")
```

The map below represents the spatial distribution of 563 station exit points in Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col= "gray50", border.col ="white", alpha = 0.6) +
tm_shape(mrts) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source:data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("MRT & LRT Station Exits in Singapore") +
  tm_layout(frame = FALSE)
```

#### Supermarkets

The supermarket data in KML format is loaded into R for further processing.

```{r}
supermarkets <- st_read("data/rawdata/SupermarketsKML.kml")%>%
  st_transform(crs = 3414)%>%
  st_zm(drop = TRUE, what = "ZM")
```

```{r}
#| eval: false
#| echo: false
write_rds(supermarkets, "data/geospatial/supermarkets.rds")
```

```{r}
#| echo: false
supermarkets <- read_rds("data/geospatial/supermarkets.rds")
```

The map depicts the spatial distribution of 526 supermarkets across Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col= "gray50", border.col ="white", alpha = 0.6) +
tm_shape(supermarkets) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source:data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("Supermarkets in Singapore") +
  tm_layout(frame = FALSE)
```
:::

### Locational Factors without Geographic Coordinates

In this section, we import the location-based datasets obtained from official government sources and Wikipedia. Since these datasets do not provide geographic coordinates, we perform geocoding to assign spatial coordinates to each location.

#### Geocoding Function

The geocoding function below is modified to extract coordinates from the OneMap API using location names, rather than the block-and-street inputs used earlier. Because name-based searches are more sensitive to text variations, additional preprocessing steps are performed to standardise location names before querying the API. As geocoding can be time-consuming, a progress bar is included in the function to monitor progress and estimated completion time (ETA). A demonstration of this procedure can be found in the **Good Primary Schools** section.

```{r}
#| code-fold: true
#| code-summary: "CODE"
geocode_one <- function(data) {
  data$LATITUDE <- NA
  data$LONGITUDE <- NA
  
  cat("Geocoding", nrow(data), "data...\n")
  
  # Create progress bar
  pb <- progress_bar$new(
    format = "[:bar] :percent | ETA: :eta | :current/:total | :name",
    total = nrow(data),
    clear = FALSE,
    width = 80
  )
  
  for (i in 1:nrow(data)) {
    data_name <- data[i, 1]
    
    pb$tick(tokens = list(name = substr(data_name, 1, 20)))
    
    temp_output <- get_coords(data_name)
    
    data$LATITUDE[i] <- as.numeric(temp_output$results.LATITUDE[1])
    data$LONGITUDE[i] <- as.numeric(temp_output$results.LONGITUDE[1])
  }
  
  successful <- sum(!is.na(data$LATITUDE) & !is.na(data$LONGITUDE))
  failed <- nrow(data) - successful
  
  cat("\nGeocoding completed:\n")
  cat("successful:", successful, "/", nrow(data), "\n")
  cat("failed:", failed, "/", nrow(data), "\n")
  
  if (failed > 0) {
    cat("Failed data:\n")
    failed_data <- data[is.na(data$LATITUDE) | is.na(data$LONGITUDE), 1]
    print(failed_data)
  }
  
  return(data)
}

```

```{r}
#| code-fold: true
#| code-summary: "CODE"
get_coords <- function(name) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  query <- list(
    "searchVal" = name,
    "returnGeom" = "Y",
    "getAddrDetails" = "N",
    "pageNum" = "1"
  )
  
  tryCatch({
    # Add delay to avoid rate limiting
    Sys.sleep(0.1)
    
    res <- GET(base_url, query = query)
    
    # Check if request was successful
    if (status_code(res) != 200) {
      warning(paste("HTTP error for", name, ":", status_code(res)))
      return(data.frame(results.LATITUDE = NA, results.LONGITUDE = NA))
    }
    
    restext <- content(res, as = "text", encoding = "UTF-8")
    
    # Check if response is valid JSON
    if (restext == "" || is.na(restext)) {
      warning(paste("Empty response for", name))
      return(data.frame(results.LATITUDE = NA, results.LONGITUDE = NA))
    }
    
    output <- fromJSON(restext)
    
    # Check if results are found
    if (is.null(output$results) || length(output$results) == 0) {
      warning(paste("No results found for", name))
      return(data.frame(results.LATITUDE = NA, results.LONGITUDE = NA))
    }
    
    # Extract coordinates from first result
    output <- output$results[1, ] %>%
      select(LATITUDE, LONGITUDE) %>%
      rename(results.LATITUDE = LATITUDE, results.LONGITUDE = LONGITUDE)
    
    return(output)
    
  }, error = function(e) {
    warning(paste("Error geocoding", name, ":", e$message))
    return(data.frame(results.LATITUDE = NA, results.LONGITUDE = NA))
  })
}
```

::: panel-tabset
#### Primary Schools

Easy access to We extracted a list of 179 primary schools from publicly-available `Generalinformationofschools.csv` (data.gov.sg) dataset (version updated on 24 Sep 2025). However, the Ministry of Education (MOE) directory lists 182 primary schools, indicating that three schools were missing: Catholic High School (Primary Section), CHIJ St. Nicholas Girls’ School (Primary Section), and Maris Stella High School (Primary Section). To ensure full coverage, we manually identified and added these three schools to our dataset.

The following code standardises primary school names, geocodes each school using the `geocode_one` function to obtain latitude and longitude, and subsequently transforms the coordinates into a projected CRS to enable precise distance measurement.

```{r}
#| eval: false
prisch<- read_csv("data/rawdata/prischmoe_rename.csv") %>%
  mutate(name = gsub("St\\.", "SAINT",name))

prisch <- geocode_one(prisch)
prisch <- st_as_sf(prisch,
                         coords = c("LONGITUDE", "LATITUDE"),
                         crs = 4326) %>%
  st_transform(crs = 3414)
```

```{r}
#| eval: false
#| echo: false
write_rds(prisch, "data/geospatial/prisch.rds")
```

```{r}
#| echo: false
prisch <- read_rds("data/geospatial/prisch.rds")
```

The following table presents the first 10 primary school records, including their geocoded locations.

```{r}
kable(head(prisch,10), caption = "Preview of Primary School Locations")
```

The map below demonstrates the spatial arrangement of 182 primary schools across Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(prisch) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: MOE",
            position = c("right", "bottom")) +
  tm_title("Primary Schools in Singapore") +
  tm_layout(frame = FALSE)
```

#### Good Primary Schools

Access to good primary schools is an important consideration for many families when purchasing a home. In Singapore, education is highly valued, living within the catchment area of a strong school can therefore be an advantage for families planning their children’s long-term development.

However Singapore does not publish an official ranking of primary schools to avoid unhealthy competition. As a result, parents often rely on alternative indicators. One commonly referenced metric is the popularity score, derived from the Phase 2B registration ratio (i.e., number of applicants divided by the number of available places), which serves as a proxy for school demand and perceived quality.

For this study, we identify the top 20 primary schools using popularity scores sourced from a local education resource [Creative Campus](https://www.creativecampus.com.sg/best-primary-schools-in-singapore-2024). Only schools with a popularity score of 1.7 or higher were included. These schools are widely regarded as offering enriched learning environments and are highly sought after by parents and students.

The following code applies the same procedure as before: importing the dataset, geocoding the locations, transforming the coordinates to the designated projected CRS, and storing the output as an `sf` object.

```{r}
poppris <- read_csv("data/rawdata/top_primary_schools.csv") %>%
  mutate(School = gsub("ST\\.", "SAINT",School))

poppris <- geocode_one(poppris)
poppris <- st_as_sf(poppris,
                    coords = c("LONGITUDE", "LATITUDE"),
                    crs = 4326) %>%
  st_transform(crs = 3414)
```

```{r}
#| eval: false
#| echo: false
write_rds(poppris, "data/geospatial/poppris.rds")
```

```{r}
#| echo: false
poppris <- read_rds("data/geospatial/poppris.rds")
```

The following table presents the first 10 records of the identified top primary schools, including their geocoded locations.

```{r}
kable(head(poppris,10), caption = "Preview of Good Primary School Locations")
```

The 20 primary schools by popularity score are displayed in the map below.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(poppris) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: creativecampus.com.sg",
            position = c("right", "bottom")) +
  tm_title("Good Primary Schools in Singapore") +
  tm_layout(frame = FALSE)
```

#### Shopping Malls

Next, we import and process the dataset of 90 shopping malls in Singapore, which was sourced from Wikipedia on 29 October 2025.

```{r}
#| eval: false
malls <- read_csv("data/rawdata/malls.csv") 

malls <- geocode_one(malls)
malls <- st_as_sf(malls,
                         coords = c("LONGITUDE", "LATITUDE"),
                         crs = 4326) %>%
  st_transform(crs = 3414)
```

```{r}
#| eval: false
#| echo: false
write_rds(malls, "data/geospatial/malls.rds")
```

```{r}
#| echo: false
malls <- read_rds("data/geospatial/malls.rds")
```

The preview table shows the first 10 shopping malls and their locations.

```{r}
kable(head(malls,10), caption = "Preview of Shopping Mall Locations")
```

The map below shows the spatial locations of 90 shopping malls across Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(malls) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: wikipedia",
            position = c("right", "bottom")) +
  tm_title("Shopping Malls in Singapore") +
  tm_layout(frame = FALSE)
```

#### Hospitals

Next, we import and process data on 27 general hospitals in Singapore, obtained from Wikipedia on 29 October 2025. The Institute of Mental Health (IMH) in Hougang is excluded, as it serves as a specialised facility for mental health treatment rather than a general hospital providing general healthcare services.

```{r}
#| eval: false
hospitals <- read_csv("data/rawdata/hospitals.csv")

hospitals <- geocode_one(hospitals)
hospitals <- st_as_sf(hospitals,
                         coords = c("LONGITUDE", "LATITUDE"),
                         crs = 4326) %>%
  st_transform(crs = 3414)
```

```{r}
#| eval: false
#| echo: false
write_rds(hospitals, "data/geospatial/hospitals.rds")
```

```{r}
#| echo: false
hospitals <- read_rds("data/geospatial/hospitals.rds")
```

The preview table shows the first 10 hospital locations.

```{r}
kable(head(hospitals,10), caption = "Preview of Hospital Locations")
```

The map below illustrates the locations of 27 hospitals across Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(hospitals) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: wikipedia",
            position = c("right", "bottom")) +
  tm_title("Hospitals in Singapore") +
  tm_layout(frame = FALSE)
```

#### Sports Facilities

Regular exercise contributes to both physical and mental well-being. In Singapore, the government provides extensive, affordable sports facilities through ActiveSG, making recreational sports accessible to the public. Dual-use school facilities are excluded from this study, as they are primarily located within primary schools and are already captured under the primary school variable.

We manually compiled a list of 104 facilities from the ActiveSG Circle website (accessed 28 October 2025), including:

-   18 sports halls

-   24 swimming pools

-   13 stadiums

-   28 gyms

-   21 sports centres

After careful cross-checking and consolidation, we identified **52** unique sports locations. Facilities housed within the same complex were counted only once; for example, an ActiveSG gym located inside a larger sports centre was considered part of the same location to avoid duplication.

The code chunk below imports and processes the sports facilities dataset.

```{r}
#| eval: false
sports <- read_csv("data/rawdata/sports.csv")

sports <- geocode_one(sports)
sports <- st_as_sf(sports,
                         coords = c("LONGITUDE", "LATITUDE"),
                         crs = 4326) %>%
  st_transform(crs = 3414)
```

```{r}
#| eval: false
#| echo: false
write_rds(sports, "data/geospatial/sports.rds")
```

```{r}
#| echo: false
sports <- read_rds("data/geospatial/sports.rds")
```

The preview table shows the first 10 hospital locations.

```{r}
kable(head(sports,10), caption = "Preview of Sports Locations")
```

The map below represents the 52 Sports centre/complex in Singapore.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(sports) +
  tm_dots(fill = "red", size = 0.3, shape =17, lwd = 2) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: wikipedia",
            position = c("right", "bottom")) +
  tm_title("Sports Facilities in Singapore") +
  tm_layout(frame = FALSE)
```

#### CBD

Easy access to the Central Business District (CBD) is an important consideration when choosing where to live, as many residents commute to the city centre for work and services. In Singapore, the CBD is located within the Downtown Core planning area of the Central Region. We obtained the CBD coordinates (1.287953, 103.851784) from Google and stored the location as an `sf` point object in the projected SVY21 coordinate system (EPSG:3414).

```{r}
cbd <- data.frame(name = "CBD", LATITUDE = 1.287953, LONGITUDE = 103.851784) %>%
  st_as_sf(coords = c("LONGITUDE","LATITUDE"), crs = 4326) %>%
  st_transform(crs = 3414)
```

```{r}
cbd
```

```{r}
#| eval: false
#| echo: false
write_rds(cbd, "data/geospatial/cbd.rds")
```

```{r}
#| echo: false
cbd <- read_rds("data/geospatial/cbd.rds")
```

The extracted map shows the Central Region, where the CBD location is marked with a red diamond, and the Downtown Core is highlighted in yellow.

```{r}
#| code-fold: true
#| code-summary: "CODE"
central_region <- mpsz[mpsz$REGION_N=="CENTRAL REGION", ]
downtown_core <- mpsz[mpsz$PLN_AREA_N=="DOWNTOWN CORE", ]

tm_shape(central_region) +
  tm_polygons(col = "gray50", border.col = "white", alpha = 0.6) +
tm_shape(downtown_core) +
  tm_polygons(col = "orange", alpha = 0.3, border.col = "red",border.lwd = 2) +
tm_shape(cbd) +
  tm_dots(col = "red", size = 1.2, shape = 18,border.col = "darkred",border.lwd = 1.5) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Google & data.gov.sg",
            position = c("left", "bottom")) +
  tm_title("Singapore Central Business District") +
   tm_layout(frame = FALSE)
```

```{r}
#| echo: false
HDBresale_sf <- read_rds("data/geospatial/HDBresale_sf.rds")
```
:::

## Data Wrangling

In this section, we further prepare the independent variables defined earlier in the study.

### Checking of Overlapping Points

Before fitting the GWR model, it is necessary to ensure that no unintended point-overlap occurs among the spatial datasets (e.g., HDB locations or public amenities). Overlapping coordinates may indicate geocoding inaccuracies or duplicated facility records, which could distort local parameter estimates. Accordingly, we assess point overlap across all spatial layers:`cbd`, `eldercare` , `hawker` , `mrts`, `parks`, `poppri`, `malls`, `hospitals`, `communityclubs`, `sports`, `childcares`, `kindergartens`, `clinics`, `BusStop_in_SG`, `prisch`, and HDB resale data (`HDBresale_sf`).

We differentiate genuine co-located features from unintended duplicates.Then a helper function below employs `st_equals()` to detect identical geometries; the presence of `TRUE` (or `1`) values indicates overlapping points. When overlaps are identified, `st_jitter(amount = 1)` is applied to nudge points by one metre, ensuring spatial separation while maintaining positional accuracy.

```{r}
#| code-fold: true
#| code-summary: "CODE"
fix_overlapping_points <- function(var_list) {
  cat("Checking for overlapping points in", length(var_list), "datasets...\n")

  
  for (i in seq_along(var_list)) {
    dataset_name <- names(var_list)[i]
    sf_data <- var_list[[i]]
    
    # Check for overlapping points
    overlapping_points <- sf_data %>%
      mutate(overlap = lengths(st_equals(., .)) > 1)
    
    overlap_count <- sum(overlapping_points$overlap)
    
    if (overlap_count > 0) {
      cat(dataset_name, "-", overlap_count, "overlapping points found\n")
      
      # Apply jitter 1m
      sf_data <- sf_data %>%
        st_jitter(amount = 1)
      
      # Update the dataset in the list
      var_list[[i]] <- sf_data
      cat("Jitter applied to", dataset_name, "\n")
    }
  }
  
  cat("Overlap fixing completed!\n")
  return(var_list)
}
```

The following code loads all previously processed spatial datasets stored in the *data/geospatial* folder (in `.rds` format) and compile them into a list named `var_list` to streamline subsequent processing steps.

```{r}
#| code-fold: true
#| code-summary: "CODE"
# Get all .rds files in the geospatial directory
rds_files <- list.files("data/geospatial", pattern = "\\.rds$", full.names = TRUE)

# Create empty list
var_list <- list()

# Load each RDS file
for (file in rds_files) {
  # Extract filename without extension for the list name
  var_name <- gsub("\\.rds$", "", basename(file))  
  # Load the data and assign to list
  var_list[[var_name]] <- read_rds(file)
  
  cat("Loaded:", var_name, "\n")
}

cat("Total datasets loaded:", length(var_list), "\n")
```

We then apply the `fix_overlapping_points` function to slightly separate points that share the same location.

```{r}
var_list <- fix_overlapping_points(var_list)
```

The following code unpacks the processed `sf` objects into the R environment.

```{r}
# Unpack all objects into the global environment
list2env(var_list, envir = .GlobalEnv)
```

In addition, we re-examine the `HDBresale_sf` dataset to confirm that no overlapping coordinates exist, as the overlap count returned zero.

```{r}
overlapping_points <- HDBresale_sf %>%
  mutate(overlap = lengths(st_equals(., .)) >1)
sum(overlapping_points$overlap)
```

### Creating Structural Factors

In this section, we reconstruct key attributes from `HDBresale_sf` to derive structural factors that could potentially affect resale prices. The processed variables are `storey_range` and `remaining_lease`.

In addition, the completion status of the Upgrading Programme (`MUP/HIP`) is included in the model specification.

::: panel-tabset
#### Floor Level

As previously noted, the `storey_range` field in the HDBresale dataset is stored as a text string (e.g., “04 TO 06”). Floor height is considered a relevant explanatory variable, as higher floors may contribute to price premiums due to reduced noise exposure and improved views. To use this variable for modelling, we transform each range into a numeric value by computing its midpoint (e.g., “04 TO 06” to 5).

The following code displays the original text strings in the dataset. The `storey_range` values span from "01 TO 03" to "49 TO 51".

```{r}
sort(unique(HDBresale_sf$storey_range))
```

The code below creates a new column, `FLOOR`, and uses `st_extract`() to extract the first and last numeric values from the `storey_range` field. These values are then averaged (divided by two) to obtain the **midpoint** of the storey range.

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    FLOOR = (as.numeric(str_extract(storey_range, "^[0-9]+")) +
                     as.numeric(str_extract(storey_range, "[0-9]+$"))) / 2
  )
```

Following the transformation, the `FLOOR` variable ranges from 2 to 50, accurately reflecting the storey range in the original dataset (from “01 TO 03” to “49 TO 51”).

```{r}
summary(HDBresale_sf$FLOOR)
```

#### Remaining Lease

The age of a property is another important structural factor that may influence its value. Newer flats are often more desirable as they tend to be in better physical condition and have more modern layouts. To account for this, we include property age as an explanatory variable in our model.

In Singapore, HDB flats are sold on a 99-year leasehold basis. Within the HDBresale dataset, lease information is provided in two formats: `remaining_lease`, which appears as text (e.g., “53 years 1 month”), and `lease_commence_date`, which specifies the lease start year. As these variables represent equivalent information, we convert `remaining_lease` into a numeric measure by calculating the total number of months remaining.

```{r}
head(HDBresale_sf$remaining_lease,4)
```

The code chunk below extracts the numeric values before the text “years” and “months” and stores the calculated lease duration (in months) in a new column, `LEASE.` Since there are variations in the `remaining_lease` text format (e.g., no month specified, “1 month”, or multiple months), we include an `if_else()` condition in the code: if the month value is `NA`, it is replaced with `0`; otherwise, the extracted month value is used.

```{r}
HDBresale_sf <- HDBresale_sf %>% 
  mutate(
    lease_years = if_else(
      is.na(str_extract(remaining_lease, "\\d+(?= years)")), 
      0, 
      as.numeric(str_extract(remaining_lease, "\\d+(?= years)"))),
    lease_months = if_else(
      is.na(str_extract(remaining_lease, "\\d+(?= month)")), 
      0, 
      as.numeric(str_extract(remaining_lease, "\\d+(?= month)"))),
    LEASE = lease_years * 12 + lease_months)
```

The following code The output below verifies that the transformed lease variable spans from 485 months to 1,144 months, consistent with the expected lease durations in the dataset.

```{r}
summary(HDBresale_sf$LEASE)
```

#### Age of the unit in Months

The sales records indicate that the median TOP year for flats in the dataset is 2002. The earliest completion year observed is 1967 (approximately 58 years old as of 2025), whereas the most recent completions are around 4 years old. Under HDB regulations, most owners must fulfil a Minimum Occupation Period (MOP) of 5 years before they can sell their flats. However, upon reviewing the data and cross-checking certain transactions, we observed that several units completed in 2020/2021 have already been resold. This implies that exceptions to the MOP rule may exist, such as special approval cases or specific scheme-based exemptions.

```{r}
summary(HDBresale_sf$lease_commence_date)
```

Following the previous step, we then subtract `LEASE` from 99 years to derive the property’s age in months at the time of the transaction.

```{r}
HDBresale_sf <- HDBresale_sf %>% 
  mutate(AGE = 99 * 12 -LEASE)
```

The output below shows that the property age ranges from 44 months to 703 months, which is consistent with the values in the original dataset.

```{r}
summary(HDBresale_sf$AGE)
```

#### Main Upgrading Program (MUP) completed

As discussed earlier, internal conditions are an important structural factor in property valuation. For many flats in our dataset,typically around 20 years old, renovation can meaningfully enhance market value. While private renovation details are not available, government upgrading programmes such as the Main Upgrading Programme (MUP) and Home Improvement Programme (HIP) provide useful proxies for housing condition.

Flats that have undergone MUP or HIP are assigned a value of `1` in the corresponding variables. We then perform a left-join between the MUP/HIP dataset and `HDBresale_sf` using the `block`, `street_name`, and `town` as the joining keys. We assume that all units within a block benefited from the upgrading works if the block appears in the programme list. However, because MUP was fully completed by 2012, its effect on today’s resale prices may have weakened.

```{r}
HDBresale_sf <- left_join(HDBresale_sf, HIP, by = c('block', 'street_name', 'town'))
```

```{r}
length(unique(HDBresale_sf$geometry))
```

```{r}
sum(HDBresale_sf$HIP)
```

```{r}
sum(HDBresale_sf$MUP)
```

Our results show that 2,967 flats in the dataset have undergone HIP and 257 flats have undergone MUP, accounting for approximately 37% of the total resale transactions.

```{r}
(sum(HDBresale_sf$MUP) + sum(HDBresale_sf$HIP)) /length(unique(HDBresale_sf$geometry))
```

Note: there are also 30 flats in the dataset that have received both HIP and MUP upgrades.

```{r}
sum(HDBresale_sf$HIP == 1 & HDBresale_sf$MUP ==1 )
```
:::

### Locational Factors

Location strongly influences property value. Homes close to schools, transport, healthcare, shopping, and recreation offer greater convenience, community access, and easier CBD commuting.

In this section, we construct two sets of locational variables:

-   Proximity-based measures: distance from each flat to key amenities (`cbd`, `eldercares`,`hawkers`, `mrts`, `parks`, `poppris`, `malls`, `hospitals`, `communityclubs`, `sports`)

-   Accessibility counts: number of nearby facilities within a specified radius ( `childcares`, `kindergartens`, `clinics`,`busstops` within 350m or `prisch` within 1 km).

#### Proximity

We use `st_distance()` to compute the distance between each HDB resale flat and the amenities listed above. Given that each computation requires roughly two minutes, a progress bar is incorporated into the function to track execution progress.

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "CODE"
proximity <- function(data, variable, varname) {
  
  cat("Calculating proximity for", nrow(data), "locations...\n")
  
  # Create progress bar
  pb <- progress_bar$new(
    format = "[:bar] :percent | ETA: :eta | :current/:total",
    total = nrow(data),
    clear = FALSE,
    width = 80
  )
  
  # Initialise column
  data[[varname]] <- NA
  
  for (i in 1:nrow(data)) {
    # Calculate distance from current data point to all variable points
    dis <- st_distance(data[i, ], variable)
    # Find minimum distance and store it
    data[[varname]][i] <- as.numeric(min(dis))
    
    # Update progress bar
    pb$tick(tokens = list(name = varname))
  }
  
  # Calculate success/failure stats
  successful <- sum(!is.na(data[[varname]]))
  failed <- sum(is.na(data[[varname]]))
  
  cat("\nProximity calculation completed:\n")
  cat("successful:", successful, "/", nrow(data), "\n")
  cat("failed:", failed, "/", nrow(data), "\n")
  
  if (failed > 0) {
    cat("Failed rows (indices):\n")
    failed_indices <- which(is.na(data[[varname]]))
    print(failed_indices)
  }
  
  return(data)
}
```

::: panel-tabset
##### CBD

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, cbd, "DIS_TO_CBD")
```

##### Eldercare Centres

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, eldercares, "DIS_TO_ELDERCARE")
```

##### Hawker Centres

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, hawkers, "DIS_TO_HAWKER")
```

##### MRTs &LRTs

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, mrts, "DIS_TO_MRT")
```

##### Good Primary Schools

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, poppris, "DIS_TO_POPPRIM")
```

##### Shopping Malls

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, malls, "DIS_TO_MALL")
```

##### Community Clubs

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, communityclubs, "DIS_TO_COMMUNITYCLUB")
```

##### Hospitals

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, hospitals, "DIS_TO_HOSPITAL")
```

##### Sports Facilities

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, sports, "DIS_TO_SPORTS")
```

##### Parks

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, parks, "DIS_TO_PARK")
```

##### Supermarkets

```{r}
#| eval: false
HDBresale_sf <- proximity(HDBresale_sf, supermarkets, "DIS_TO_SUPERMARKET")
```
:::

#### Number of Nearby Facilicites

The following code constructs a function that counts the number of facilities located within predefined buffer radii from each HDB flat. To track execution progress, a progress bar is incorporated into the procedure. Each computation takes approximately three minutes to complete.

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "CODE"
countnearby <- function(data, variable, varname, distance) {
  
  cat("Counting facilicities within", distance,"m for", nrow(data), "locations...\n")
  
  # Create progress bar
  pb <- progress_bar$new(
    format = "[:bar] :percent | ETA: :eta | :current/:total",
    total = nrow(data),
    clear = FALSE,
    width = 80
  )
  
  # Initialise column
  data[[varname]] <- NA
  
  for (i in 1:nrow(data)) {
   # Calculate which facilities are within distance for this specific point
  within_ <- st_is_within_distance(data[i,], variable, dist = distance)
   # Count the number of facilities within distance
  data[[varname]][i] <- length(within_[[1]])
    
    # Update progress bar
    pb$tick(tokens = list(name = varname))
  }
  
  # Calculate success/failure stats
  successful <- sum(!is.na(data[[varname]]))
  failed <- sum(is.na(data[[varname]]))
  
  cat("\nCount completed:\n")
  cat("successful:", successful, "/", nrow(data), "\n")
  cat("failed:", failed, "/", nrow(data), "\n")
  
  if (failed > 0) {
    cat("Failed rows (indices):\n")
    failed_indices <- which(is.na(data[[varname]]))
    print(failed_indices)
  }
  
  return(data)
}
```

::: panel-tabset
##### Childcare Centres within 350 Metres

```{r}
#| eval: false
HDBresale_sf <- countnearby(HDBresale_sf, childcares, "N_CHILDCARE_350M", 350)
```

##### Kindergartens within 350 Metres

```{r}
#| eval: false
HDBresale_sf <- countnearby(HDBresale_sf, kindergartens, "N_KINDY_350M", 350)
```

##### CHAS Clinics within 350 Metres

```{r}
#| eval: false
HDBresale_sf <- countnearby(HDBresale_sf, clinics, "N_CLINIC_350M", 350)
```

##### Bus Stops within 350 Metres

```{r}
#| eval: false
HDBresale_sf <- countnearby(HDBresale_sf, BusStop_in_SG, "N_BUS_350M", 350)
```

##### Primary Schools within 1 Kilometre

```{r}
#| eval: false
HDBresale_sf <- countnearby(HDBresale_sf, prisch, "N_PRI_1KM", 1000)
```

```{r}
#| eval: false
#| echo: false
write_rds(HDBresale_sf, "data/rds/HDBresale_sf.rds")
```

```{r}
#| echo: false
HDBresale_sf <- read_rds("data/rds/HDBresale_sf.rds")
```
:::

#### Pre-modelling Data Checks

In this step, we verify that the prepared `HDBresale_sf` dataset contains no missing values, and we perform variable renaming and selection to retain only the fields required for modelling. The processed dataset is then saved and renamed as `HDBresale` for efficient coding and analysis.

First the code chunk below shows there is no missing value in the `HDBresale_sf` dataset.

```{r}
sum(is.na(HDBresale_sf))
```

Columns such as `flat_type`, `block`, `street_name`, and `storey_range` etc are either pre-processed or redundant. Therefore, we prepare a dataset that contains only the variables required for modelling.

```{r}
colnames(HDBresale_sf)
```

Here we select the required columns by using the `select()` function.

```{r}
HDBresale_final <- HDBresale_sf %>% select(2,12,11,13,18,19,17,16,7,20:35)
```

For consistency, we convert all selected column names to uppercase to distinguish them from the original dataset.

```{r}
HDBresale_final <- HDBresale_final %>%
  rename(AREA_SQM = floor_area_sqm) %>%
  rename_with(toupper)  # Convert all column names to uppercase
```

The dataset consists of 25 columns. The variable `RESALE_PRICE` serves as the dependent variable, while the remaining variables from the fourth column onwards are the independent variables used for explanatory modelling.

```{r}
colnames(HDBresale_final)
```

The dataset is saved and renamed as `HDBresale`, and a validation check confirms that no missing values are present in the loaded dataset.

```{r}
#| eval: false
#| echo: false
write_rds(HDBresale_final, "data/rds/HDBresale_final.rds")
```

```{r}
HDBresale <- read_rds("data/rds/HDBresale_final.rds")
```

```{r}
sum(is.na(HDBresale))
```

We quickly generate summary statistics for each column.

```{r}
summary(HDBresale)
```

## Data Visualisation

With the data prepared and the majority of independent variables processed, a preliminary visual check is essential before progressing to modelling. This section presents plots generated using the **ggplot** package to examine the distribution of each variable.

### HDB Resale Overview: January to September 2025

From January 2025 to September 2025, 3-room, 4-room, and 5-room flats were the top three transacted flat types. Among them, 4-room units recorded the highest number of transactions (8,679), likely due to their adequate supply and practical size (approximately 90–100 sqm), which suits families with one to two children, a typical demographic profile in Singapore.

```{r}
count <- ggplot(HDBresale_all) +
  geom_histogram(aes(x=flat_type), stat= "count", color = "black", fill = "lightblue") +
  labs(title = "Distribution of Flat Types" , x= "Flat Type", y = "Count")
count
```

The code below generates a series of interactive boxplots of average monthly resale prices by flat type. As expected, 4-room flats fall near the middle of the price range across all HDB flat types.

```{r}
summary_data <- HDBresale_all %>%
  mutate(month = as.Date(paste0(month, "-01"))) %>%  # Convert to proper Date
  group_by(month, flat_type) %>%
  summarize(avg_resale_price = mean(resale_price),
            .groups = 'drop')

summary_boxplot <- ggplot(summary_data, aes(x = flat_type, y = avg_resale_price, fill = flat_type)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(
    labels = scales::comma_format(),
    breaks = scales::pretty_breaks()
  ) +
  labs(title = "Distribution of Average Resale Prices by Flat Type in 2025", 
       x = "Flat Type", 
       y = "Average Resale Price",
       fill = "Flat Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))  # Rotate x-axis labels

ggplotly(summary_boxplot)
```

### HDB Resale 4-ROOM Flats: January to September 2025

This section explores the distributional characteristics of the 4-room flat dataset. A helper function is implemented to produce histograms for all variables, and the plots are arranged into grids using `ggarrange()` from the **ggpubr** package when necessary.

```{r}
plot_histograms <- function(col) {
  ggplot(HDBresale, aes(x = .data[[col]])) +
    geom_histogram(bins = 20, color = "black", fill = "lightblue") +
    ggtitle(paste("Distribution of", col)) +
    xlab(col) 
}
```

#### Resale Price

##### Distribution Analysis

The code chunk below creates the histogram for distribution of `RESALE_PRICE` by using the helper function `plot_histogram()`.

```{r, fig.height=4, fig.width=10}
plot_histograms("RESALE_PRICE") %>%
  ggplotly()
```

The histogram indicates that the distribution of resale prices for 4-room flats is **right-skewed**, meaning most units are sold within the lower price range, with fewer high-value transactions at the upper end.

A common way to address right-skewed data is to apply a **log** transformation, which helps normalise the distribution. The code chunk below demonstrates this by creating a new variable, `LOG_RESALE_PRICE`, using `mutate()` from the **dplyr** package and `log()` from **base R**.

Although a log transformation is not compulsory, it can improve model robustness by reducing skewness and stabilising variance. In the log-transformed model, coefficients are interpreted in percentage terms: for example, a one-unit increase in a explanatory variable corresponds to the percentage change in resale price. Thus, a one-square-metre increase in floor area would be interpreted as resulting in an approximate x% change in resale price, holding all other variables constant.

```{r}
HDBresale <- HDBresale %>%
  mutate(`LOG_RESALE_PRICE` = log(RESALE_PRICE))
```

```{r}
plot_histograms("LOG_RESALE_PRICE") 
```

```{r}
#| echo: false
#| eval: false
write_rds(HDBresale, "data/rds/HDBresalelg.rds")
```

```{r}
#| echo: false
HDBresale <-read_rds("data/rds/HDBresalelg.rds")
```

The distribution becomes less skewed after the transformation.

##### Spatial Visualisation

The following code generates a map illustrating the spatial distribution of HDB resale prices in Singapore using functions from **tmap** package.

```{r}
#| code-fold: true
#| code-summary: "CODE"
rp<- tm_shape(mpsz) +
  tm_polygons(fill = "gray50",
              border.col = "white", 
              alpha = 0.6) +
tm_shape(HDBresale) +    #event points
  tm_dots(fill = "RESALE_PRICE",
          fill_alpha = 0.6,
          size = 0.2,
          fill.scale = tm_scale_intervals(
            style = "quantile", n = 5, values = "-brewer.rd_bu")) +
  tm_view(set_zoom_limits = c(11,14)) +
  tm_compass(type = "8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("HDB Resale Prices in Singapore") +
  tm_layout(frame = FALSE)
rp
```

#### Multiple Histogram Plots: Distribution of Independent Variables

We utilise the helper function to draw trellis plots (multiple histograms) for the independent variables and arrange them into a grid.

The code below organises the explanatory variables into lists by their characteristics, structural and locational, to streamline plotting.

```{r}
names(HDBresale)
structural<- names(HDBresale)[4:9]
structural
locational <- names(HDBresale)[10:25]
locational
```

##### Structral Factors

The code below produces bar charts showing the distribution of structural factors. We observe that the majority of resale flats are located below the 20th storey. Very few blocks received MUP, while about one-third underwent HIP. It is high chance that MUP will not contribute much explanatory power.

```{r, fig.height=4, fig.width=10}
plots <- lapply(structural, plot_histograms)
ggarrange(plotlist = plots, ncol = 3, nrow = 2) 
```

The distributions of `AGE` and `LEASE` mirror each other, as one is derived from the other. Since they are highly correlated, only one should be retained in the model to avoid multicollinearity, that is validated through statistical diagnostics later. The `AREA_SQM` distribution also shows that most 4-room flats are approximately 90 sqm in size.

##### Locational Factors

The histograms below illustrate the spatial accessibility characteristics of a typical 4-room HDB resale flat. Most units are located within 15 km of the CBD, although the distribution is left-skewed, indicating that only a smaller proportion of flats are very close to the CBD.

-   Eldercare centres: mostly within 800 m

-   Hawker centres: within 500 m

-   MRT stations: typically within 500 m

-   Top primary schools: within 2 km

-   Shopping malls: generally within 1 km

-   Community clubs: around 500 m

-   Hospitals: within 2 km

-   Sports complexes: often within 1 km

-   Parks: mostly within 800 m

-   Supermarkets: usually within 300 m

-   Buffer-based amenity counts also highlight high urban accessibility:

-   Childcare centres (350 m): typically 4–5

-   Kindergartens (350 m): usually 1–2

-   Clinics (350 m): commonly \~3

-   Bus stops (350 m): typically 7–8

-   Primary schools (1 km): generally 2–3

Overall it indicates a generally thoughtful planning pattern across residential areas, with many flats having reasonable access to essential amenities.

```{r, fig.height=10, fig.width=16}
          
plots <- lapply(locational, plot_histograms)
ggarrange(plotlist = plots, ncol = 4, nrow = 4) 
```

## Hedonic Pricing Modelling

In this section, we develop the following modelling approaches. First, we build a Multiple Linear Regression (MLR) model using the `lm()` function in **base R**, followed by a Geographically Weighted Regression (GWR) model using the **GWmodel** package to account for potential spatial variation in coefficients. These models form the core of our explanatory framework, as they provide interpretable parameter estimates and statistical significance measures.

To complement the explanatory models and capture potential non-linear relationships, we also introduce a Random Forest model. The non-spatial and spatial Random Forest is implemented using **SpatialRF** package. Random Forest aggregates multiple decision trees and does not require prior feature transformation, making it well suited for modelling complex interactions.

Our primary objective is explanatory modelling, therefore, MLR and GWR serve as the main analytical tools. The Random Forest models are included not as causal inference methods, but to validate variable importance patterns, ensuring robustness in the interpretation of key determinants of HDB resale prices.

### Multiple Linear Regression

In this section, we first conduct multicollinearity tests to identify and remove highly correlated explanatory variable. Once the final variable set is determined, we fit the model and evaluate its robustness by verifying adherence to standard regression assumptions.

The general form of a Multiple Linear Regression (MLR) model is shown below. MLR expresses the dependent variable as a linear combination of multiple independent variables, plus an error term to account for unexplained variation.

$$Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \cdots + \beta_kX_k + \varepsilon$$

#### Computing Correlation Matrix

Before building the multiple linear regression model, we assess multicollinearity to ensure that the independent variables are not highly correlated with one another. High multicollinearity suggests redundancy among predictors and can distort coefficient estimates, resulting in unstable and unreliable model outcomes.

To examine this, we use the [`ggcorrmat()`](https://indrajeetpatil.github.io/ggstatsplot/articles/web_only/ggcorrmat.html) function from the **ggstatsplot** package to visualise the correlation matrix of the variables. The geometry column is excluded from this analysis, as most statistical plotting functions do not operate on spatial objects. A commonly accepted rule of thumb is that correlation values above 0.8 indicate potential multicollinearity concerns.

```{r, fig.width=14, fig.height=10}
HDBresale_nogeo <- HDBresale %>%
  st_drop_geometry()
ggstatsplot::ggcorrmat(HDBresale_nogeo[, 2:25])
```

The output above shows a triangular matrix of pairwise correlation values between variables.

| Correlation r value | Interpretation         |
|---------------------|------------------------|
| (0,1)               | positive relationships |
| =0                  | no correlation         |
| (-1,0)              | negative relationships |

The correlation matrix shows that `AGE` and `LEASE` are perfectly negatively correlated (r = –1), which is expected as `AGE` was derived directly from `LEASE` using the 99-year lease rule. To avoid redundancy, only one of these variables should be included in the model. We retain `AGE`, as it is more intuitive and interpretable.

In addition, several variables display extremely weak correlation with resale price (\|r\| ≤ 0.02), such as `N_KINDY_350M`, `DIS_TO_SPORTS`, and `DIS_TO_SUPERMARKET`. Although these variables exhibit minimal pairwise correlation with resale price, they are still included in the regression model. Their inclusion increases model performance (higher R^2^ and lower AIC), suggesting suppressor effects that they enhance explanatory power by accounting for variance not captured by other variables.

#### Building Baseline MLR model

Next we build the hedonic pricing models for HDB `RESALE_PRICE` using [`lm()`](https://www.rdocumentation.org/packages/stats/versions/3.5.2/topics/lm) of **base R**.

The following code estimates the multiple linear regression model with the selected explanatory variables: `FLOOR`, `HIP`, `MUP`, `AGE`, `AREA_SQM`, `DIS_TO_CBD`, `DIS_TO_ELDERCARE`, `DIS_TO_HAWKER`, `DIS_TO_MRT`, `DIS_TO_POPPRIM`, `DIS_TO_MALL`, `DIS_TO_COMMUNITYCLUB`, `DIS_TO_HOSPITAL`, `DIS_TO_PARK`, `N_CHILDCARE_350M`, `N_CLINIC_350M`, `N_BUS_350M`, `N_PRI_1M`, `N_KINDY_350M`, `DIS_TO_SPORTS`, and `DIS_TO_SUPERMARKET`.

```{r}
#| code-fold: true
#| code-summary: "CODE"
start_time <- Sys.time()
cat("Starting multiple linear regression at:", format(Sys.time(), "%H:%M:%S"), "\n")
hdb_mlr <- lm(
  RESALE_PRICE ~ FLOOR + HIP + MUP + AGE + AREA_SQM +
    DIS_TO_CBD + DIS_TO_ELDERCARE + DIS_TO_HAWKER + DIS_TO_MRT
    + DIS_TO_POPPRIM + DIS_TO_MALL +
    DIS_TO_COMMUNITYCLUB + DIS_TO_HOSPITAL + DIS_TO_PARK +
    N_CHILDCARE_350M + N_CLINIC_350M + N_BUS_350M + N_PRI_1KM + N_KINDY_350M + DIS_TO_SPORTS +DIS_TO_SUPERMARKET,
  data = HDBresale)

# End time and duration
end_time <- Sys.time()
duration <- end_time - start_time

cat("Model completed at:", format(end_time, "%Y-%m-%d %H:%M:%S"), "\n")
cat("Processing time:", round(duration, 2), attr(duration, "units"), "\n")
```

```{r}
#| eval: false
#| echo: false
write_rds(hdb_mlr, "data/model/hdb_mlr.rds")
```

```{r}
#| echo: false
hdb_mlr <-read_rds("data/model/hdb_mlr.rds")
```

In this section, we present the summary report of the `hdb_mlr` model, which provides the statistical results of the multiple linear regression analysis. Before interpreting the coefficients, we first assess the statistical validity of the model and the significance of the individual explanatory variables.

Two null hypotheses are evaluated:

-   H~0~ (overall model): The regression model is not statistically significant.

-   H~0~ (individual coefficients): Each regression coefficient equals zero (i.e., the explanatory variable has no effect).

if a p-value of 0.05 or less, we reject the null hypothesis at the 95% confidence level. As shown in the regression table, all p-values are below this threshold, confirming that both the overall model and all explanatory variables are statistically significant.

Although some independent variables exhibit weak pairwise correlation with the target variable, sensitivity testing below shows that the model performs best when these variables are retained (see comparison table below). This provides evidence of suppressor effects, where certain explanatory variables enhance model performance only in the presence of others.

```{r}
#| code-fold: true
#| code-summary: "CODE"
results_table <- data.frame(
  Model = c("Baseline", "Model 2", "Model 3", "Model 4"),
  Removed_variable = c("None", "SUPERMARKET", "SUPERMARKET + SPORTS", "SUPERMARKET + SPORTS + KINDY"),
  R2 = c(0.815, 0.814, 0.812, 0.811),
  Adj_R2 = c(0.815, 0.814, 0.811, 0.810),
  AIC = c(218087, 218134, 218243, 218285),
  sigma = c(69131, 69323, 69764, 69933),
  Interpretation = c("Best performance", "Tiny performance drop", "Worse", "Worst")
)

kable(results_table, caption = "Table: Model Performance Comparison with Sequential Removal of Weak Variables", format = "html", align = "c")
```

The code below generates a well-formatted regression summary table using [`tbl_regression()`](https://www.danieldsjoberg.com/gtsummary/reference/tbl_regression.html) from [**gtsummary**](https://www.danieldsjoberg.com/gtsummary/index.html) package. Additional model evaluation metrics are included using [`add_glance_table()`](https://www.danieldsjoberg.com/gtsummary/reference/add_glance.html).

```{r}
#| code-fold: true
#| code-summary: "CODE"
tbl_regression(hdb_mlr, intercept = TRUE) %>%
  bold_labels() %>%
  bold_p(t = 0.05) %>%
  add_glance_source_note(
    label = list(sigma~ "\U03C3"),
    include = c(r.squared, adj.r.squared, AIC, statistic, p.value, sigma)
  ) %>%
  modify_caption("Table: Baseline Model Statistical Summary")
  
```

The model achieves an R^2^ of 0.815 and an Adjusted R^2^ of 0.815, indicating that approximately 81.5% of the variation in 4-room HDB resale prices is explained by the selected variables.

Based on the estimated coefficients, the Multiple Linear Regression (MLR) model for HDB resale price can be expressed as follows:

ResalePrice = 634,871 + 5,982×FLOOR + 10,235×HIP - 48,303×MUP - 507×AGE + 4,583×AREA_SQM - 19.88×DIS_TO_CBD - 15.23×DIS_TO_ELDERCARE - 15.81×DIS_TO_HAWKER - 52.77×DIS_TO_MRT - 9.56×DIS_TO_POPPRIM - 17.35×DIS_TO_MALL - 15.60×DIS_TO_COMMUNITYCLUB + 9.91×DIS_TO_HOSPITAL - 9.12×DIS_TO_PARK - 2,594×N_CHILDCARE_350M + 6,432×N_CLINIC_350M + 1,265×N_BUS_350M - 11,340×N_PRI_1M + 6,321×N_KINDY_350M + 14.41×DIS_TO_SPORTS + 32.50×DIS_TO_SUPERMARKET

The MLR model indicates that both structural attributes (floor, size, age) and locational factors (proximity to MRT, CBD, amenities) are strong determinants of 4-room HDB resale prices. While most amenity-related effects are small in dollar term, upgrading programmes (HIP/MUP) and access to transport and education facilities play important roles in shaping valuation. For instance, when all other variables are held constant, the presence of HIP is associated with an average increase of approximately \$10,235 in HDB resale price.

#### Building Log-MLR Model

The following code fits a log-linear MLR model, where the dependent variable is the log-transformed resale price (`LOG_RESALE_PRICE`). The same explanatory variables used in the base model is retained for consistency.

The formula is as follows:

$$\log(Y_i) = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \cdots + \beta_kX_{ki} + \varepsilon_i$$

Where:

-   $Y_i$ = Resale price of HDB flat $i$

-   $\log(Y_i)$ = Natural logarithm of resale price

-   $\beta_0$ = Global intercept (average log price when all independent variables = 0)

-   $\beta_j$ = Global coefficient for independent variables $j$ ($j = 1, 2, \dots, k$)

-   $X_{ji}$ = Value of independent variables $j$ for observation $i$

-   $\varepsilon_i$ = Error term for observation $i$ \~ $N(0, \sigma^2)$

-   $i$ = Observation index ($i = 1, 2, \dots, n$)

-   $k$ = Number of independent variables

-   $n$ = Total number of observations (8,679)

```{r}
#| code-fold: true
#| code-summary: "CODE"
start_time <- Sys.time()
cat("Starting multiple linear regression log at:", format(Sys.time(), "%H:%M:%S"), "\n")
loghdb_mlr <- lm(
  LOG_RESALE_PRICE ~ FLOOR + HIP + MUP + AGE + AREA_SQM +
    DIS_TO_CBD + DIS_TO_ELDERCARE + DIS_TO_HAWKER + DIS_TO_MRT
    + DIS_TO_POPPRIM + DIS_TO_MALL +
    DIS_TO_COMMUNITYCLUB + DIS_TO_HOSPITAL + DIS_TO_PARK +
    N_CHILDCARE_350M + N_CLINIC_350M + N_BUS_350M + N_PRI_1KM + N_KINDY_350M + DIS_TO_SPORTS +DIS_TO_SUPERMARKET,
  data = HDBresale)

# End time and duration
end_time <- Sys.time()
duration <- end_time - start_time

cat("Model completed at:", format(end_time, "%Y-%m-%d %H:%M:%S"), "\n")
cat("Processing time:", round(duration, 2), attr(duration, "units"), "\n")

```

The code below outputs a statistical summary table for the log-transformed MLR model, similar to the base MLR model. For clearer interpretation, the intercept is hidden and included as a footnote. All p-values are below 0.05, allowing us to reject the null hypotheses and conclude that both the model and all coefficients are statistically significant at the 95% confidence level.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tbl_regression(loghdb_mlr, intercept = FALSE) %>%
  bold_labels() %>%
  bold_p(t = 0.05) %>%
  add_glance_source_note(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, adj.r.squared, AIC, statistic, p.value, sigma)) %>%
  modify_caption("Table: Log_MLR Model Statistical Summary") %>%
  modify_table_body(~ .x %>%
      mutate(across(c(estimate, conf.low, conf.high), ~(exp(.)-1) * 100)) %>% # change to percentage
      arrange(desc(abs(estimate)))) %>%  #sort
  modify_header(
    estimate = "**% Change in Price**",
    conf.low = "**CI Lower**", 
    conf.high = "**CI Upper**") %>%
  modify_fmt_fun(
    estimate = function(x) style_number(x, digits = 3),
    conf.low = function(x) style_number(x, digits = 3),
    conf.high = function(x) style_number(x, digits = 3))%>%
  modify_footnote(
    estimate ~ "Intercept: Estimate :1.330e+01, Std. Error: 1.546e-02, p value < 2e-16")%>%
   modify_caption("Table: Log-MLR Model Statistical Summary")
```

The log-MLR model achieves an R^2^ of 0.836 and an Adjusted R^2^ of 0.835, indicating that approximately 83.6% of the variation in 4-room HDB resale prices is explained by the selected explanatory variables. This reflects a stronger model fit. Consistent with the base MLR results, structural factors remain the most influential drivers of resale value in the log-linear specification.

In the log-MLR models, coefficients are interpreted in percentage terms. For example, the coefficient for HIP indicates that, holding other variables constant, flats located in blocks that have undergone the Home Improvement Programme are associated with an approximate 1.42% increase in resale price on average. Similarly, each additional floor level increases price by around 0.71%, while each additional year of age reduces price by about 0.89% per year.

```{r}
#| eval: false
#| echo: false
write_rds(loghdb_mlr, "data/model/loghdb_mlr.rds")
```

```{r}
#| echo: false
loghdb_mlr <-read_rds("data/model/loghdb_mlr.rds")
```

#### Regression Diagnostics

Regression models rely on several key assumptions, including linear relationships between independent variable and the target variable, independence of errors, constant variance of residuals (homoscedasticity), normally distributed errors, and the absence of high multicollinearity among explanatory variables. In this section, we conduct diagnostic checks to verify that the log-MLR model satisfies these assumptions. The **olsrr** and **performance** packages are used to implement these diagnostics, as they provide a comprehensive suite of tools for assessing regression model validity, including heteroskedasticity tests, collinearity diagnostics, model fit evaluations, influence measures, and residual analysis. These checks ensure that the fitted model is statistically sound and appropriate for inference.

Since diagnostic requirements are the same for MLR models and the log transformation should enhances normality and variance stability, regression assumptions are evaluated only for the log-MLR model.

::: panel-tabset
##### Multicollinearity Test

Multicollinearity arises when independent variables are highly correlated with one another. This condition inflates the variance of coefficient estimates, making them unstable and less reliable. As a result, variables that are genuinely important may appear statistically insignificant due to shared explanatory power.

To assess multicollinearity in our log-MLR model, we use the [`check_collinearity()`](https://easystats.github.io/performance/reference/check_collinearity.html) function from the **performance** package, which reports Variance Inflation Factors (VIFs) and related diagnostics.

```{r}
logmlr.vif <- check_collinearity(loghdb_mlr)
plot(logmlr.vif) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Log-MLR Model VIF")
```

+---------------+------------------------------------------------------+
| VIF value     | Interpretation                                       |
+===============+======================================================+
| 1             | No correlation between explanatory variables (ideal) |
+---------------+------------------------------------------------------+
| 1-5           | Moderate correlation; acceptable in most models.     |
+---------------+------------------------------------------------------+
| \>5           | Potential multicollinearity concern                  |
+---------------+------------------------------------------------------+
| \>10          | Serious multicollinearity problems                   |
+---------------+------------------------------------------------------+

All VIF values are below 10, indicating no signs of multicollinearity in the explanatory variables.

##### Test for Non-Linearity

The code below generates diagnostic plots to examine whether the residuals exhibit any non-linear patterns that may violate the linearity assumption.

The residual pattern largely centres around zero, consistent with the linearity assumption. A slight curve in the reference line suggests that some explanatory variables may have non-linear effects. This observation provides justification for considering complementary non-linear models. While small amount of outliers are visible, they do not exert substantial influence on the overall pattern.

```{r}
check_model(loghdb_mlr,   #performance pacakge
            check = "linearity")
```

##### Test for Normality Assumption

If the residuals follow a normal distribution, the normality assumption holds, ensuring valid hypothesis tests and confidence intervals.

The code chunk below uses `check_normality()` from the **performance** package to examine the normality of residuals via a Q–Q plot. The results show that most points lie closely along the reference line, indicating that the residuals are approximately normally distributed. Although minor deviations are observed at the tails, this is common in large datasets and does not materially affect the validity of the model.

```{r}
plot(check_normality(loghdb_mlr), type ="qq")
```

The code chunk below uses [`check_model()`](https://easystats.github.io/performance/reference/check_model.html) from performance package to test residuals normality.

```{r}
check_model(loghdb_mlr,
            check= "normality")
```

The figure reveals that the residuals of the log-linear model resemble normal distribution.

##### Testing for Spatial Autocorrelation

Since the hedonic model relies on spatially referenced data, it is important to test for spatial autocorrelation in the regression residuals. Standard OLS assumes that observations are independent. However, in spatial datasets, nearby locations often exhibit similar or dissimilar values, consistent with Tobler’s First Law of Geography. If spatial autocorrelation is present, coefficient estimates and p-values may be biased, leading to model misspecification. To address this, we apply Moran’s I test on the residuals. A significant Moran’s I would suggest that the model has not fully accounted for spatial dependence and that a spatial modelling approach (e.g., GWR) may be required.

```{r}
ols_test_correlation(loghdb_mlr)
```

The OLS autocorrelation value of 0.998 suggests strong positive autocorrelation that the residuals are not independent. This violates the OLS assumption of error independence and should be further tested using Moran’s I for spatial dependence. Significant spatial autocorrelation in the residuals means the model is not capturing the full spatial story, some spatial dependence is left unexplained.

The following code first saves the model `residuals` as a data frame and then joins them back to the `HDBresale` spatial object. This allows each record to retain both its geographic location and residual value. Then we visualise the spatial distribution of residuals using the **tmap** package. This enables us to assess whether residuals exhibit any spatial clustering or systematic geographic patterns that the model has not captured.

```{r}
#| code-fold: true
#| code-summary: "CODE"
logmlr.output <- as.data.frame(loghdb_mlr$residuals)
HDBresale <- cbind(HDBresale, logmlr.output) %>%
  rename(`LOGMLR_RES`= `loghdb_mlr.residuals`)
```

```{r}
tm_shape(mpsz) +
  tm_polygons(fill_alpha =0.4) +
  tm_shape(HDBresale) +
  tm_dots( fill = "LOGMLR_RES",
           size = 0.4,
           col = "black",
           fill.scale = tm_scale(
             n= 10,
             value = rev(brewer.pal(11, "RdBu")),
             style = "quantile",
             midpoint = NA),
           fill.legend = tm_legend(title = "Residuals")) +
  
  tm_title("Log MLR Residuals (Quantile Classification)") +
  tm_layout(legdend.outside = TRUE)
```

Each dot represents a condo resale residual (actual - predicted):

-   Green dots (positive): model underestimated (actual \> predicted).

-   Purple dots (negative): model overestimated (actual \< predicted).

If residuals were spatially random, green and purple dots would be evenly scattered across Singapore. However, this is not observed. Further, the coexistence of positive and negative residuals within the same region indicates that the OLS model cannot fully capture local variations in housing market dynamics.

To formally verify this spatial dependence, we proceed with Moran’s I test on the model residuals.

First, we compute distance-based weight matrix using [`st_dist_band()`](https://sfdep.josiahparry.com/reference/st_dist_band) function of **sfdep.** A 1.5 km distance band is selected as it reflects a reasonable neighbourhood interaction range in Singapore. Most daily amenities are designed to be within short walking or public-transport access, and extending the radius beyond 1.5 km risks diluting local price relationships by mixing flats from different neighbourhood markets.

```{r}
HDBresale <- HDBresale %>%
  mutate(nb = st_dist_band(st_geometry(.), upper = 1500),
         wt = st_weights(nb, style = "W"))
```

Then, we use [`global_moran_perm()`](https://sfdep.josiahparry.com/reference/global_moran_perm) of **sfdep** package conduct Moran's I test for residual spatial autocorrelation with Monte Carlo Simulations (500 times).

```{r}
#| code-fold: true
#| code-summary: "CODE"
#| eval: false
start_time <- Sys.time()
cat("Starting Global Moran's I Simulation at:", format(Sys.time(), "%H:%M:%S"), "\n")
set.seed(1234)
gmperm <-global_moran_perm(HDBresale$LOGMLR_RES,
                  nb = HDBresale$nb,
                  wt = HDBresale$wt,
                  alternative = "two.sided",
                  nsim = 499)
# End time and duration
end_time <- Sys.time()
duration <- end_time - start_time

cat("Simulation completed at:", format(end_time, "%Y-%m-%d %H:%M:%S"), "\n")
cat("Processing time:", round(duration, 2), attr(duration, "units"), "\n")
```

```{r}
#| eval: false
#| echo: false
write_rds(gmperm, "data/rds/gmperm.rds")
```

```{r}
#| echo: false
gmperm <- read_rds("data/rds/gmperm.rds")
```

```{r}
gmperm
```

The Global Moran’s I value of 0.3389 (p \< 0.05) indicates statistically significant positive spatial autocorrelation in the residuals. This suggests the presence of mild spatial clustering across the study area at the confidence level of 95%. That nearby locations tend to exhibit similar residual patterns.
:::

### Geographically Weighted Regression

Since the OLS model cannot fully account for the spatial dependence in HDB resale prices, we next apply a Geographically Weighted Regression (GWR) model to examine how the effects of explanatory variables vary across Singapore. Unlike the global OLS model, GWR allows for localised coefficient estimation, providing deeper insights into the spatial heterogeneity of housing market dynamics.

In this section,we build hedonic pricing models using both the fixed and adaptive bandwidth schemes. The `bw.gwr()` function in the **GWModel** package selects the optimal bandwidth for GWR. With `adaptive = FALSE`, it searches for a fixed distance bandwidth, while with `adaptive = TRUE`, it finds an adaptive (k-nearest-neighbour) bandwidth.

The formula for GWR is as follows:

$$\log(price_i) = \beta_0(u_i,v_i) + \beta_1(u_i,v_i)X_{1i} + \beta_2(u_i,v_i)X_{2i} + \cdots + \beta_k(u_i,v_i)X_{ki} + \varepsilon_i$$ $\log(price_i)$ = Natural log of resale price for observation $i$

Where:

-   $\beta_0(u_i,v_i)$ = Local intercept at coordinates $(u_i, v_i)$

-   $\beta_j(u_i,v_i)$ = Local coefficient for predictor $j$ at location $(u_i, v_i)$

-   $X_{ji}$ = Value of predictor $j$ for observation $i$

-   $\varepsilon_i$ = Error term for observation $i$

-   $(u_i, v_i)$ = Spatial coordinates (longitude, latitude)

To ensure consistency and comparability with the earlier OLS results, the log-transformed resale price is retained as the dependent variable, along with the same set of explanatory variables.

After estimating the GWR model, we extract diagnostic and model evaluation measures stored in the `sf` DataFrame under the `SDF` output list. These include observed and predicted values, residuals, condition numbers, local R^2^, coefficient estimates, standard errors, and corresponding t-values.

The local R^2^ represents the spatially varying explanatory power of the model (ranging from 0 to 1). Residuals, calculated as the differences between observed and predicted values, can be visualised to illustrate spatial patterns of model performance.

Finally, the standard errors of coefficients reflect the reliability of local parameter estimates. Smaller standard errors relative to the magnitude of coefficients suggest stable and well-estimated relationships, whereas larger standard errors point to possible local collinearity issues. By jointly examining the local R^2^, residual and coefficient estimates maps, we can identify spatial variations in model performance, providing a more nuanced understanding of the underlying factors influencing HDB resale prices across Singapore.

#### Fixed Bandwidth GWR Model

In this section, we build a GWR model using a fixed bandwidth, where a fixed distance is applied to define the neighbourhood used in each local regression. To determine the optimal fixed distance, we first evaluate the bandwidths through kernel and validation comparisons then select the combination that provides the best model fit. Two kernel types were tested: Gaussian and Bisquare. The Gaussian kernel generally produces smoother local estimates and performs better in handling sparse areas, whereas the Bisquare kernel provides more localised estimates and can delineate clearer spatial boundaries due to its sharp cut-off.

Under the cross-validation (CV) method, the bandwidth with the lower CV score is selected, while under the AICc method, the bandwidth with the lower AICc value is preferred. The selected kernel and its corresponding optimal bandwidth are then applied in the `gwr.basic()` function to calibrate the models.

::: panel-tabset
##### Selecting Optimal Bandwidth

###### CV Method

The following code performs a cross-validation-based bandwidth optimisation using the `bw.gwr()` function with `adaptive = FALSE`, specifying a fixed bandwidth. The optimisation process took approximately 22 minutes, and the results returned the following optimal fixed bandwidths:

-   **Gaussian kernel**: 679.2717 metres (CV score = 17.80391)

-   **Bisquare kernel**: 6,790.265 metres (CV score = 31.69816)

Since a lower CV score indicates a better model fit, the Gaussian kernel with a bandwidth of **679.27** metres was selected for the subsequent GWR modelling.

```{r}
#| code-fold: true
#| code-summary: "CODE"
#| eval: false
# Start timing
kernels_to_test <- c("gaussian", "bisquare")
results_fixed_cv <- list()

start_time <- Sys.time()
cat("=== GWR Fixed Bandwidth Selection - Kernel Comparison ===\n")

for(kernel_type in kernels_to_test) {
  cat("\n--- Testing", kernel_type, "kernel ---\n")
  
  # Run the bandwidth selection
  bw_result <- bw.gwr(
    formula = LOG_RESALE_PRICE ~ FLOOR + HIP + MUP + AGE + AREA_SQM +
      DIS_TO_CBD + DIS_TO_ELDERCARE + DIS_TO_HAWKER + DIS_TO_MRT +
      DIS_TO_POPPRIM + DIS_TO_MALL +
      DIS_TO_COMMUNITYCLUB + DIS_TO_HOSPITAL + DIS_TO_PARK +
      N_CHILDCARE_350M + N_CLINIC_350M + N_BUS_350M + N_PRI_1KM + 
      N_KINDY_350M + DIS_TO_SPORTS + DIS_TO_SUPERMARKET,
    data = HDBresale,
    approach = "CV",
    kernel = kernel_type,
    adaptive = FALSE,
    longlat = FALSE)
  
  results_fixed_cv[[kernel_type]] <- bw_result
  cat("-", kernel_type, "bandwidth:", round(bw_result, 4), "\n")}

# End timing and display results
end_time <- Sys.time()
duration <- end_time - start_time

cat("\n", rep("=", 50), "\n", sep="") 
cat("COMPLETED: GWR Fixed Bandwidth Selection\n")
cat("Processing time: ", round(duration, 2), attributes(duration)$units, "\n")
```

```{r}
#| eval: false
#| echo: false
write_rds(results_fixed_cv, "data/model/bwresults_fixed_cv.rds")
```

```{r}
#| echo: false
bwresults_fixed_cv <- read_rds("data/model/bwresults_fixed_cv.rds")
```

```{r}
bwresults_fixed_cv
```

###### AIC Method

The following code performs an AICc-based fixed-bandwidth search using the `bw.gwr()` function. The optimisation process took approximately 1.51 hours and returned the following results:

-   **Gaussian kernel**: optimal fixed bandwidth 418.6772 metres, AICc –30,246.29

-   **Bisquare kernel**: optimal fixed bandwidth6,790.236 metres, AICc –24,127.01

Since lower AICc values indicate a better model fit, the Gaussian kernel with a bandwidth of **418.6772** metres is selected as the optimal fixed bandwidth for the subsequent GWR modelling.

```{r}
#| code-fold: true
#| code-summary: "CODE"
#| eval: false
# Start timing
kernels_to_test <- c("gaussian", "bisquare")
bwresults_fixed_aic <- list()

start_time <- Sys.time()
cat("=== GWR Fixed Bandwidth Selection (AIC Apporach) - Kernel Comparison ===\n")

for(kernel_type in kernels_to_test) {
  cat("\n--- Testing", kernel_type, "kernel ---\n")
  
  # Run the bandwidth selection
  bw_result <- bw.gwr(
    formula = LOG_RESALE_PRICE ~ FLOOR + HIP + MUP + AGE + AREA_SQM +
      DIS_TO_CBD + DIS_TO_ELDERCARE + DIS_TO_HAWKER + DIS_TO_MRT +
      DIS_TO_POPPRIM + DIS_TO_MALL +
      DIS_TO_COMMUNITYCLUB + DIS_TO_HOSPITAL + DIS_TO_PARK +
      N_CHILDCARE_350M + N_CLINIC_350M + N_BUS_350M + N_PRI_1KM + 
      N_KINDY_350M + DIS_TO_SPORTS + DIS_TO_SUPERMARKET,
    data = HDBresale,
    approach = "AIC",
    kernel = kernel_type,
    adaptive = FALSE,
    longlat = FALSE)
  
  bwresults_fixed_aic[[kernel_type]] <- bw_result
  cat("-", kernel_type, "bandwidth:", round(bw_result, 4), "\n")}

# End timing and display results
end_time <- Sys.time()
duration <- end_time - start_time

cat("\n", rep("=", 50), "\n", sep="") 
cat("COMPLETED: GWR Fixed Bandwidth Selection\n")
cat("Processing time: ", round(duration, 2), attributes(duration)$units, "\n")
```

```{r}
#| eval: false
#| echo: false
write_rds(bwresults_fixed_aic, "data/model/bwresults_fixed_aic.rds")
```

```{r}
#| echo: false
bwresults_fixed_aic <- read_rds("data/model/bwresults_fixed_aic.rds")
```

```{r}
bwresults_fixed_aic
```
:::

##### Calibrating Fixed Bandwidth GWR Model

::: panel-tabset
###### CV Method

The code chunk below use Gaussian Kernel and fixed optimal bandwidth 679.2717 meteres to calibrate GWR model.

```{r}
#| code-fold: true
#| code-summary: "CODE"
#| eval: false
# Start timing
start_time <- Sys.time()
cat("Building Fixed Bandwidth GWRModel (CV)... This may take a while.\n")

gwr_fixed_cv <- gwr.basic(
 formula = LOG_RESALE_PRICE ~ FLOOR + HIP + MUP + AGE + AREA_SQM +
    DIS_TO_CBD + DIS_TO_ELDERCARE + DIS_TO_HAWKER + DIS_TO_MRT
    + DIS_TO_POPPRIM + DIS_TO_MALL +
    DIS_TO_COMMUNITYCLUB + DIS_TO_HOSPITAL + DIS_TO_PARK +
    N_CHILDCARE_350M + N_CLINIC_350M + N_BUS_350M + N_PRI_1KM + N_KINDY_350M + DIS_TO_SPORTS +DIS_TO_SUPERMARKET,
  data = HDBresale,
  bw= 679.2717,
  kernel = "gaussian",
  adaptive = FALSE,
  longlat = FALSE)

# End timing and display results
end_time <- Sys.time()
duration <- end_time - start_time

cat("COMPLETED: Fixed Bandwidth GWRModel\n")
cat("Processing time: ", round(duration, 2), attr(duration, "units"), "\n")
```

```{r}
#| eval: false
#| echo: false
write_rds(gwr_fixed_cv, "data/model/gwr_fixed_cv.rds")
```

```{r}
#| echo: false
gwr_fixed_cv <- read_rds("data/model/gwr_fixed_cv.rds")
```

The model output shows that the fixed-bandwidth GWR achieves an adjusted R^2^ of 0.9620 and an AICc of –29,674.81, representing 15% improvement over the global regression model, which recorded an adjusted R^2^ of 0.8354.

```{r}
gwr_fixed_cv
```

###### AIC Method

Using the optimal Gaussian fixed bandwidth of 418.6772 metres identified through the AICc-based search, we construct the GWR model. The results show an R^2^ of 0.9745 and an adjusted R^2^ of 0.9677, with an AICc of –30,246.89. Overall, the AICc-optimised bandwidth leads to a better-fitting and more efficiently specified GWR model.

```{r}
#| code-fold: true
#| code-summary: "CODE"
#| eval: false
# Start timing
start_time <- Sys.time()
cat("Building Fixed Bandwidth GWRModel (AIC, gaussian)... This may take a while.\n")

gwr_fixed_aic <- gwr.basic(
 formula = LOG_RESALE_PRICE ~ FLOOR + HIP + MUP + AGE + AREA_SQM +
    DIS_TO_CBD + DIS_TO_ELDERCARE + DIS_TO_HAWKER + DIS_TO_MRT
    + DIS_TO_POPPRIM + DIS_TO_MALL +
    DIS_TO_COMMUNITYCLUB + DIS_TO_HOSPITAL + DIS_TO_PARK +
    N_CHILDCARE_350M + N_CLINIC_350M + N_BUS_350M + N_PRI_1KM + N_KINDY_350M + DIS_TO_SPORTS +DIS_TO_SUPERMARKET,
  data = HDBresale,
  bw= 418.6772,
  kernel = "gaussian",
  adaptive = FALSE,
  longlat = FALSE)

# End timing and display results
end_time <- Sys.time()
duration <- end_time - start_time

cat("COMPLETED: Fixed Bandwidth GWRModel\n")
cat("Processing time: ", round(duration, 2), attr(duration, "units"), "\n")
```

```{r}
#| eval: false
#| echo: false
write_rds(gwr_fixed_aic, "data/model/gwr_fixed_aic.rds")
```

```{r}
#| echo: false
gwr_fixed_aic <- read_rds("data/model/gwr_fixed_aic.rds")
```

```{r}
gwr_fixed_aic
```
:::

#### Adaptive Bandwidth GWR Model

Next, we construct the GWR model using an adaptive bandwidth. The adaptive bandwidth automatically adjusts the spatial extent of each local regression, ensuring that areas with dense observations use smaller neighbourhoods, while regions with fewer observations draw from a wider spatial context to maintain model stability and reliability.

The optimisation uses either `AIC` or `CV`, controlled by the approach argument. In this study, we evaluate both the AICc and CV approaches and the Gaussian and Bisquare kernels to identify an adaptive bandwidth that best captures local spatial variation.

##### Selecting Optimal Bandwidth

::: panel-tabset
###### CV Method

The following code performs the adaptive bandwidth selection using the `approach = "CV"` option in the `bw.gwr()` function. The optimisation process returns an optimal adaptive bandwidth of **140**, meaning that each local regression in the GWR model is calibrated using the 140 nearest neighbouring observations. This conclusion is based on the lower CV score, where the Gaussian kernel produced a score of 20.12471 at a bandwidth of 140 neighbours, compared with the Bisquare kernel, which produced a higher CV score of 30.59457 at 2,116 neighbours. Since lower CV values indicate a better-fitting model, the Gaussian adaptive bandwidth of 140 was selected.

```{r}
#| code-fold: true
#| code-summary: "CODE"
#| eval: false
# Start timing
kernels_to_test <- c("gaussian", "bisquare")
bwresults_adpative_cv <- list()

start_time <- Sys.time()
cat("=== GWR Adaptive Bandwidth Selection - Kernel Comparison ===\n")

for(kernel_type in kernels_to_test) {
  cat("\n--- Testing", kernel_type, "kernel ---\n")
  
  # Run the bandwidth selection
  bw_result <- bw.gwr(
    formula = LOG_RESALE_PRICE ~ FLOOR + HIP + MUP + AGE + AREA_SQM +
      DIS_TO_CBD + DIS_TO_ELDERCARE + DIS_TO_HAWKER + DIS_TO_MRT +
      DIS_TO_POPPRIM + DIS_TO_MALL +
      DIS_TO_COMMUNITYCLUB + DIS_TO_HOSPITAL + DIS_TO_PARK +
      N_CHILDCARE_350M + N_CLINIC_350M + N_BUS_350M + N_PRI_1KM + 
      N_KINDY_350M + DIS_TO_SPORTS + DIS_TO_SUPERMARKET,
    data = HDBresale,
    approach = "CV",
    kernel = kernel_type,
    adaptive = TRUE,
    longlat = FALSE)
  
  bwresults_adpative_cv[[kernel_type]] <- bw_result
  cat("-", kernel_type, "bandwidth:", round(bw_result, 4), "\n")}

# End timing and display results
end_time <- Sys.time()
duration <- end_time - start_time

cat("\n", rep("=", 50), "\n", sep="") 
cat("COMPLETED: GWR Adaptive Bandwidth Selection\n")
cat("Processing time: ", round(duration, 2), attributes(duration)$units, "\n")
```

```{r}
#| eval: false
#| echo: false
write_rds(bwresults_adpative_cv, "data/model/bwresults_adpative_cv.rds")
```

```{r}
#| echo: false
bwresults_adpative_cv <- read_rds("data/model/bwresults_adpative_cv.rds")
```

```{r}
bwresults_adpative_cv
```

###### AIC Method

Similarly, the AICc-based optimisation returns the same optimal adaptive bandwidth. The Bisquare kernel produced an AICc of –24,445.65 at 2,116 neighbours, while the Gaussian kernel achieved a substantially lower AICc of –28,135.92 at 140 neighbours. Since lower AICc values indicate a better model fit, the Gaussian adaptive bandwidth of **140** neighbours is again selected as the optimal specification.

```{r}
#| code-fold: true
#| code-summary: "CODE"
#| eval: false
# Start timing
kernels_to_test <- c("gaussian", "bisquare")
bwresults_adaptive_aic <- list()

start_time <- Sys.time()
cat("=== GWR Adaptive Bandwidth Selection (AIC Apporach) - Kernel Comparison ===\n")

for(kernel_type in kernels_to_test) {
  cat("\n--- Testing", kernel_type, "kernel ---\n")
  
  # Run the bandwidth selection
  bw_result <- bw.gwr(
    formula = LOG_RESALE_PRICE ~ FLOOR + HIP + MUP + AGE + AREA_SQM +
      DIS_TO_CBD + DIS_TO_ELDERCARE + DIS_TO_HAWKER + DIS_TO_MRT +
      DIS_TO_POPPRIM + DIS_TO_MALL +
      DIS_TO_COMMUNITYCLUB + DIS_TO_HOSPITAL + DIS_TO_PARK +
      N_CHILDCARE_350M + N_CLINIC_350M + N_BUS_350M + N_PRI_1KM + 
      N_KINDY_350M + DIS_TO_SPORTS + DIS_TO_SUPERMARKET,
    data = HDBresale,
    approach = "AIC",
    kernel = kernel_type,
    adaptive = TRUE,
    longlat = FALSE)
  
  bwresults_adaptive_aic[[kernel_type]] <- bw_result
  cat("-", kernel_type, "bandwidth:", round(bw_result, 4), "\n")}

# End timing and display results
end_time <- Sys.time()
duration <- end_time - start_time

cat("\n", rep("=", 50), "\n", sep="") 
cat("COMPLETED: GWR Adaptive Bandwidth Selection\n")
cat("Processing time: ", round(duration, 2), attributes(duration)$units, "\n")
```

```{r}
#| eval: false
#| echo: false
write_rds(bwresults_adaptive_aic, "data/model/bwresults_adaptive_aic.rds")
```

```{r}
#| echo: false
bwresults_adaptive_aic <- read_rds("data/model/bwresults_adaptive_aic.rds")
```

```{r}
bwresults_adaptive_aic
```
:::

#### Calibrating Adaptive Bandwidth GWR Model

In this section, we construct the Geographically Weighted Regression (GWR) model using an adaptive bandwidth of 140 nearest neighbours and a Gaussian kernel, as this specification was identified as the optimal choice under both the AICc and CV bandwidth selection approaches.

```{r}
#| code-fold: true
#| code-summary: "CODE"
#| eval: false
# Start timing
start_time <- Sys.time()
cat("Building Adaptive Bandwidth GWRModel... This may take a while.\n")

gwr_adaptive <- gwr.basic(
 formula = LOG_RESALE_PRICE ~ FLOOR + HIP + MUP + AGE + AREA_SQM +
    DIS_TO_CBD + DIS_TO_ELDERCARE + DIS_TO_HAWKER + DIS_TO_MRT
    + DIS_TO_POPPRIM + DIS_TO_MALL +
    DIS_TO_COMMUNITYCLUB + DIS_TO_HOSPITAL + DIS_TO_PARK +
    N_CHILDCARE_350M + N_CLINIC_350M + N_BUS_350M + N_PRI_1KM + N_KINDY_350M + DIS_TO_SPORTS +DIS_TO_SUPERMARKET,
  data = HDBresale,
  bw=bw_adaptive_cv,
  kernel = "gaussian",
  adaptive = TRUE,
  longlat = FALSE)

# End timing and display results
end_time <- Sys.time()
duration <- end_time - start_time

cat("COMPLETED: Adaptive Bandwidth GWRModel\n")
cat("Processing time: ", round(duration, 2), attr(duration, "units"), "\n")
```

```{r}
#| eval: false
#| echo: false
write_rds(gwr_adaptive, "data/model/gwr_adaptive.rds")
```

```{r}
#| echo: false
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
```

```{r}
gwr_adaptive
```

#### GWR statistics

The GWR model summary indicates a substantial improvement in model performance compared with the global log-MLR. The global model achieved an Adjusted R^2^ of 0.8354 and an AICc of −17,707.73, whereas the best GWR model produced an Adjusted R^2^ of 0.96770 and an AICc of -30,246.89. This demonstrates that the GWR explains a much larger proportion of variation in resale prices and provides a better overall model fit. With R^2^ of 0.97452, the model explains approximately 97.45% of the variation in HDB resale prices.

The following code creates an interactive summary table using [`reactable()`](https://glin.github.io/reactable/articles/examples.html#basic-usage) from **reactable** package.

```{r}

model_results <- data.frame(
  Model = c("Global", "GWR (fixed bw=679)", "GWR (fixed bw=418)", "GWR (adaptive=140)"),
  AICc = c(-17707.86, -29674.81, -30246.89, -28135.92),
  R_squared = c(0.8358, 0.9666535, 0.9745183, 0.9564602),
  Adjusted_R_squared = c(0.8354, 0.9620098, 0.9677013, 0.9528076))

reactable(model_results,
          columns = list(
            Model = colDef(name = "Model"),
            AICc = colDef(name = "AICc", format = colFormat(digits = 2)),
            R_squared = colDef(name = "R_2", format = colFormat(digits = 4)),
            Adjusted_R_squared = colDef(name = "Adj. R_2", format = colFormat(digits = 4))),
            bordered = TRUE, highlight = TRUE, striped = TRUE)
```

#### Residual Diagnostics

The code below generates two residual choropleth maps using contiguous classification, plotted on the same scale to enable direct comparison between the log-MLR and GWR models. The results show that the GWR model produces much lighter colours overall, indicating smaller residuals and a better fit. Extreme values are markedly reduced, and spatial clustering of residuals is less pronounced compared to the MLR model. These improvements demonstrate that GWR more effectively captures local spatial heterogeneity in HDB resale prices.

```{r}
# Map residuals
#| code-fold: true
#| code-summary: "CODE"
gwrres<- tm_shape(mpsz) +
  tm_polygons(alpha = 0.3) +
  tm_shape(gwr_fixed_aic$SDF) +
  tm_dots(fill = "residual",
          size = 0.25,
          fill.scale = tm_scale_continuous()) +  
  tm_title("GWR Residuals (Continuous)") +
  tm_layout(legdend.outside = TRUE)
gwrres
```

```{r}
#| code-fold: true
#| code-summary: "CODE"
mlrres <- tm_shape(mpsz) +
  tm_polygons(alpha = 0.3) +
  tm_shape(HDBresale) +  # Changed from HDB_resale to HDBresale
  tm_dots(fill = "LOGMLR_RES",
          size = 0.25,
          fill.scale = tm_scale_continuous()) +  
  tm_title("Log MLR Residuals (Continuous)") +
  tm_layout(legend.outside = TRUE)  # Fixed typo: legdend -> legend

mlrres
```

A global Moran’s I test is applied to the residuals using `global_moran_perm()` from **sfdep** package.

```{r}
gwr_residuals <- gwr_fixed_aic$SDF$residual

gmperm_gwr <- global_moran_perm(
  gwr_residuals,                   
  nb = HDBresale$nb,
  wt = HDBresale$wt,
  alternative = "two.sided",
  nsim = 499)
```

```{r}
#| eval: false
#| echo: false
write_rds(gmperm_gwr, "data/rds/gmperm_gwr_fixed_aic.ds")
```

```{r}
#| echo: false
gmperm_gwr <- read_rds("data/rds/gmperm_gwr_fixed_aic.ds")
```

```{r}
gmperm_gwr
```

::: callout-note
Although there remains statistically significant spatial autocorrelation at the 95% confidence level, the 94.8% reduction in Moran’s I (from 0.33889 to 0.01778) indicates that the GWR model effectively accounted for most of the spatial dependence present in the data.
:::

#### Local R^2^ Mapping

In this section, we first prepare the SDF output from the GWR model into a data frame and extract the local R^2^ values. With functions from the tmap package, these values are then visualised at the national level and further examined by region to compare model fit across different areas.

::: panel-tabset
##### National Level

The following code is used to extract the SDF from the gwrm object and convert it to an sf data frame in SVY21 / EPSG:3414 for subsequent analysis.

```{r}
HDBresale_fixed <- st_as_sf(gwr_fixed_aic$SDF) %>%
  st_transform(3414) %>%
  mutate(across(where(is.character), as.numeric))
```

```{r}
#| eval: false
#| echo: false
write_rds(HDBresale_fixed, "data/model/HDBresale_fixed.rds")
```

```{r}
#| echo: false
HDBresale_fixed <- read_rds("data/model/HDBresale_fixed.rds")
```

```{r}
glimpse(HDBresale_fixed)
```

Using **tmap** pacakge, we visualise the spatial distribution of local R^2^, which summarises the location-specific explanatory power of the hedonic GWR.

```{r}
#| code-fold: true
#| code-summary: "CODE"
localr2<-tm_shape(mpsz) +
  tm_polygons(fill = "gray50",
              border.col = "white", 
              alpha = 0.6) +
tm_shape(HDBresale_fixed) +    #event points
  tm_dots(fill = "Local_R2",
          fill_alpha = 0.6,
          size = 0.2,
          fill.scale = tm_scale_intervals(
            style = "quantile", n = 5, values = "-viridis")) +
  tm_view(set_zoom_limits = c(11,14)) +
  tm_compass(type = "8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("GWR Model Local R^2") +
  tm_layout(frame = FALSE)
localr2
```

Interpretation:

High local R^2^ clusters (shown in purple) are concentrated in central and north-western regions, such as Kallang, Bukit Merah, Queenstown, Clementi, Jurong West, and Woodlands, where the model explains price variation very well (R^2^ 0.968–0.998). These areas are typically mature estates with stable housing attributes and well-established market dynamics, that the selected structural and locational variables capture price differences effectively.

Conversely, lower local R^2^ values (yellow to green tones) are observed in new-town and developing areas, including Punggol and Sengkang in the northeast, and parts of Tampines and Pasir Ris in the east, where model fit weakens (R^2^ 0.701–0.902).These estates are undergoing rapid urban development, and their pricing may be influenced by emerging infrastructure, future growth expectations, and unmeasured locational factors, which the current model does not fully capture.

##### Central Region

In the Central Region, the local R^2^ values are consistently high, mostly around or above 0.968. This reflects the fact that the central estates are well-established and mature, with stable housing characteristics and well-defined amenity structures. Consequently, the explanatory variables included in the model effectively capture the spatial effects influencing resale prices in these areas.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz[mpsz$REGION_N %in% c("CENTRAL REGION"), ]) + 
  tm_polygons(fill = "gray50",
              border.col = "white", 
              alpha = 0.6) +
tm_shape(HDBresale_fixed) +    #event points
  tm_dots(fill = "Local_R2",
          fill_alpha = 0.6,
          size = 0.2,
          fill.scale = tm_scale_intervals(
            style = "quantile", n = 5, values = "-viridis")) +
  tm_view(set_zoom_limits = c(11,14)) +
  tm_compass(type = "8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("GWR Model Local R^2 - Central Region") +
  tm_layout(frame = FALSE)
```

##### East Region

When examining the eastern region, distinct spatial clusters of local R^2^ can be observed. Within Tampines, Tampines North exhibits the strongest model fit, with R^2^ values reaching approximately 0.968. Moving westward, the local R^2^ gradually decreases to around 0.951, and continues to decline eastward towards Tampines East, where values range between 0.701 and 0.902. This gradient pattern suggests that the explanatory variables capture resale price variations more effectively in Tampines North, while the eastern precincts may be influenced by unobserved locational or developmental factors not fully represented in the model.

A comparable pattern is evident in Pasir Ris, where the eastern side of Pasir Ris Drive 8 records lower local R^2^ values (0.701–0.902, yellow). In contrast, the western side demonstrates moderately higher R^2^ values (0.903-0.950, light green).

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz[mpsz$REGION_N %in% c("EAST REGION"), ]) + 
  tm_polygons(fill = "gray50",
              border.col = "white", 
              alpha = 0.6) +
tm_shape(HDBresale_fixed) +    #event points
  tm_dots(fill = "Local_R2",
          fill_alpha = 0.6,
          size = 0.2,
          fill.scale = tm_scale_intervals(
            style = "quantile", n = 5, values = "-viridis")) +
  tm_view(set_zoom_limits = c(11,14)) +
  tm_compass(type = "8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("GWR Model Local R^2- East Region") +
  tm_layout(frame = FALSE)
```

##### North- East Region

In the North-East region, a distinct green belt can be observed across Hougang, Sengkang, and Punggol, where local R^2^ values range between 0.903 and 0.950. Surrounding these towns, areas shaded in yellow (R^2^ 0.701 to 0.902) are particularly along the outer fringes and Buangkok. This pattern suggests that while the model explains resale price variations reasonably well within the central parts of these towns, unexplained factors may influence pricing dynamics in the surrounding areas. Such factors could include ongoing urban development, varying accessibility, or unmeasured neighbourhood amenities that are not fully captured by the current model specification.

```{r, fig.height=4, fig.width=10}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz[mpsz$REGION_N %in% c("NORTH-EAST REGION"), ]) + 
  tm_polygons(fill = "gray50",
              border.col = "white", 
              alpha = 0.6) +
tm_shape(HDBresale_fixed) +    #event points
  tm_dots(fill = "Local_R2",
          fill_alpha = 0.6,
          size = 0.2,
          fill.scale = tm_scale_intervals(
            style = "quantile", n = 5, values = "-viridis")) +
  tm_view(set_zoom_limits = c(11,14)) +
  tm_compass(type = "8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("GWR Model Local R^2- North West Region") +
  tm_layout(frame = FALSE)
```

##### North Region

In the Northern region, the local R^2^ values are generally high, exceeding 0.968 across most parts of Woodlands and Sembawang, and ranging between 0.903 and 0.982 in Yishun. A notable yellow patch (R^2^ 0.701–0.902) appears in the Canberra, western corner of Sembawang, where the model performs less effectively. This weaker fit may be attributed to the area’s ongoing development as a new town, limited connectivity to established amenities, and proximity to large green and park spaces, which introduce spatial variation not fully captured by the current set of explanatory variables.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz[mpsz$REGION_N %in% c("NORTH REGION"), ]) + 
  tm_polygons(fill = "gray50",
              border.col = "white", 
              alpha = 0.6) +
tm_shape(HDBresale_fixed) +    #event points
  tm_dots(fill = "Local_R2",
          fill_alpha = 0.6,
          size = 0.2,
          fill.scale = tm_scale_intervals(
            style = "quantile", n = 5, values = "-viridis")) +
  tm_view(set_zoom_limits = c(11,14)) +
  tm_compass(type = "8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("GWR Model Local R^2- North Region") +
  tm_layout(frame = FALSE)
```

##### West Region

In the Western region, the local R^2^ values are generally high, exceeding 0.932 across the central areas of Jurong West, Clementi, and Choa Chu Kang, indicating strong model performance. In contrast, lower R^2^ values are observed along the edges near Bukit Batok, particularly around the Brickworks area, which borders the new-town Tengah. These peripheral zones exhibit greater variation in housing characteristics and ongoing development activities. Overall, outside the core heartland areas, local R^2^ values typically range between 0.903 and 0.950.

```{r}
#| code-fold: true
#| code-summary: "CODE"
tm_shape(mpsz[mpsz$REGION_N %in% c("WEST REGION"), ]) + 
  tm_polygons(fill = "gray50",
              border.col = "white", 
              alpha = 0.6) +
tm_shape(HDBresale_fixed) +    #event points
  tm_dots(fill = "Local_R2",
          fill_alpha = 0.6,
          size = 0.2,
          fill.scale = tm_scale_intervals(
            style = "quantile", n = 5, values = "-viridis")) +
  tm_view(set_zoom_limits = c(11,14)) +
  tm_compass(type = "8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg",
            position = c("right", "bottom")) +
  tm_title("GWR Model Local R^2- West Region") +
  tm_layout(frame = FALSE)
```
:::

### Visualising Coefficient Estimates

In explanatory modelling it is important to visualise and understand which factors matter most in shaping the model results. In our case we focus on the variables that contribute most to the percentage change in resale price as reflected by their coefficient estimates.

In this section we first summarise each variable’s median coefficient estimate, its local statistical significance represented by the t value, its spatial coverage, and its standard error. We then select the fifteen coefficients based on their magnitude and contextual relevance, particularly those related to everyday amenities such as places to shop, eat,live, study, heal, and spend leisure time to highlight the factors that most strongly influence housing prices across Singapore.

#### Selection of Coefficients for Spatial Mapping

Since our model is a log-transformed GWR model, the coefficient estimates can be interpreted as the percentage change in resale price associated with a one-unit change in each explanatory variable. Because most coefficients are relatively small in magnitude, they are converted into percentage terms to represent the % change in resale price. For distance-based variables, such as proximity to amenities, coefficients are further scaled to reflect the change per 100 m, rather than per metre, to enhance interpretability.

In addition to the coefficient magnitudes, the analysis also focuses on their t-values (TVs). The t-statistic is used in statistical hypothesis testing to determine if a relationship is statistically significant. In our GMR model, a coefficient is considered locally significant when its absolute t-value exceeds 1.96, corresponding to a 95% confidence level. We also calculate the frequency of local significance across all observations to evaluate how consistently each statistically significant variable influences resale prices spatially. For example, if the variable `FLOOR` is locally significant in all 8,679 observations, it achieves a 100% significance frequency, ranking at the top of the summary table. This measure highlights variables that exert consistent spatial effects on HDB resale prices across Singapore.

Finally, we examine the standard errors (SEs) of the coefficients to assess estimate reliability. Coefficients with small SEs are regarded as highly reliable, while those with moderate or large SEs indicate greater local uncertainty in parameter estimation.

To evaluate the spatial distribution of variable significance, a complete summary table is created sorting variables by the frequency of local significance across all 8,679 observations.

```{r}
#| code-fold: true
#| code-summary: "CODE"
# 1. Calculate spatial significance summary
sig_summary <- HDBresale_fixed %>%
  st_drop_geometry() %>%
  summarise(across(ends_with("_TV"), 
                   ~ sum(abs(.) > 1.96, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Sig_Locations") %>%
  mutate(Percentage = round(Sig_Locations / nrow(HDBresale_fixed) * 100, 1),
         Variable = gsub("_TV", "", Variable)) %>%
  arrange(desc(Percentage))

# 2. Extract median coefficients and create percentage effects
median_coefficients <- sapply(HDBresale_fixed, function(x) {
  if(is.numeric(x)) median(x, na.rm = TRUE) else NA})

# Create a data frame with the coefficients we need - INCLUDING INTERCEPT
coef_vars <- c("Intercept", "FLOOR", "HIP", "MUP", "AGE", "AREA_SQM", "DIS_TO_CBD", 
               "DIS_TO_ELDERCARE", "DIS_TO_HAWKER", "DIS_TO_MRT", "DIS_TO_POPPRIM",
               "DIS_TO_MALL", "DIS_TO_COMMUNITYCLUB", "DIS_TO_HOSPITAL", "DIS_TO_PARK",
               "N_CHILDCARE_350M", "N_CLINIC_350M", "N_BUS_350M", "N_PRI_1KM", 
               "N_KINDY_350M", "DIS_TO_SPORTS", "DIS_TO_SUPERMARKET")

gwr_percent_effects <- data.frame(
  Variable = coef_vars,
  median_Coefficient = median_coefficients[coef_vars])

# Convert to percentage effects - special handling for intercept
gwr_percent_effects$Percent_Effect <- sapply(1:nrow(gwr_percent_effects), function(i) {
  coef <- gwr_percent_effects$median_Coefficient[i]
  var_name <- gwr_percent_effects$Variable[i]
  
  if(var_name == "Intercept") {
    # Intercept: convert to actual price (SGD)
    exp(coef)  # This gives the baseline price in SGD
  } else if(grepl("^DIS_TO_", var_name)) {
    # Distance variables: effect per 100 meters
    (exp(coef * 100) - 1) * 100
  } else {
    # Other variables: effect per unit
    (exp(coef) - 1) * 100}})

# Format for display - special formatting for intercept
gwr_percent_effects$Formatted_Effect <- sapply(1:nrow(gwr_percent_effects), function(i) {
  if(gwr_percent_effects$Variable[i] == "Intercept") {
    sprintf("$%.0f", gwr_percent_effects$Percent_Effect[i])  # Dollar format for intercept
  } else {
    sprintf("%.3f%%", gwr_percent_effects$Percent_Effect[i])}})# Percentage for others

gwr_percent_effects$Direction <- ifelse(gwr_percent_effects$Percent_Effect > 0, "Positive", "Negative")
gwr_percent_effects$Units <- ifelse(gwr_percent_effects$Variable == "Intercept", "Baseline Price",ifelse(grepl("^DIS_TO_", gwr_percent_effects$Variable), "per 100m", "per unit"))

# 3. Combine both tables and include SE
combined_table <- sig_summary %>%
  rename(Variable_Sig = Variable) %>%
  # Join with percentage effects table
  left_join(gwr_percent_effects, by = c("Variable_Sig" = "Variable")) %>%
  # Get median SE for each variable
  mutate(SE_Column = paste0(Variable_Sig, "_SE")) %>%
  rowwise() %>%
  mutate(median_SE = ifelse(SE_Column %in% names(median_coefficients),
                      median_coefficients[SE_Column], NA)) %>%
  ungroup() %>%
  select(Variable = Variable_Sig, Formatted_Effect, Direction, Units,
         Sig_Locations, Percentage, median_SE) %>%
  # Format SE nicely
  mutate(Formatted_SE = ifelse(Variable == "Intercept", 
                         sprintf("%.3f", median_SE),
                         sprintf("%.2e", median_SE)))

# 4. Sort the combined table (intercept first, then by significance percentage)
intercept_row <- combined_table %>% filter(Variable == "Intercept")
non_intercept <- combined_table %>% filter(Variable != "Intercept") %>% arrange(desc(Percentage))
combined_table_sorted <- bind_rows(intercept_row, non_intercept)

# 5. Display the comprehensive combined table
combined_table_sorted %>%
  select(Variable, Formatted_Effect, Direction, Units, Percentage, Formatted_SE) %>%
  gt() %>%
  tab_header(
    title = md("**Comprehensive GWR Analysis of HDB Price Determinants (Median Effect)**"),
    subtitle = "Combining effect sizes, spatial significance, and estimate precision") %>%
  cols_label(
    Variable = "Explanatory Variable",
    Formatted_Effect = "Price Effect",
    Direction = "Direction",
    Units = "Units",
    Percentage = "Sig. %",
    Formatted_SE = "Standard Error") %>%
  tab_style(
    style = cell_text(weight = "bold", size = px(16)),
    locations = cells_title(groups = "title")) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(weight = "bold", color = "darkgreen"),
    locations = cells_body(rows = Variable == "Intercept")) %>%
  # Color coding for significance percentage
  data_color(
    columns = Percentage,
    colors = scales::col_numeric(
      palette = c("#FFE5E5", "#FFB3B3", "#FF0000"),
      domain = c(0, 100))) %>%
  fmt_number(
    columns = Percentage,
    decimals = 1) %>%
  tab_footnote(
    footnote = "Intercept: Not meaningfully interpretable as it represents impossible zero values for all variables. Included for reference only.",
    locations = cells_body(rows = Variable == "Intercept")) %>%
  tab_footnote(
    footnote = "Sig. %: Percentage of Singapore locations where relationship is statistically significant (p < 0.05)",
    locations = cells_column_labels(columns = Percentage)) %>%
  tab_options(
    table.width = pct(95),
    table.align = "center",
    heading.title.font.size = 20,
    heading.subtitle.font.size = 14)
```

The intercept represents the baseline price in dollar terms under the impossible scenario of all variables being zero. While this specific scenario is not meaningful, the spatial variation in intercept values reveals important geographic patterns in base housing costs across Singapore. All other coefficients represent the spatially varying relationships between each variable and HDB realse prices.

Based on the summary table, the following variables are identified as the most influential in explaining price variation: `FLOOR, AGE, AREA_SQM, HIP, DIS_TO_CBD, DIS_TO_MRT, DIS_TO_HAWKER, DIS_TO_MALL, DIS_TO_SUPERMARKET, DIS_TO_PARK, DIS_TO_ELDERCARE, N_BUS_350M, N_PRI_1KM, N_CHILDCARE, and N_CLINIC_350M`.

#### Geographical Patterns of Coefficients

Following the summary table, we implement two mapping functions that visualises local coefficient estimates and each coefficient’s t-value from the GWR. Observations are classified as:

-   Very significant positive: t ≥ 2.58 shown in dark purple.

-   Significant positive: 1.96 ≤ t \< 2.58 in light purple.

-   Significant negative: −2.58 \< t ≤ −1.96 in light orange.

-   Very significant negative: t ≤ −2.58 in dark orange.

-   Not significant: \|t\| \< 1.96 in light blue.

This framework follows the recommendations of Mennis (2006) and Matthews and Yang (2012) for mapping local coefficients. It simultaneously conveys both the direction of the relationship (price increasing versus price decreasing) and the strength of statistical evidence for each local effect. Conceptually, it is similar to the colour themed local Gi\* mapping used in Hot and Cold Spot Analysis, which visually differentiates areas of strong positive and negative influence across space.

The following code defines helper functions to create choropleth maps for coefficient estimates (effects on resale price) and for coefficient significance.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map <- function(coef_var, title) {

  HDBresale_fixed$transformed_coef <- HDBresale_fixed[[coef_var]]
  
  regions <- mpsz %>%
    group_by(REGION_N) %>%
    summarise(do_union = TRUE)
  
    # Many unique values - use intervals
    tm_shape(regions) +
      tm_borders(col = "gray50", lwd = 1.5) +
      tm_shape(HDBresale_fixed) +
      tm_dots(fill = "transformed_coef",
              fill.scale = tm_scale_intervals(
                #values = c("#8C2D04", "#D94801", "#F16913", "#6A51A3", "#453190", "#1F0C5A"), # orange to purple
                values = c("#8B0000", "#D73027", "#F46D43", "#74C476", "#31A354", "#006837"), # red to green
                #values = c("#CA0020", "#EF3B2C", "#F4A582", "#92C5DE", "#0571B0", "#053061"), # red to blue
                style = "fisher", n = 6),
              fill.legend = tm_legend(title = "Effect (raw)"),
              size = 0.3, 
              shape = 18) +
      tm_title(paste(title, "- Price Effect")) +
      tm_compass(type = "8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
      tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
      tm_grid(lwd = 0.1, alpha = 0.2) +
      tm_credits("Source: data.gov.sg",
                position = c("right", "bottom")) +
      tm_layout(na.show = FALSE)}
```

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_significance_map <- function(tv_var, title) {
  HDBresale_fixed$classification <- case_when(
    HDBresale_fixed[[tv_var]] >= 2.58 ~ "Very Sig Positive",      
    HDBresale_fixed[[tv_var]] >= 1.96 ~ "Sig Positive",          
    HDBresale_fixed[[tv_var]] <= -2.58 ~ "Very Sig Negative",     
    HDBresale_fixed[[tv_var]] <= -1.96 ~ "Sig Negative",          
    TRUE ~ "Not Significant")
  
  HDBresale_fixed$classification <- factor(
    HDBresale_fixed$classification,
    levels = c("Not Significant","Very Sig Negative", "Sig Negative", 
               "Sig Positive", "Very Sig Positive"))
  
  #Check the distribution
  #print(paste("Classification distribution for", title, ":"))
  #print(table(HDBresale_fixed$classification))
  
  regions <- mpsz %>%
    group_by(REGION_N) %>%
    summarise(do_union = TRUE)
  
  tm_shape(regions) +
    tm_borders(col = "gray80", lwd = 1.5) +
    tm_shape(HDBresale_fixed) +
    tm_dots(
      fill = "classification",
      size = 0.25, shape = 15,
      fill.scale = tm_scale_categorical(
        values = c("Very Sig Positive" = "#542788",
                   "Sig Positive" = "turquoise",
                   "Not Significant" = "grey50",
                   "Sig Negative" = "#f1a340",
                   "Very Sig Negative" = "#e66101")),
      fill.legend = tm_legend(title = title)) +
    tm_title(paste(title, "- Significance"))+
    tm_compass(type = "8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
    tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
    tm_grid(lwd = 0.1, alpha = 0.2) +
    tm_credits("Source: data.gov.sg",
              position = c("right", "bottom")) +
    tm_layout(na.show = FALSE)}
```


```{r}
#| code-fold: true
#| code-summary: "CODE"
create_se_map <- function(se_var, title) {
  # Extract the corresponding SE variable (e.g., "DIS_TO_MRT_SE")
  HDBresale_fixed$se_value <- HDBresale_fixed[[se_var]]
  
  regions <- mpsz %>%
    group_by(REGION_N) %>%
    summarise(do_union = TRUE)
  
  tm_shape(regions) +
    tm_borders(col = "gray50", lwd = 1.5) +
    tm_shape(HDBresale_fixed) +
    tm_dots(fill = "se_value",
            fill.scale = tm_scale_intervals(
              values = c("#F7FBFF", "#D2E3F3", "#9ECAE1", "#6BAED6", "#3182BD", "#08519C"), # blue gradient
              style = "fisher", n = 6),
            fill.legend = tm_legend(title = "Standard Error"),
            size = 0.3, 
            shape = 18) +
    tm_title(paste(title, "- Estimation Uncertainty")) +
    tm_compass(type = "8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
    tm_scalebar(bg.color = "white", bg.alpha = 0.5) +
    tm_grid(lwd = 0.1, alpha = 0.2) +
    tm_credits("Source: data.gov.sg",
              position = c("right", "bottom")) +
    tm_layout(na.show = FALSE)
}
```


##### Structoral Factors

We use the mapping functions to visualise structural factors, floor level, flat size, and building age, by local t-values and its coefficient estimates.

::: panel-tabset
###### Floor level

The results show that floor level is consistently strong positive across nearly all observations: higher floors contribute to a price premium throughout Singapore. Based on the median effect, for every one unit increase in floor level, the resale price increases by about 0.673 percent, holding other factors constant.



```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("FLOOR", "Floor Level")
create_significance_map ("FLOOR_TV", "Floor Level") 
create_se_map("FLOOR_SE", "Floor Level")
```

###### Floor Area

Similarly, larger flat sizes exhibit a significant positive relationship with resale prices across approximately 87 percent of the observations. Based on the median local effect, a one-sqm increase in floor area is associated with about a 0.567 percent rise in resale price, holding other factors constant. However, in the Taman Jurong area (shown in orange on the significance map), the effect of floor area turns significantly negative. Since the standard error in this area is not particularly low, the magnitude should be interpreted with caution.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("AREA_SQM", "Floor Area")
create_significance_map ("AREA_SQM_TV", "Floor Area")
create_se_map("AREA_SQM_SE", "Floor Area")
```

###### Block Age

In contrast, building age appears predominantly in dark orange, reflecting a significant negative relationship, older properties are generally associated with lower resale prices. Based on the median local effect, a one-year increase in building age is associated with about a 0.071 percent decrease in resale price, holding other factors constant. These negative effects are strong in mature estates such as Clementi, Bukit Merah, and Toa Payoh, whereas flats in Yew Tee at the northern end of Choa Chu Kang display a small positive relationship with resale price, although the magnitude of this effect remains modest.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("AGE", "Block Age")
create_significance_map ("AGE_TV", "Block Age")
create_se_map("AGE_SE", "Block Age")
```

###### HIP Completion

HIP completion has a spatially uneven and locally significant effect, influencing about 31.5 percent of all observations. The pattern suggests that HIP’s price impact depends strongly on estate characteristics and location. In mature estates such as Bukit Merah, Toa Payoh, and Clementi, the effect is often postive, where completed upgrading may signal improved living quality. Conversely, slightly negative or zero effects appear in northern and north-eastern towns like Woodlands, Choa Chu Kang, Jurong East, possibly reflecting overall newer building conditions and buyer preferences for newer flats. These results highlight that the HIP upgrades is highly context dependent.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("HIP", "HIP Completion")
create_significance_map ("HIP_TV", "HIP Completion")
create_se_map("HIP_SE", "HIP Completion")
```
:::

##### Locational Factors

We use the mapping functions to visualise locational factors.

In this section, choropleth maps are created to visualise the local coefficients for locational factors, categorised by statistical significance and magnitude of price effect.

::: panel-tabset
###### Proximity to MRTs

Proximity to MRTs generally represents negative effects, that shorter distances to MRT stations are associated with higher resale prices. Based on the median effect, for every 100 metres closer to an MRT station, the resale price increases by about 0.742 percent, holding other factors constant.This relationship is statistically significant for about 40.9% of all observations, suggesting that accessibility to MRT stations is an important driver of housing value, particularly in suburban estates. However, in highly accessible central locations such as Rochor, a slight positive relationship is observed, where flats located farther from MRT stations tend to command marginally higher prices.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("DIS_TO_MRT", "MRT proximity")
create_significance_map ("DIS_TO_MRT_TV", "MRT proximity")
create_se_map("DIS_TO_MRT_SE", "MRT proximity")
```

###### Proximity to CBD

Proximity to the CBD is statistically significant in approximately 35.1 percent of the dataset. At a glance, most coefficients are negative, indicating that as the distance from the CBD increases, resale prices tend to decline.However, areas such as Little India Rochor display a positive relationship with proximity to the CBD, similar to the pattern observed for MRT accessibility. Based on the median effect, every 100 metres closer to the CBD is associated with an estimated 0.463 percent increase in resale price,holding other factors constant.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("DIS_TO_CBD", "CBD proximity")
create_significance_map ("DIS_TO_CBD_TV", "CBD proximity")
create_se_map("DIS_TO_CBD_SE","CBD proximity") 
```

###### Proximity to Hawker Centres

Proximity to hawker centres is locally significant in about 35.0 percent of observations. The raw coefficient estimates range from approximately −0.02 to 0.0084, with a median effect of around 0.134 percent, holding other factors constant. The direction of significance is mixed, showing both positive and negative local effects. Negative relationships are more common in Bukit Batok, Choa Chu Kang, Bishan, and Redhill Bukit Merah, where hawker centres are scarcer and therefore more valued as essential community amenities that enhance flat values. In contrast, positive effects are more evident in areas such as Toa Payoh, Ting Bahru, Jurong West where the abundance of food options may dilute the unique value of hawker centres. In these dense neighbourhoods, proximity reduces housing desirability due to noise, hygiene concerns, or heavy foot traffic.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("DIS_TO_HAWKER", "Hawker Centre Proximity")
create_significance_map ("DIS_TO_HAWKER_TV", "Hawker Proximity")
create_se_map("DIS_TO_HAWKER_SE","Hawker Centre Proximity")  
```

###### Proximity to Malls

In terms of proximity to malls, which is locally significant in about 33.1 percent of observations, areas showing significance generally exhibit a negative relationship between distance and resale price. Based on the median converted effect, every 100 metres farther from a mall is associated with approximately a 0.874 percent decrease in resale price.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("DIS_TO_MALL", "Mall Proximity")
create_significance_map ("DIS_TO_MALL_TV", "Mall Proximity")
create_se_map("DIS_TO_MALL_SE", "Mall Proximity")  
```

###### Proximity to Eldercare Centres

Proximity to eldercare centres is locally significant in approximately 31.1 percent of observations across Singapore. The presence of eldercare facilities tends to add value to nearby HDB resale flats in areas such as Kallang, Hougang, and Punggol. Outside these locations, however, the variable shows little or no positive influence on resale prices. Based on the median effect, every 100 metres closer to an eldercare centre is associated with an estimated 0.322 percent decrease in resale price, holding other factors constant.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("DIS_TO_ELDERCARE", "Eldercare Proximity")
create_significance_map ("DIS_TO_ELDERCARE_TV", "Eldercare")
create_se_map("DIS_TO_ELDERCARE_SE", "Eldercare Proximity") 
```

###### Proximity to Supermarkets

Proximity to supermarkets is statistically significant in approximately 23.5 percent of observations. The results indicate that supermarket accessibility does not appear to significantly enhance property values in areas such as Bukit Merah, Queenstown, and Geylang, where the local coefficients are positive. However, since the standard errors in some of these areas are moderate, the results should be interpreted with caution.
In general, ease of access to supermarkets is regarded as a desirable amenity in most other regions. Based on the median effect, every 100 metres farther from a supermarket corresponds to an estimated 0.200 percent decrease in resale price,holding other factors constant.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("DIS_TO_SUPERMARKET", "Supermarket Proximity")
create_significance_map ("DIS_TO_SUPERMARKET_TV", "Supermarket")
create_se_map("DIS_TO_SUPERMARKET_SE","Supermarket Proximity") 
```

###### Proximity to Parks

Proximity to parks is locally significant in about 30.8 percent of observations, but its effects are not uniformly distributed across Singapore. In general, proximity to parks adds value to most properties located near green spaces. However, negative and significant effects are observed in Bukit Batok, Queenstown, and Kallang, where parks are already abundant.

```{r}
#| code-fold: true
#| code-summary: "CODE"

create_coefficient_map("DIS_TO_PARK", "Proximity to Park")
create_significance_map ("DIS_TO_PARK_TV", "Park Proximity")
create_se_map("DIS_TO_PARK_SE",    "Proximity to Park") 
```

###### Number of Primary Schools within 1km

The number of primary schools within 1 kilometre is statistically significant in about 19.1 percent of observations. In regions where this variable is significant, the availability of nearby schools is generally perceived as a desirable neighbourhood feature. Based on the median converted effect, the presence of one additional primary school within 1 kilometre is associated with approximately a 0.099 percent increase in resale price, holding other factors constant. However, in areas with an abundance of schools or where the local demographic composition includes fewer families with primary-school-aged children, this effect tends to diminish or become insignificant.

```{r}
#| code-fold: true
#| code-summary: "CODE"

create_coefficient_map("N_PRI_1KM", "No of Primary Schools within 1km")
create_significance_map ("N_PRI_1KM_TV", "N_Prisch_1KM") 
create_se_map("N_PRI_1KM_SE","Primary Schools within 1 km") 
```

###### Number of Bus Stops within 350m

In general, the number of bus stops within 350 metres is statistically significant in about 24.4 percent of observations. The effect tends to add value to neighbourhoods where MRT stations are not nearby, where bus routes offer greater network versatility. Based on the median converted effect, each additional bus stop within the defined buffer is associated with approximately a 0.114 percent increase in resale price, holding other factors constant.This finding suggests that bus accessibility is an important transport amenity for HDB residents. However, the magnitude of the effect should be interpreted with caution, as the standard errors are relatively high, indicating some degree of uncertainty in the local estimates.

```{r}
#| code-fold: true
#| code-summary: "CODE"

create_coefficient_map("N_BUS_350M", "No of Bus Stops within 350m")
create_significance_map ("N_BUS_350M_TV", "N_BUS_350M")
create_se_map("N_BUS_350M_SE","Bus Stops within 350 m")
```

###### Number to Childcare Centres within 350m

The number of childcare centres within 350 metres shows a marginal effect on resale prices, with local significance in about 18.0 percent of observations. The effect is generally positive, but it does not add value to flats in Yishun, Queenstown, Sengkang, and Hougang, likely because these  towns already have a high concentration of childcare facilities. After conversion, the median effect indicates that the presence of one additional childcare centre within 350 metres is associated with an estimated 0.025 percent increase in resale price.

```{r}
#| code-fold: true
#| code-summary: "CODE"
create_coefficient_map("N_CHILDCARE_350M", "No of Chilcare within 350m")
create_significance_map ("N_CHILDCARE_350M_TV", "# CC 350m")
create_se_map("N_CHILDCARE_350M_SE", "# Childcare within 350m")   
```

###### Number to Clinics within 350m

Although only 17.2 percent of observations show a statistically significant local relationship between the number of clinics within 350 metres and resale prices, the median converted effect is approximately 0.132 percent. Notably, in areas such as Bendemeer, Woodlands East, and Bukit Panjang, the local coefficients are negative and significant, likely reflecting a saturation effect. In these mature estates, where healthcare infrastructure is already abundant or hospitals are located nearby, the presence of additional clinics may weaken housing values. This finding is consistent with the pattern observed for proximity to hospitals, where the median effect is positive, indicating that being closer to a hospital tends to slightly depress resale prices, holding all other variables constant.

```{r}
#| code-fold: true
#| code-summary: "CODE"

create_coefficient_map("N_CLINIC_350M", "No of Clincs within 350m")
create_significance_map ("N_CLINIC_350M_TV", "# Clinics 350m")
create_se_map("N_CLINIC_350M_SE", "Clinics within 350 m") 
```
:::

### Explanatory Random Forest Model

In this section, we apply the **SpatialRF** method to develop a non-linear Random Forest model to identify feature importance and compare its performance with the log-GWR model described earlier. The following code uses the same set of explanatory variables and target variable to ensure consistency and comparability between the two modelling approaches.

```{r}
#| echo: false
HDBresale <- read_rds("data/rds/HDBresalelg.rds")
```

```{r}
#| code-fold: true
#| code-summary: "CODE"
predictors <- names(HDBresale)[c(4:7, 9:25)]
predictors
dependent <- names(HDBresale)[26]
dependent
```

#### Non-spatial Random Forest

We first randomly sample 1,500 observations from the dataset due to computational constraints (Benito (2025) recommends the sample size between 100 to 5000), and then define the distance matrix and distance threshold to create a non-spatial Random Forest model `rf()` from [**spatialRF**](https://cran.r-project.org/web/packages/spatialRF/index.html) pacakge.

```{r}
#| code-fold: true
#| code-summary: "CODE"
#| eval: false

set.seed(1234)

coords <- st_coordinates(HDBresale)
HDBresale_nogeom <- HDBresale %>%
  st_drop_geometry()
HDB_sample <- HDBresale %>%
  mutate(x = coords[,1], y = coords[,2]) %>%
  sample_n(1500)
coords_sample <- st_coordinates(HDB_sample)
distance.matrix <- as.matrix(dist(coords_sample))
distance.thresholds <- c(200,600,1200,3000)
xy <- HDB_sample[, c("x","y")] %>%
  st_drop_geometry()
HDB_sample <- HDB_sample %>% st_drop_geometry()
HDB_sample_non_spatial <- spatialRF::rf(
  data = HDB_sample,
  dependent.variable.name = dependent,
  predictor.variable.names = predictors,
  distance.matrix = distance.matrix,
  distance.thresholds = distance.thresholds,
  xy = xy, #not needed by rf, but other functions read it from the model
  seed = 1234,
  verbose = FALSE)
```

```{r}
#| eval: false
#| echo: false
write_rds(HDB_sample_non_spatial, "data/model/HDB_sample_non_spatial.rds")
```

```{r}
#| echo: false
HDB_sample_non_spatial <- read_rds("data/model/HDB_sample_non_spatial.rds")
```

The `summary()` function output for the `rf` object contains several key pieces of information that describe the model’s structure, performance, and feature importance metrics.

```{r}
summary(HDB_sample_non_spatial)
```

The following code uses `plot_residuals_diagnostic()` to produce the residual diagnostics. The residuals are approximately normally distributed, and the residuals versus fitted values plot shows no clear pattern. Finally, the multiscale Moran’s I results indicate that spatial autocorrelation remains present in the model, suggesting some neighbourhood effects are still not fully captured.

```{r, fig.height=8, fig.width=10}
spatialRF::plot_residuals_diagnostics(
  HDB_sample_non_spatial,
  verbose = FALSE)
```

By using the `print_performance()` function, the model demonstrates an R^2^ of 0.8943.

```{r}
spatialRF::print_performance(HDB_sample_non_spatial)
```

The following code outputs a table showing the global variable importance of the non-spatial Random Forest model, including several key diagnostic measures. These measures comprise the average minimum tree depth (where lower values indicate greater importance), the total number of node splits (higher values suggest more frequent use in decision-making), and the number of times a variable appears as a root node (higher frequency implies greater criticality). The accompanying p-values indicate the statistical significance of each variable’s contribution, with p \< 0.05 denoting significance at the 95% confidence level.

The results highlight several influential predictors, including distance to the CBD, building age, floor level (moderately important, though not statistically significant), proximity to hawker centres, floor area (not statistically significant), distance to MRT, and proximity to popular primary schools. These findings align with the GWR analysis.

```{r}
importance.df <- randomForestExplainer::measure_importance(
  HDB_sample_non_spatial,
  measures = c("mean_min_depth", "no_of_nodes", "times_a_root", "p_value"))
kableExtra::kbl(
  importance.df %>% 
    dplyr::arrange(mean_min_depth) %>% 
    dplyr::mutate(p_value = round(p_value, 4)),
  format = "html", caption = "Global Variable Importance for Non Spatial RF Model") %>%
  kableExtra::kable_paper("hover", full_width = F)
```

#### Spatial Random Forest

The following code builds a spatial model for HDB sample by using `rf_spatial()`.

```{r}
#| eval: false

cat("Starting Building Spatial Random Forest, this might take hours...\n")
cat("Processing time: ", round(duration, 2), attr(duration, "units"), "\n")

start_time <- Sys.time()
 HDB_sample_spatial <- rf_spatial(
  model = HDB_sample_non_spatial,
  xy = xy, #not needed by rf, but other functions read it from the model
  seed = 1234,
  verbose = FALSE,
  method ="mem.moran.sequential")

# End timing and display results
end_time <- Sys.time()
duration <- end_time - start_time


cat("COMPLETED: Spatial RF Model\n")
cat("Processing time: ", round(duration, 2), attr(duration, "units"), "\n")
```

```{r}
#| eval: false
#| echo: false
write_rds(HDB_sample_spatial, "data/model/HDB_sample_spatial.rds")
```

```{r}
#| echo: false
HDB_sample_spatial <- read_rds("data/model/HDB_sample_spatial.rds")
```

The Multiscale Moran’s I results indicate that a small degree of spatial autocorrelation (Moran’s I approximately 0.01) remains in the spatial Random Forest model. However, this represents a clear improvement compared to the non-spatial model, which recorded a higher Moran’s I value of approximately 0.05.

```{r}
spatialRF::plot_moran(
  HDB_sample_spatial, 
  verbose = FALSE)
```

```{r, fig.height=8, fig.width=15}
p1 <- spatialRF::plot_importance(
  HDB_sample_non_spatial, 
  verbose = FALSE) + 
  ggplot2::ggtitle("HDB Sample Non-spatial model") 

p2 <- spatialRF::plot_importance(
  HDB_sample_spatial,
  verbose = FALSE) + 
  ggplot2::ggtitle("HDB Sample Spatial model")

p1 | p2 
```

The `plot_importance()` function generates the feature importance plots for both the spatial and non-spatial models. The most influential predictors in both models are largely consistent with those identified in the log-GWR model, including variables such as proximity to the CBD, building age, Home Improvement Programme (HIP), floor level, proximity to MRT stations and hawker centres, and number of primary schools in 1km.

## Conclusion & Limitations

In this study, we utilise Multiple Linear Regression (MLR) and Geographically Weighted Regression (GWR) models to construct a hedonic pricing model for HDB 4-room resale flats from January 2025 to September 2025 with handful of carefully selected structural and locational factors. The explanatory variables cover a wide range of characteristics, including flat attributes (age, size, and floor level), government upgrading programmes, proximity to the CBD, accessibility to surrounding amenities and services, and counts of essential facilities such as schools, clinics, and bus stops.

The log-GWR model achieves an R^2^ of 0.97, indicating that it could explain 97 percent of the spatial variation in local resale prices. It outperforms the MLR model, effectively capturing local spatial heterogeneity, with the best model performance observed in central regions. We also visualise the spatial distribution of selected coefficient estimates to deepen the understanding of each variable’s local influence and statistical significance.

In addition, we experiment with the SpatialSF Random Forest method, which validats the variable importance rankings but provides limited explanatory insight compared to the GWR framework, given its black-box nature.


It is worth noting that, the hedonic model itself does not capture external market factors such as tax policies, housing regulations, or private property market dynamics, all of which may influence HDB resale prices. Furthermore, ongoing urban redevelopment and new town expansions may account for the lower local R^2^ values observed in several North-East and East regions. Finally, the model assumes a linear relationship between explanatory variables and prices. Future research could explore non-linear or hybrid spatial approaches to address this limitation. While machine learning models are primarily predictive, they can help uncover complex, non-linear structures in housing data. In contrast, spatial econometric models offer stronger explanatory power and can be used in tandem to enhance both prediction and interpretability in future studies.

Overall, the findings provide valuable insights into current HDB resale price determinants and may serve as a useful reference for policymakers, buyers, and urban planners.



## Reference

Kam, T. S. 13 Calibrating Hedonic Pricing Model for Private Highrise Property with GWR Method. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap13

Lim, A. (2025, February 24). 13+ housing & household statistics in Singapore (2025). SmartWealth Singapore. https://smartwealth.sg/housing-household-statistics-singapore/

Lim, J. (2025, October 24). HDB resale price growth in Q3 slows; private home prices driven by new launches. The Straits Times. Retrieved from https://www.straitstimes.com/singapore/hdb-resale-price-growth-in-q3-near-5-year-low-private-home-prices-driven-by-new-launches

Anselin, L., & Bera, A. K. (1998). Spatial dependence in linear regression models with an introduction to spatial econometrics. In A. Ullah & D. E. A. Giles (Eds.), Handbook of applied economic statistics (pp. 237–289). New York: Marcel Dekker

Brunsdon, C., Fotheringham, A. S., & Charlton, M. E. (1996). Geographically Weighted Regression: A Method for Exploring Spatial Nonstationarity. Geographical Analysis, 28(4), 281–298.

Mennis, J. (2006). Mapping the Results of Geographically Weighted Regression. The Cartographic Journal, 43(2), 171–179.

Matthews, S. A., & Yang, T.-C. (2012). Mapping the Results of Local Statistics: Using Geographically Weighted Regression. Demographic Research, 26, 151–166.

Xu, T. (2017). Hedonic Modelling of Singapore’s Resale Public Housing Market.

Chia, W.-M., Li, M., & Tang, H. (2017). Booms, Busts, and Household Mobility: The Role of Fundamentals and Policies in Singapore’s Housing Market. Journal of Housing Economics, 37, 16–33.

Wang, J., et al. (2022). Taxi Travel Time-Based Geographically Weighted Regression Model for Modeling Public Housing Prices in Singapore.

Phang, S.-Y. (2015). Singapore’s Housing Policies: Responding to the Challenges of Economic Transitions.

Mennis, Jeremy (2006) “Mapping the Results of Geographically Weighted Regression”, The Cartographic Journal, Vol.43 (2), p.171-179.

Stephen A. Matthews ; Tse-Chuan Yang (2012) “Mapping the results of local statistics: Using geographically weighted regression”, Demographic Research, Vol.26, p.151-166.

Benito B. (2025). spatialRF. GitHub. https://blasbenito.github.io/spatialRF/
