---
title: "09:Modelling Geographical Accessibility "
author: "Yiqiong PAN"
date: "Nov 04, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 4

editor: visual
---

## Introduction

Geographical accessibility refers to the ease with which individuals can physically reach certain locations, services, or opportunities, that is generally measured through distance or travel time, with consideration of transportation networks and spatial barriers.

In this exercise, we learn how to model geoprahpical accessibility by using R's geospatial analysis package.

### Learning Outcome:

-   Import GIS polygon data into R and save them as *simple feature data frame* by using appropriate functions of **sf** package of R;
-   Import aspatial data into R and save them as *simple feature data frame* by using appropriate functions of **sf** package of R;
-   Computer accessibility measure by using **Hansen’s potential model** and **Spatial Accessibility Measure** (SAM); and
-   Visualise the accessibility measures by using **tmap** and **ggplot2** packages.

### The Data

+--------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+
| Dataset Name             | Description                                                                                                                                                                                                                                                                                                         | Format         |
+==========================+=====================================================================================================================================================================================================================================================================================================================+================+
| `MP14_SUBZONE_NO_SEA_PL` | URA Master Plan 2014 subzone boundary GIS data, downloaded from [data.gov.sg](data.gov.sg).                                                                                                                                                                                                                         | ERSI Shapefile |
+--------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+
| `hexagons`               | A 250m radius hexagons GIS data, created by using [`st_make_grid()`](https://r-spatial.github.io/sf/reference/st_make_grid.html) of **sf** package.                                                                                                                                                                 | ESRI shapefile |
+--------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+
| `ELDERCARE`              | Locations of eldercare service, downloaded from [data.gov.sg](https://data.gov.sg/collections/714/view). Two formats were available, ESRI shapefile and Google KML. The Shapefile format was selected as it is a geospatial-native format that preserves coordinate systems and spatial attributes more accurately. | ESRI shapefile |
+--------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+
| `OD_Matrix`              | A travel distance matrix between the origin hexagon cells and eldercare service locations, measured by **metres**. It includes:                                                                                                                                                                                     | CSV            |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `origin_id`: hexagon identifier                                                                                                                                                                                                                                                                                 |                |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `destination_id`: eldercare facility identifier                                                                                                                                                                                                                                                                 |                |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `entry_cost`: perpendicular distance from the origin to nearest road entry                                                                                                                                                                                                                                      |                |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `network_cost`: actual travel distance on the road                                                                                                                                                                                                                                                              |                |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `exit_cost`: perpendicular distance from destination to nearest road connection                                                                                                                                                                                                                                 |                |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `total_cost`: total travel cost, computed as `entry_cost+network_cost+exit_cost`                                                                                                                                                                                                                                |                |
+--------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+

## Getting Started

There are **six** packages to be used in this exercise.

+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| Package                                                                                                                                                 | Description                                                                                                                |
+=========================================================================================================================================================+============================================================================================================================+
| [sf](https://r-spatial.github.io/sf/)                                                                                                                   | Handles importing, managing, and processing vector-based geospatial data                                                   |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| [tidyverse](https://www.tidyverse.org/)                                                                                                                 | A collection of R packages for data wrangling and visualisation, including **readr**, **dplyr**, and **ggplot2**.          |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| [tmap](https://r-tmap.github.io/tmap/)                                                                                                                  | Creates cartographic quality static and interactive thematic (choropleth) maps.                                            |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| [Spatial](https://cran.r-project.org/web/packages/SpatialML/index.html)[ACC](https://cran.r-project.org/web/packages/SpatialAcc/refman/SpatialAcc.html) | Provides a set of spatial accessibility measures from demand to supply                                                     |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| [ggstatsplot](https://indrajeetpatil.github.io/ggstatsplot/)                                                                                            | An extension of [`{ggplot2}`](https://github.com/tidyverse/ggplot2) package to create enriched plots for statistical tests |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| [reshape2](https://www.rdocumentation.org/packages/reshape2/versions/1.4.4)                                                                             | Provides tools to restructure and combine data.                                                                            |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+

Let's install and launch these packages into R environment.

```{r}
pacman::p_load(tmap, SpatialAcc, sf, ggstatsplot, reshape2, tidyverse)
```

## Geospatial Data Wrangling

### Importing Geospatial Data

The code chunk below uses `st_read()` of **sf** package to import the three ERSI Shapefile files into R as `sf` objects.

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL")
```

```{r}
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") 
```

```{r}
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE") 
```

The import reports reveal the geometry type of each file, and improtant information like dimension and projected CRS. We notice `mpsz` file does not have proper "SVY21/ Singapore TM".

### Updating CRS Information

The code chunk below assigns the correct ESPG(3414) to our newly imported `sf` objects.

```{r}
mpsz <- st_transform(mpsz, 3414)
eldercare <- st_transform(eldercare, 3414)
hexagons <- st_transform(hexagons, 3414)
```

Then we can verify the update via calling `st_crs()` of **sf** package. The EPSG has been updated accordingly.

```{r}
st_crs(mpsz)
```

Next we quickly plot two maps for hexagons and eldercare centers in Singapore.

```{r}
tm_shape(mpsz) +
  tm_polygons(alpha = 0.2, border.col = "gray") +
  tm_dots(col = "red", size = 0.1, alpha = 0.6) +
  tm_title("Eldercare Facilities Distribution")
```

```{r}
tm_shape(hexagons) +
  tm_polygons(col = "lightblue", alpha = 0.5, 
              border.col = "darkblue") +
  tm_title("Hexagonal Grid")
```

### Cleaning and Updating Attribute fields of the Geospatial Data

After inspecting the origin and destination datasets, we reorganise the attributes and retain only the required ID fields. We then create a **demand** field for the hexagon layer, representing the estimated number of people who require eldercare services within each hexagonal area. Likewise, we add a **capacity** field for the eldercare layer, which indicates the service capacity or supply at each facility (e.g., number of beds or service slots available).

For the purpose of this exercise, both demand and capacity are assigned a constant value of **100** to simplify the demonstration. In practice, actual population data and real facility capacity figures should be used.

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

## Apsaital Data Handling and Wrangling

### Importing Distance Matrix

The code chunk below uses `read_cvs()` of **readr** package to import the matrix dataset into R as a tibble data.frame .

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

### Tidying Distance Matrix

The imported ODMatrix is structured that the cost component appears as a separate column.

```{r}
knitr::kable(head(ODMatrix, 10))
```

A standardised matrix format stores origins as rows and destinations as columns, allowing the distance between any origin–destination pair to be retrieved by referencing the corresponding row and column indices.

This transformation can be performed using the `spread()` or `pivot_wider()` function from the **tidyr** package to convert the long (thin) table format into a wide (fat) matrix format.

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  pivot_wider(names_from = destination_id, values_from = total_cost) %>% #spread(destination, total_cost)
  select(-'origin_id')  #remove the column origin_id
```

Next, we rescale the distance values from metres to kilometres. `as.matrix()` is used here to convert the `tbl_df` data frame into a numeric `matrix` format.

```{r}
distmat_km <- as.matrix(distmat/1000)
```

## ac() in the SpatialAcc Package

ac() computes spatial accessibility of demand locations (e.g., residential areas) to service locations (e.g., eldercare centres). It implements several well-known measures:

-   "Hansen":Hansen gravity-type accessibility

-   "KD2SFCA": Kernel/Downgraded 2SFCA (distance-decay)

-   "SAM": Spatial Accessibility Measure

**Usage**:

-   ac(p, n, D, d0, power=2, family="SAM")

**Arguments**:

-   p: numeric vector of demand at each origin (e.g., number of senior residents per hexagon).

-   n: numeric vector of supply at each destination (e.g., number of beds per eldercare centre).

-   D: distance or time matrix separating origins (rows) from destinations (columns). Prefer road-network distance or travel time.

-   d0: catchment threshold (distance/time) defining which OD pairs “count”.

-   power: distance-decay exponent (commonly 2 in gravity models).

-   family: which accessibility method to compute: "SAM", "2SFCA", "KD2SFCA", or "Hansen" (default "SAM").

All four methods are **gravity-based** accessibility measures where accessibility **improves** with greater service capacity and proximity, and **decreases** with distance and population demand due to competition for limited resources.

**Return value**

A numeric vector of accessibility scores for each origin (length = length(p)).

## Modelling and Visualising Accessibility using Hansen Method

The code chunk below uses `ac()` to compute the Hansen’s accessibility and saves the output in a data frame by using data.frame().

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km,
                            power = 2,
                            family = "Hansen"))
```

Let's have a quick review on the Hansen accessibility measure, after rename the column name to `accHansen`.

```{r}
colnames(acc_Hansen) <- "accHansen" 
```

We convert the data table into tibble format by using the code chunk below.

```{r}
acc_Hansen <- acc_Hansen %>%
  as_tibble()
```

```{r}
knitr::kable(head(acc_Hansen, 10), 
             align = 'c',  # column alignment: c=center, l=left, r=right
             format = "html")  # or "latex", "markdown", "pandoc"
```

Lastly, we join the `acc_Hansen` `tibble` with the `hexagon` `sf` data frame using `bind_cols()`. Because the `sf` object is on the **left** side of the operation, the resulting dataset automatically inherits the **sf** class, preserving its spatial attributes.

```{r}
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

```{r}
knitr::kable(head(hexagon_Hansen,10),
             align = "c",
            format = "markdown")
```

### Visualising Hansen’s Accessibility

#### Extracting Map Extend

First, we extract the extent of the hexagon sf layer using `st_bbox()` of sf package.

```{r}
mapex <- st_bbox(hexagons)
```

Next, we create a choropleth map with the **tmap** package to visualise eldercare service accessibility across Singapore.

```{r}
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: Hansen method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

The map indicates that accessibility is highest in areas with clusters of eldercare centres, particularly in the Central Region and established heartland estates. In contrast, large peripheral areas exhibit zero accessibility.

### Statistical Graphic Visualisation

Next, we create a choropleth map with the **tmap** package to visualise eldercare service accessibility across Singapore.

In this section, we compare the distribution of Hansen accessiblity scores across URA Planning regions. To do so, we first append the main dataset `mpsz` to the `hexagon_Hansen` `sf` data frame using `st_join()`.

```{r}
hexagon_Hansen <- st_join(hexagon_Hansen, mpsz, join= st_intersects)
```

Then we use `ggplot()`, `geom_boxplot()` and `stat_summary()` , to create boxplots for each region and display the mean accessibility value.

```{r}
ggplot(data=hexagon_Hansen, 
       aes(y = log(accHansen), x= REGION_N)) +
  geom_boxplot() +
  stat_summary(fun="mean", colour ="red", size=0.5)
```

The boxplots of Hansen’s accessibility show that the Central Region has the highest average level of accessibility, followed by the North-East Region. In contrast, the West Region records the lowest average accessibility. The East Region displays the largest variability, as indicated by its long lower whiskers, suggesting uneven access to eldercare services across the region, possibly influenced by extensive non-residential areas, the airport.

The average Hansen accessibility scores rank the regions from highest to lowest as follows: Central, North-East, North, East, and West.

## Modelling and Visualising Accessibility using KD2SFCA Method

KD2SFCA stands for Kernel Density 2-Step Floating Catchment Area. It uses a distance decay function where accessibility decreases as distance increases and accounts for competing demand, which is commonly used in healthcare facility access.

In this section, we repeat the similar steps to compute and visualise the accessibility using KD2SFCA method.

```{r}
acc_KD2SFCA <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            d0 = 5,
                            power = 2, 
                            family = "KD2SFCA"))

colnames(acc_KD2SFCA) <- "accKD2SFCA"
acc_KD2SFCA <- as_tibble(acc_KD2SFCA)
hexagon_KD2SFCA <- bind_cols(hexagons, acc_KD2SFCA)
```

The d0 = 5 parameter in the ac() function represents the bandwidth or search radius in kilometers.

5 means the maximum distance within which eldercare facilities can serve the people in the hexagon area.

If beyond 5km, accessibility becomes zero.

The code chunk below displays the data frame `hexagon_KD2SFCA` that shows the eldercare service accessibility per hexagon.

```{r}
knitr::kable(head(hexagon_KD2SFCA, 10),
            format = "pandoc")
```

17.7.2 Visualising KD2SFCA’s Accessibility

We reuse the hexagon extent and create a high cartographic quality map for accessibility to eldercare centres in Singapore.

```{r}
tm_shape(hexagon_KD2SFCA,
         bbox = mapex) + 
  tm_polygons(fill = "accKD2SFCA", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: KD2SFCA method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

The KD2SFCA-based accessibility map shows a spatial pattern similar to the Hansen model, although the magnitude of accessibility scores differs.

### Statistical Graphic Visualisation

In this section, we compare the distribution of KD2SFCA accessiblity scores across URA Planning regions. To do so, we first append the main dataset `mpsz` to the `hexagon_KD2SFCA` `sf` data frame using `st_join()`.

```{r}
hexagon_KD2SFCA <- st_join(hexagon_KD2SFCA, mpsz, join = st_intersects)
```

Then we use `ggplot()`, `geom_boxplot()` and `stat_summary()` , to create boxplots for each region and display the mean accessibility value.

```{r}
ggplot(data=hexagon_KD2SFCA, 
       aes(y = log(accKD2SFCA), x= REGION_N)) +
  geom_boxplot() +
  stat_summary(fun="mean", colour ="red", size=0.5)
```

The boxplot for the KD2SFCA method shows that the average accessibility scores, from highest to lowest, are ranked as follows: Central, North-East, West, East, and North.

## Modelling and Visualising Accessibility using Spatial Accessibility Measure (SAM) Method

### Computing SAM Accessibility

We repeat the similar code to calculate SAM Accessibility.

```{r}
acc_SAM <- data.frame(ac(hexagons$demand,
                         eldercare$capacity,
                         distmat_km, 
                         d0 = 5,
                         power = 2, 
                         family = "SAM"))

colnames(acc_SAM) <- "accSAM"
acc_SAM <- as_tibble(acc_SAM)
hexagon_SAM <- bind_cols(hexagons, acc_SAM)
```

```{r}
knitr::kable(head(hexagon_SAM, 10),
            format = "markdown")
```

### Visualising SAM’s Accessibility

Similarly we plot the choropleth map for SAM's accessibility.



```{r}
tm_shape(hexagon_SAM,
         bbox = mapex) + 
  tm_polygons(fill = "accSAM", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: SAM method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```


The SAM map displays a spatial pattern very similar to the Hansen results. However, the SAM values are much more interpretable, instead of many zero values and long decimal numbers, the scores range from approximately 0.1 to 740, making the scale more user-friendly.

### Statistical Graphic Visualisation

In this section, we compare the distribution of SAM accessiblity scores across URA Planning regions. To do so, we first append the main dataset `mpsz` to the `hexagon_SAM` `sf` data frame using `st_join()`.

```{r}
hexagon_SAM <- st_join(hexagon_SAM, mpsz, join = st_intersects)
```

Then we use `ggplot()`, `geom_boxplot()` and `stat_summary()` , to create boxplots for each region and display the mean accessibility value.

```{r}
ggplot(data=hexagon_SAM, 
       aes(y = log(accSAM), x= REGION_N)) +
  geom_boxplot() +
  stat_summary(fun="mean", colour ="red", size=0.5)
```

SAM accessibility ranking (from highest to lowest) (consistent with Hansen): Central,North-East, North, East and West.

The Central Region contains two notable outliers, Crawford and Lavender from Kallang subzone, with SAM accessibility scores of approximately 740, far exceeding values observed elsewhere.

## Reference

Kam, T. S. 17 Modelling Geographical Accessibility. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap17.html
