---
title: "In-class Ex05"

author: "Yiqiong PAN"
date: "Sep 27, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 4

editor: visual
---

## Overview

[sfdep](https://sfdep.josiahparry.com/) built on [spdep](https://cran.r-project.org/web/packages/spdep/index.html) ( tools for computing spatial weights and calculating spatially lagged variables), provides a tidyverse-friendly interface with list-columns and additional functionality, more importantly inlcuding functions to analyse spatial temporal data.

```{r}
pacman:: p_load(sf, tmap, tidyverse, sfdep)
```

## Getting the Data into R environment

:::panel-tabset
### Importing Geospatial data in Shapefile

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "hunan")
st_crs(hunan)
```

```{r}
hunan <- hunan %>%
  st_transform(crs = 32650)
st_crs(hunan)
```

### Importing Aspatial data in csv

```{r}
hunan2012 <- read_csv ("data/aspatial/hunan_2012.csv")
```

### Performing Rational Join

```{r}
hunan_GDPPC <- left_join(hunan, hunan2012) %>%
  select(1:4, 7,15)
```



### Visualising Regional Development Indicator

The code chunk below uses [`qtm()`](https://jamescheshire.github.io/learningR/mapping-in-r.html) from **tmap** to plot base and choropleth maps to show the ditribtution of GDPPC 2012 in Hunan. **Equal interval groups by same range of values whereas Quantile defines the group by rank/count, but value ranges may vary.**

More mapping tips can be found [here](https://r.geocompx.org/adv-map#animated-maps).

```{r, fig.width=12, fig.height=8}
equal <- tm_shape(hunan_GDPPC) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues"
              )) +
  tm_borders(fill_alpha= 0.5) + 
  tm_layout(legend.position = c("left", "bottom"),
            main.title = "Equal interval classification")

quantile <- tm_shape(hunan_GDPPC) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(legend.position = c("left", "bottom"),
            main.title = "Quantile classfication")

tmap_arrange(equal, quantile, asp = 1, ncol = 2)

```
:::


## Global Measures of Spatial Association


### Computing Contiguity Spatial Weights

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(
           nb, style = "W"),
         .before = 1)
head(wm_q)
```

### Computing Global Moran's I

```{r}
set.seed(1234)
```

```{r}
global_moran_perm(wm_q$GDPPC,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99) #100 simulations since the it starts with 0.
```

**At the 95% confidence level, p-value is less than 0.05, we reject the null hypothesis. Moran's I value is 0.3 which is a sign of moderate clustering.**


:::callout-caution

When we fail to reject the null hypothesis, we should continue to examine further, since there is still possibility that at local level there are hot spot and cold spot areas.

:::


## Emerging Hot Spot Analysis (EHSA)

- Build a space–time cube.
- Calculate local Gi*.
- Evaluate the hot/cold-spot trend using the Mann–Kendall test.
Categorise each study-area location by combining the trend (MK z-score and p-value) with the - per-bin hot/cold status (Gi* z-score and p-value).

```{r}
pacman:: p_load(sf, tmap, tidyverse, sfdep)
```


```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "hunan")
st_crs(hunan)
```


- Build a space–time cube.

:::callout-important
The GDPPC data set must follow certain orders to avoid the errors (time (integer), space, value).

:::

```{r}
GDPPC_st <- spacetime(GDPPC, hunan, .loc_col = "County", .time_col ="year")
#is_spacetime
```


- Calculate local Gi*.

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>% ## this is important
  mutate ()
  
```



Gi* is computed per year (space–time cube bins). We report both z-scores and p-values; bins with p≥0.05 are labelled Insignificant.

Mann–Kendall (MK) is a non-parametric trend test:
H₀: no monotonic trend (flat).
H₁: a monotonic trend exists (consistently increases or decreases).

Kendall’s 
τ ranges from −1 to +1 (direction = sign; strength = magnitude).

In Kendall::MannKendall, sl is the p-value.

Outliers/clusters vs hot/cold spots: Local Moran’s I (LISA) finds high–high/low–low clusters and outliers; EHSA/Gi* focuses on hot/cold spots.

We select counties manually to show their Gi* time series and MK result (trend line figure).

Bar chart: summarises counts (or proportions) of Hot, Cold, Insignificant bins by year (analogous to a Moran’s-I summary, but based on Gi* z-scores/p-values).
group_by(Year) # that is why the table format is useful that we can calculate by year.


Visualising EHSA should incorporate the p-value.
Update the plot on slide p18 to include an “Insignificant” legend entry, consistent with the HCSA map in Chapter 10.

Master the explanation chart for hot/cold-spot definitions.
For example, for a Sporadic Hot Spot, point to the region on p18 and cross-reference its Gi* trend line on p11.

Select sample locations for each category to explain the spatial patterns.

Be careful with the steps (document thresholds, weights, and binning).

