---
title: "In-class Ex04"
author: "Yiqiong PAN"
date: "Sep 20, 2025"
date-modified: "last-modified"
execute:
  echo: true #display the code
  eval: true #run the code
  message: false
  warning: false
  freeze: true # not render if nothing edited
editor: visual
format: 
  html:
    number-sections: true
    number-depth: 4
    code-link: true
    toc: true
---

## Overview

In this exercise, we apply spatial weights in a practical implementation in R.

Spatial data are rarely independent which often violates the independence assumptions of classical statistics (e.g., regression or ANOVA).Properly defining and weighting neighbours is crucial to account for spatial autocorrelation and avoid biased outcomes. Through it, we can understand who influences whom and how strong the influence is.

## The Study Area and Data

The following two data sets are used here:

1.  Geospatial data: Hunan county boundary layer in `shapefile` format

2.  Aspatial data: `Hunan_2012.csv` for Hunan’s local development indicators in 2012.

### Getting started

A total of **five** R packages will be used in this exercise.

| Package | Description |
|------------------------------------|------------------------------------|
| [sf](https://r-spatial.github.io/sf/) | Simple Features, a new R package which handles importing, managing, and processing vector-based geospatial data. |
| [spdep](https://cran.r-project.org/web/packages/spdep/index.html) | Tools for computing spatial weights and calculating spatially lagged variables. |
| [knitr](https://cran.r-project.org/web/packages/knitr/index.html) | Provides useful functions for dynamic document generation. |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static or interactive choropleth maps. |
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data import, cleaning, transformation, and visualisation (e.g., readr, dplyr, tidyr, ggplot2). |

: {tbl-colwidths="\[15,85\]"}

After installation via `install.packages()`, we load them into R environment using the code below.

```{r}
pacman:: p_load(sf, spdep, knitr, tmap, tidyverse, GWmodel, ggstatsplot)
```

## Importing Data into R

In this section, we import above-mentioned two data sets into the R environment by using `st_read()` from `sf` package for geospatial data and `read_csv()` from `readr` of `tiyeverse` family.

### Importing `sharefile` File

The code chunk below uses `st_read()` to import the Hunan county layer in `sharefile` format to R as an `sf` object for spatial analysis.

```{r}
hunan_sf <- st_read(dsn = "data/geospatial",
                layer = "Hunan")
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")

hunan_sf <- hunan_sf %>%
  left_join(hunan2012, by = "County")
hunan_sf <- hunan_sf %>%
   select(1:3, 7, 15, 16, 31, 32)

hunan_sp <- hunan_sf %>%
  as_Spatial()

```

```{r}
bw_CV <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "CV",
                adaptive = FALSE,
                kernel = "bisquare",
                longlat = T)
bw_CV
```                
  
  
```{r}
                
bw_AIC <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "AIC",
                adaptive = FALSE,
                kernel = "bisquare",
                longlat = T)
```


```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               adaptive = FALSE,
              longlat = T)
```


```{r}
gwstat_df <- as.data.frame((gwstat$SDF))
```


Since there is no common clumn which is essental to left_join, we can use cbind() when joining two tables together without unique identifiers, given the sequence is same. 


```{r}
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```


```{r}
tm_shape(hunan_gstat) +
  tm_fill("GDPPC_LM", n = 5, style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Distribution of Geographically Weighted Mean",
            main.title.position = "center",
            main.title.size = 2.0,
            legend.text.size = 1.2,
            legend.height = 1.5,
  legend.show = FALSE   # <- turns legend off
  )
```








