---
title: "In-class Ex04"
author: "Yiqiong PAN"
date: "Sep 20, 2025"
date-modified: "last-modified"
execute:
  echo: true #display the code
  eval: true #run the code
  message: false
  warning: false
  freeze: true # not render if nothing edited
editor: visual
format: 
  html:
    number-sections: true
    number-depth: 4
    code-link: true
    toc: true
---

## Overview

**Geographically Weighted Summary Statistics (GWSS)** compute local measures such as mean, standard deviation, or median. Nearby observations are given higher weights (via a kernel function and bandwidth), so each location reflects its own neighbourhood. This highlights spatial heterogeneity by revealing how statistical patterns vary across space.

##Application of GWSS

GWSS helps us see how relationships between data change in areas like the environment, health, and socio-economics. It can be used to study changes in ecosystems, find differences in health across regions, connect human activities to changes in the landscape, and spot unusual patterns in soil or minerals. Instead of giving one overall result, GWSS shows local patterns, making it easier to design policies and manage resources in a more focused way.

## The Study Area and Data

The following two data sets are used here:

1.  Geospatial data: Hunan county boundary layer in `shapefile` format

2.  Aspatial data: `Hunan_2012.csv` for Hunan’s local development indicators in 2012.

### Getting started

A total of **seven** R packages will be used in this exercise.

| Package | Description |
|------------------------------------|------------------------------------|
| [sf](https://r-spatial.github.io/sf/) | Simple Features, a new R package which handles importing, managing, and processing vector-based geospatial data. |
| [spdep](https://cran.r-project.org/web/packages/spdep/index.html) | Tools for computing spatial weights and calculating spatially lagged variables. |
| [knitr](https://cran.r-project.org/web/packages/knitr/index.html) | Provides useful functions for dynamic document generation. |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static or interactive choropleth maps. |
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data import, cleaning, transformation, and visualisation (e.g., readr, dplyr, tidyr, ggplot2). |
| [GWmodel](https://www.tidyverse.org/) | A R package for Geographically-Weighted Models (e.g., GW summary statistics,GW principal components analysis, GW discriminant analysis,GW principal components analysis. |
| [ggstatsplot](https://indrajeetpatil.github.io/ggstatsplot/) | An extension packages to ggplot to visualise statisital tests (e.g., violin plots, histograms, scatterplots, correlation matrices, bar charts, etc). |

: {tbl-colwidths="\[15,85\]"}

After installation via `install.packages()`, we load them into R environment using the code below.

```{r}
pacman:: p_load(sf, spdep, knitr, tmap, tidyverse, GWmodel, ggstatsplot)
```

## Importing Data into R

In this section, we import above-mentioned two data sets into the R environment by using `st_read()` from `sf` package for geospatial data and `read_csv()` from `readr` of `tiyeverse` family.

### Importing `sharefile` File

The code chunk below uses `st_read()` to import the Hunan county layer in `sharefile` format to R as an `sf` object for spatial analysis.

```{r}
hunan_sf <- st_read(dsn = "data/geospatial",
                layer = "Hunan")
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")

hunan_sf <- hunan_sf %>%
  left_join(hunan2012, by = "County")
hunan_sf <- hunan_sf %>%
   select(1:3, 7, 15, 16, 31, 32)
```

### Visualisation

The code chunk below plots a base map of Hunan Counties and a thematic map for geographic distribution of GDPPC of Hunan, side by side for quick overview.

```{r}
tmap_mode("plot")
```

```{r, fig.width=14, fig.height=7}

basemap <- tm_shape(hunan_sf) +
  tm_polygons() +
  tm_text("NAME_3", size=0.4)

gdppc <- qtm(hunan_sf, "GDPPC") +
tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp = 1, ncol =2)

```


### Converting to SpatialPolygonDataFrame

**GWmodel** requires spatial dataframe i.e. `SpatialPointsDataFrame` or` SpatialPolygonsDataFrame` as defined in polder ackage `sp`. Here we transfer the `sf` data frame to `sp` data frame by using `as_Spatial()`.

```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()

```


## Geographically Weighted Summary Statistics (GSWW) with adaptive bandwidth

### Determine adaptive bandwidth

The code chunk below uses `bw.gwr()` for bandwith selection to calibrate a basic **GWR** model. 
- `kernel` could be given by `gaussian`, `bisquare`, `exponential`, `tricube`, and `boxcar`. 
- `approach` is specified by `CV` for cross-validation approach or by `AIC` corrected (AICc) approach.
- `adaptive` is `FALSE` by default. Here we enforce it to compute an adaptive kernel.

```{r}
bw_CV <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T) #if TRUE, great circle distances will be calculated
bw_CV
```
```{r}
                
bw_AIC <- bw.gwr(GDPPC ~ 1, #~ 1 says: model GDPPC as the dependent variable, without predictors.
                data = hunan_sp,
                approach = "AIC",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
bw_AIC
```
### Computing geographically wieghted summary statistics (GWSS)

The code chunk below calculate basic and robust GWSS including geographically weighted means,
standard deviations and skew. 

```{r}
gwstat_bisquare <- gwss(data = hunan_sp,
               vars = "GDPPC", #a vector of variable names to be summarised
               bw = bw_AIC, 
               kernel ="bisquare",
               quantile = TRUE,
               adaptive = TRUE,
               longlat = T)

gwstat_bisquare
```


The code chunk below compute the GWSS by other four kernel methods.

```{r}
gwstat_gaussian <- gwss(data = hunan_sp,
               vars = "GDPPC", #a vector of variable names to be summarised
               bw = bw_AIC, 
               kernel ="gaussian",
               quantile = TRUE,
               adaptive = TRUE,
               longlat = T)

gwstat_gaussian
```


```{r}
gwstat_exponential <- gwss(data = hunan_sp,
               vars = "GDPPC", #a vector of variable names to be summarised
               bw = bw_AIC, 
               kernel ="exponential",
               quantile = TRUE,
               adaptive = TRUE,
               longlat = T)

gwstat_exponential
```

```{r}
gwstat_tricube <- gwss(data = hunan_sp,
               vars = "GDPPC", #a vector of variable names to be summarised
               bw = bw_AIC, 
               kernel ="tricube",
               quantile = TRUE,
               adaptive = TRUE,
               longlat = T)

gwstat_tricube
```

```{r}
gwstat_boxcar <- gwss(data = hunan_sp,
               vars = "GDPPC", #a vector of variable names to be summarised
               bw = bw_AIC, 
               kernel ="boxcar",
               quantile = TRUE,
               adaptive = TRUE,
               longlat = T)

gwstat_boxcar
```



### Preparing the output data


:::panel-tabset

#### Plotting with gaussian method 

The code chunk converts the extracted SPF data to data.frame.

```{r}
gwstat_gaussian_df <- as.data.frame((gwstat_gaussian$SDF))
```

Since there is no common column which is essential to `left_join()`, we can use `cbind()` when joining two tables together without unique identifiers, given the sequence is same.

```{r}
hunan_gstat_gaussian <- cbind(hunan_sf, gwstat_gaussian_df)
head(hunan_gstat_gaussian)
```

```{r}
ad_g <- tm_shape(hunan_gstat_gaussian) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically wieghted mean (gaussian)",
            main.title.position = "center",
            main.title.size = 2.0,
            frame = TRUE)
ad_g
```

#### Plotting GWSS with Bisquare Method


The code chunk converts the extracted SPF data to data.frame.

```{r}
gwstat_bisquare_df <- as.data.frame((gwstat_bisquare$SDF))
```

Since there is no common column which is essential to `left_join()`, we can use `cbind()` when joining two tables together without unique identifiers, given the sequence is same.

```{r}
hunan_gstat_bisquare <- cbind(hunan_sf, gwstat_bisquare_df)
head(hunan_gstat_bisquare)
```

```{r}
ad_bi <- tm_shape(hunan_gstat_bisquare) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically wieghted mean (bisquare)",
            main.title.position = "center",
            main.title.size = 2.0,
            frame = TRUE)
ad_bi 
```

#### Plotting GWSS with Exponential Method


The code chunk converts the extracted SPF data to data.frame.

```{r}
gwstat_exponential_df <- as.data.frame((gwstat_exponential$SDF))
```

Since there is no common column which is essential to `left_join()`, we can use `cbind()` when joining two tables together without unique identifiers, given the sequence is same.

```{r}
hunan_gstat_exponential <- cbind(hunan_sf, gwstat_exponential_df)
head(hunan_gstat_exponential)
```

```{r}
ad_exp <- tm_shape(hunan_gstat_exponential) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically wieghted mean (exponential)",
            main.title.position = "center",
            main.title.size = 2.0,
            frame = TRUE)

ad_exp 
```

#### Plotting GWSS with Tricube Method

The code chunk converts the extracted SPF data to data.frame.

```{r}
gwstat_tricube_df <- as.data.frame((gwstat_tricube$SDF))
```

Since there is no common column which is essential to `left_join()`, we can use `cbind()` when joining two tables together without unique identifiers, given the sequence is same.

```{r}
hunan_gstat_tricube <- cbind(hunan_sf, gwstat_tricube_df)
head(hunan_gstat_tricube)
```

```{r}
ad_tri <-tm_shape(hunan_gstat_tricube) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically wieghted mean (tricube)",
            main.title.position = "center",
            main.title.size = 2.0,
            frame = TRUE)

ad_tri
```

#### Plotting GWSS with Boxcar Method

The code chunk converts the extracted SPF data to data.frame.

```{r}
gwstat_boxcar_df <- as.data.frame((gwstat_boxcar$SDF))
```

Since there is no common column which is essential to `left_join()`, we can use `cbind()` when joining two tables together without unique identifiers, given the sequence is same.

```{r}
hunan_gstat_boxcar <- cbind(hunan_sf, gwstat_boxcar_df)
head(hunan_gstat_boxcar)
```

```{r}
ad_bc <-tm_shape(hunan_gstat_boxcar) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically wieghted mean (boxcar)",
            main.title.position = "center",
            main.title.size = 2.0,
            frame = TRUE)

ad_bc
```
#### Plotting Summary


```{r fig.width= 18, fig.height=16}
tmap_arrange(gdppc, ad_g, ad_bi, ad_exp, ad_tri, ad_bc, asp = 1, ncol = 2)
```

Visually, Gaussian/ exponential shows smooth distribution, while tricube/ bisquare shows relatively less smooth. And Boxcar is not as jumpy as bisquare but also not smooth as gaussian.

:::

## GWSS with Fixed Bandwidth

### Determine fixed bandwidth


```{r}
bw_CV <- bw.gwr(GDPPC ~ 1, 
             data = hunan_sp,
             approach = "CV",
             adaptive = FALSE, 
             kernel = "bisquare", 
             longlat = T)

bw_CV
```

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1, 
             data = hunan_sp,
             approach ="AIC",
             adaptive = FALSE, 
             kernel = "bisquare", 
             longlat = T)
bw_AIC
```

### Computing adaptive bandwidth GWSS

```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare",
               adaptive = FALSE,
               quantile = FALSE,
               longlat = T)
```

### Preparing the output data

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
```


```{r}
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```


```{r}
tm_shape(hunan_gstat) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically wieghted mean",
            main.title.position = "center",
            main.title.size = 2.0,
            frame = TRUE)
```

## Geographically Weighted Correlation

Business question: Is there any relationship between GDP per capita and Gross Industry Output?

The code chunk below generates a scatter plot for conventional statistical solution. That with p-value less than the alpha, the statistically significant relationship between GDPPC and Gross Agriculture Output is evident with 95% confidence.

```{r}
ggscatterstats(
  data = hunan2012, 
  x = Agri, 
  y = GDPPC,
  xlab = "Gross Agriculture Output", ## label for the x-axis
  ylab = "GDP per capita", 
  label.var = County, 
  label.expression = Agri > 10000 & GDPPC > 50000, 
  point.label.args = list(alpha = 0.7, size = 4, color = "grey50"),
  xfill = "#CC79A7", 
  yfill = "#009E73", 
  title = "Relationship between GDP PC and Gross Agriculture Output")
```


## Geospatial analytics solution

First we determine the adaptive bandwidth using "AICc" approach, calculate *gwCorrelation* using `gwss()`, extract the result from `SDF` field then convert the data into `data.frame`. Finally we append the the new sf data with main `hunan` sf data.

```{r}
bw <- bw.gwr(GDPPC ~ GIO, #GIO is predictor
             data = hunan_sp, 
             approach = "AICc", 
             adaptive = TRUE)
```

```{r}
gwstats <- gwss(hunan_sp, 
                vars = c("GDPPC", "GIO"), 
                bw = bw,
                kernel = "bisquare",
                adaptive = TRUE, 
                longlat = T)
```


```{r}
gwstat_df <- as.data.frame(gwstats$SDF) %>%
  select(c(12,13)) %>%
  rename(gwCorr = Corr_GDPPC.GIO, #gwCorr, linear 
         gwSpearman = Spearman_rho_GDPPC.GIO) #gwSpearman, non-linear
```

```{r}
hunan_Corr <- cbind(hunan_sf, gwstat_df)
```


### Visualising Local Correlation

The code Chunk below visualises the locate correlations.

```{r}
tm_shape(hunan_Corr) +
  tm_polygons(fill = "gwSpearman",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Spearman Rho")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Spearman Rho", 
  size = 2.0) +
  tm_layout(frame = TRUE)
```
```{r}

tm_shape(hunan_Corr) +
  tm_polygons(fill = "gwCorr",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Correlation Coefficient")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Correlation Coefficient", 
  size = 2.0) +
  tm_layout(frame = TRUE)
```

Based on GWSS, the correlation between GDPPC and GIO in Hunan is strong and largely linear, as indicated by higher local Pearson coefficients compared to Spearman’s rho. This suggests that increases in GIO are generally associated with increases in GDPPC across the province. Wealthier counties such as Changsha, as well as the western highlands (less suitable for agriculture and less developed), show relatively lower correlations, whereas the surrounding counties exhibit stronger correlations.