---
title: "01: Geospatial Analytics for Public Good"
draft: true
author: "Yiqiong PAN"
date: "Sep 10, 2025"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 4

editor: visual
---

## Overview

Motivation

**Spatial Point Pattern Analysis** (SPPA) refers to the study of how points are arranged or distributed across a given surface. There points may represent:

-   **Events**, such as crimes, road accidents, or disease occurrences, or

-   **Service and facility locations**, including shops (e.g., cafes, supermarkets) and community facilities like childcare or aged care centres.

**First-order** **Spatial Point Pattern Analysis** (1-st SPPA) examines the *intensity* or *density* of points across a study area. It identifies spatial trends in point distribution without considering interaction between points helping to answer questions like:

-   Where are points most concentrated?

-   Is density uniform or variable?

-   How dispersed is the pattern?

Unlike first-order analysis, which looks only at overall density independently, second-order methods reveal relationships and influences between centres (points) based on distance.

A **spatio-temporal point process** is a random set of points, each showing when and where an event happens. Examples include disease cases, animal sightings, or natural disasters such as fires and earthquakes.

As more data is collected with both time and location, analysing spatio-temporal patterns has become more important. In recent years, several R packages have been developed for this purpose.

This exercise demonstrates how to apply the process through a case study of **New Business** in Singapore, from 1 January to 30 June 2025.

Since the period covers all of 2024 and first half of 2025, we split the data into two groups and compare them: Baseline: 2024 (Jan–Dec) Comparison: 2025 (Jan–Jun) This let us check whether some SSIC groups are growing faster in 2025, or whether patterns are stable.

## objective

The key questions we aim to address are:

**1. Are new business locations in Singapore independent in space and time?**

**2. If not, in which areas and during which periods do the forest fires tend to cluster?**

## The Data

For this exercise, the two datasets used are as follows:

1.  Master Plan 2019 Subzone Boundary (No Sea) (KML) contains xxx, downloaded from [data.gov.sg](http://data.gov.sg).

2.  **Corporate Entities dataset** contains corporate entitites regisretarion date since... Aug 2025, among with we extracted from duration and study buness based on SSIC 2020. THe orgindal data is downed from data.gov.sg

3.   Master Plan

## Installing and Loading the R Packages

A total of **six** R packages will be used in this exercise.

| Package | Description |
|------------------------------------|------------------------------------|
| [sf](https://r-spatial.github.io/sf/) | Simple Features, a new R package which handles importing, managing, and processing vector-based geospatial data. |
| [raster](https://rspatial.org/raster/) | Tools for reading, writing, and analyzing raster (gridded) spatial data in R |
| [spatstat](https://spatstat.org/) | Provides useful functions for SPPA, including kcross, Lcross etc. |
| [sparr](https://tilmandavies.github.io/sparr/index.html) | Functions for fixed/adaptive kernel density estimation and relative risk mapping via density ratios; also supports fixed-bandwidth space-time density and risk estimation with inference. |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static or interactive choropleth maps. |
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data import, cleaning, transformation, and visualization (e.g., readr, dplyr, tidyr, ggplot2). |
| [terra](https://rspatial.github.io/terra/index.html) | Modern package for raster/vector spatial data and will be used to convert spastat outputs into terra format. |

| stopp for 2rd order ST SPPA \|

| stpp \|

| magick for animation \|

| fs for renaming \|

: {tbl-colwidths="\[15,85\]"}

After installation via `install.packages()`, we load them into R environment using the code below.

```{r}
pacman:: p_load(sf, raster, rvest, spatstat, sparr, tmap, tidyverse, terra, stopp, stpp, magick, httr, ggthemes, plotly)
```

## Importing and Preparing Study Area

::: panel-tabset

### Importing Study Area

The following code chunk shows the steps to first import the `Master plan 2019 Subzone Boundary (No Sea)` data using **st_read()** function from sf package, extract the required 4 columns from the `Description` field, filter out the nearby islands, and finally save the file as `mpsz_cl` for further analysis.

```{r}
#| eval: false
mpsz <- st_read("data/geospatial/rawdata/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)


summary(mpsz)

tm_shape(mpsz) +
  tm_polygons()
```


```{r}
#| eval: false
mpsz$Description[200]
```


```{r}
#| eval: false
extract_kml_field <- function(html_text, field_name) {
  # return NA if missing or empty
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)          #parse the HTML string 
  rows <- page %>% html_elements("tr")  #get all table rows <tr>
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>% #keep row where <th> table header text = field_name
    html_element("td") %>%                                       #take its <td> table data
    html_text2()                                                 #extract clean text
  
  if (length(value) == 0) NA_character_ else value        #return NA if no match, otherwise the value
}
```


```{r}
#| eval: false
mpsz <- mpsz %>%
  mutate(
    REGION_N   = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N  = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C  = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```


```{r}
#| eval: false
mpsz_cl <- mpsz %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")

write_rds(mpsz_cl,
          "data/rds/mpsz_cl.rds")
```

```{r}
mpsz_cl <- read_rds("data/rds/mpsz_cl.rds")
summary(mpsz_cl)
```

```{r}
tmap_mode ("plot")
tm_shape(mpsz_cl) + tm_polygons()

```



:::



```{r}
#| eval: false
lu2019 <- st_read("data/geospatial/rawdata/MasterPlan2019LandUselayer.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

```{r}
#| eval: false
lu2019$Description[1]
```

```{r}
#| eval: false
lu2023 <- st_read("data/geospatial/rawdata/AmendmenttoMasterPlan2019LandUselayer.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
lu2023$Description[1:2] 
```


```{r}
#| eval: false
st_crs(lu2019) == st_crs(lu2023)
lu2019 <- st_make_valid(lu2019)
lu2023 <- st_make_valid(lu2023)
lu2019_updated <- lu2019[!st_intersects(lu2019, st_union(lu2023), sparse = FALSE)[,1], ]
lu_combined <- rbind(lu2019_updated, lu2023)

```


```{r}
nrow(lu2019); nrow(lu2019_updated); nrow(lu2023); nrow(lu_combined)

# Coverage check: does 2023 essentially cover all 2019?
a2019 <- st_area(st_union(st_make_valid(lu2019)))
a2023 <- st_area(st_union(st_make_valid(lu2023)))
as.numeric(a2023 / a2019)

```

```{r}
# parse LU_DESC from the Description column of land_use
lu <- lu_combined %>%
  mutate(
    LU_DESC = map_chr(Description, extract_kml_field, "LU_DESC"),
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

```


```{r}
table(lu$LU_DESC)

```

```{r}
lu_use <- lu %>%
  filter(!(LU_DESC %in% c("ROAD", "WATERBODY")))

```


```{r}

if (st_crs(lu_use) != st_crs(mpsz_cl)) {
  mpsz_cl <- st_transform(mpsz_cl, st_crs(lu_use))
}

# Option A: centroid join (faster)
lu_cent <- st_centroid(lu_use)

lu_with_subzone <- st_join(
  lu_cent %>% select(LU_DESC),
  mpsz_cl  %>% select(REGION_N, PLN_AREA_N, SUBZONE_N),
  join = st_within, left = TRUE
)

# Now every LU parcel has a SUBZONE_N, PLN_AREA_N, REGION_N

```
```{r}
lu_counts_subzone <- lu_with_subzone %>%
  st_drop_geometry() %>%
  count(PLN_AREA_N, SUBZONE_N, LU_DESC, sort = TRUE)

head(lu_counts_subzone, 20)

```



::: panel-tabset
### Business registry


#### Importing the ACRA Data
The code chunk imports ACRA data into R enviroment as tibble data frame. We also need to drop unnecessary fields, e.g., those with excessive missing values or containing only a single value.


```{r}
#| eval: false
folder_path <- "data/aspatial/acra"
file_list <- list.files(path = folder_path,
                        pattern = "^ACRA.*\\csv$",
                        full.names = TRUE)

acra_data <- file_list %>%
  map_dfr(read_csv,show_col_types = FALSE)
```


#### Saving the ACRA Data
The code chunk saves the consolidated ACRA data.

```{r}
#| eval: false
write_rds(acra_data, "data/rds/acra_data.rds") #rds not csv to avoid changing of field names..
```

```{r}
#| eval: false
acra_data <- read_rds("data/rds/acra_data.rds")
```

Selecting the Target Business and Period
The code chunk selects fields for the target business according to the SSIC code and registrations within the first six months of 2025. It renames long column names to ‘date’ for easier coding later, creates a new year column and both numeric and categorical month columns, and also tidies the postal code by adding a leading ‘0’.

```{r}
acra_2025 <- acra_data %>%
  select(1:24) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE),
         postal_code = str_pad(postal_code, width = 6, side = "left", pad = "0")) %>%
  filter(YEAR == 2025, MONTH_NUM <= 6)
```


```{r}
write_rds(acra_2025, "data/rds/acra_2025.rds")
```

```{r}
acra_2025 <- read_rds("data/rds/acra_2025.rds")
```

37362 vs 37012

```{r}
acra_2025l <- acra_2025 %>%
  filter(str_detect(entity_status_description, "Live"))
```


```{r}
write_rds(acra_2025l, "data/rds/acra_2025l.rds")
acra_2025l <- read_rds("data/rds/acra_2025l.rds")
```

```{r}
top20_codes <- acra_2025l %>%
  count(primary_ssic_code, sort = TRUE) %>%
  slice_max(n , n =20)   # keep top 20 codes
```


```{r}
ssic_lookup <- read_excel("data/aspatial/ssic2020-classification-structure.xlsx", skip = 4)%>%
  select(ssic_code = `SSIC 2020`, ssic_title = `SSIC 2020 Title`) %>%
  filter(!is.na(ssic_code))
```

```{r}
top20_codes <- top20_codes %>%
  mutate(primary_ssic_code = as.character(primary_ssic_code)) %>%
  left_join(ssic_lookup, by = c("primary_ssic_code" = "ssic_code"))
```


Following business types are selected as study of interest:


1. 46900 – Wholesale trade of a variety of goods without a dominant product
2. 62011 – Information technology consultancy (except cybersecurity)
3. 56140 - Stalls selling cooked food and prepared drinks (including stalls at food courts and mobile food hawkers)

Since they are among top 20, closed rate less than 1%



```{r}
# Define the SSIC codes of interest
target_codes <- c("46900", "62011", "56140", "47801", "47802" )

# Count in full dataset
counts_full <- acra_2025 %>%
  count(primary_ssic_code, name = "count_full") %>%
  filter(primary_ssic_code %in% target_codes)

# Count in live-only dataset
counts_live <- acra_2025l %>%
  count(primary_ssic_code, name = "count_live") %>%
  filter(primary_ssic_code %in% target_codes)

# Combine side by side
counts_compare <- counts_full %>%
  full_join(counts_live, by = "primary_ssic_code") %>%
  mutate(precentage =round((count_full- count_live)/ count_full * 100,2))%>%
  arrange(primary_ssic_code)

counts_compare
```



```{r}
acra_sl <- acra_2025l %>%
  select(3,17,22,25:27)

```


```{r}
biz_46900 <- acra_sl %>% filter(primary_ssic_code == "46900")
biz_62011 <- acra_sl %>% filter(primary_ssic_code == "62011")
biz_56140 <- acra_sl %>% filter(primary_ssic_code == "56140")
```


GEOCODE FOR EACH CODE


Performing Geocoding
The code chunk extracts unique postal codes from the postal_code field of a tidied tibble, queries the *SLA OneMap* `Reverse Geocode API` for each postal code. Then it store results in two data frames: - `found`: records with valid geocode results. - `not_found`: records where postal codes are not returned by the API.Clearning up the Geocoding
The code chunk extracts the first 10 columns that contain useful information for this study
Combining the Business Data with Geo Info
The code chunk appends the business sf data frame with location information.

Saving the data
The code chunk saves the data as rds format.

Converting to sf Data Frame
The code chunk converts the tibble data frame to an sf object and also assigns the correct EPSG.

```{r}
geocode_biz <- function(biz_df, out_rds_path) {
  # Extract unique postal codes
  postcodes <- unique(biz_df$postal_code)
  url <- "https://onemap.gov.sg/api/common/elastic/search"
  
  found <- data.frame()
  not_found <- data.frame(postcode = character())
  
  # Loop through postal codes
  for (pc in postcodes) {
    query <- list(
      searchVal = pc,
      returnGeom = "Y",
      getAddrDetails = "Y",
      pageNum = "1"
    )
    
    res <- GET(url, query = query)
    json <- content(res)
    
    if (json$found != 0) {
      df <- as.data.frame(json$results, stringsAsFactors = FALSE)
      df$input_postcode <- pc
      found <- bind_rows(found, df)
    } else {
      not_found <- bind_rows(not_found, data.frame(postcode = pc))
    }
  }
  
  # Keep first 10 columns
  found <- found %>% select(1:10)
  
  # Join back to business data
  biz_df <- biz_df %>%
    left_join(found, by = c("postal_code" = "POSTAL"))
  
  # Save RDS
  write_rds(biz_df, out_rds_path)
  
  # Convert to sf
  biz_sf <- st_as_sf(biz_df,
                     coords = c("X", "Y"),
                     crs = 3414)
  
  return(biz_sf)
}

```

Usage on select three codes

```{r}
biz_46900_sf <- geocode_biz(biz_46900, "data/rds/biz_46900.rds")
```


```{r}
biz_62011_sf <- geocode_biz(biz_62011, "data/rds/biz_62011.rds")
```

```{r}
biz_56140_sf <- geocode_biz(biz_56140, "data/rds/biz_56140.rds")
```




EDA
Visualising the Distribution
The code chunk shows the new business registration number by month.aes() maps data frame columns to plot aesthetics, e.g. aes(x = var1, y = var2) maps var1 to the x-axis and var2 to the y-axis.


```{r}
ggplot(data = biz_46900_sf,
       aes(x = MONTH_ABBR)) +
  geom_bar()
```









  
Visualsing the business
The code chunk plots the new business in interactive map.


```{r}

mpsz_lbl <- mpsz_cl %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6) %>%
  st_point_on_surface() %>%                 # safer than st_centroid for skinny polygons
  filter(area_km2 > 0.30)                   # label only big ones; tweak threshold

tmap_mode("view")

tm_shape(mpsz_cl) +
  tm_polygons(alpha = 0.15, border.col = "grey50") +
tm_shape(mpsz_lbl) +
  tm_text("SUBZONE_N", size = 0.6, col = "blue", auto.placement = TRUE) +
tm_shape(biz_46900_sf) + tm_dots()
```


```{r}
tmap_mode('view') 
tm_shape(biz_46900_sf) + tm_dots()
```
  

The code chunk reverses the tmap_mode from interactive to static.
```{r}
tmap_mode('plot')
```


EDA
Visualising the Distribution
The code chunk shows the new business registration number by month.aes() maps data frame columns to plot aesthetics, e.g. aes(x = var1, y = var2) maps var1 to the x-axis and var2 to the y-axis.


```{r}
ggplot(data = biz_62011_sf,
       aes(x = MONTH_ABBR)) +
  geom_bar()
```


  
Visualsing the business
The code chunk plots the new business in interactive map.


```{r}
tmap_mode('view') 
tm_shape(biz_62011_sf) + tm_dots()
```
  

The code chunk reverses the tmap_mode from interactive to static.
```{r}
tmap_mode('plot')
```


```{r}
ggplot(data = biz_56140_sf,
       aes(x = MONTH_ABBR)) +
  geom_bar()
```


  
Visualsing the business
The code chunk plots the new business in interactive map.


```{r}
tmap_mode('view') 
tm_shape(biz_56140_sf) + tm_dots()
```
  

The code chunk reverses the tmap_mode from interactive to static.
```{r}
tmap_mode('plot')
```








Online marketplaces for goods (including food)

```{r}
biz_63201 <- acra_2025 %>%
  filter(primary_ssic_code == 63201)
```

```{r}
biz_63205 <- acra_2025 %>%
  filter(primary_ssic_code == 63205) 
```


we realise that \`preimiarly ssic code has missing leading 0 probably due to transmission and registartion date date type is ymd.

Inside map_dfr(), vroom(.x) → reads each CSV filter(between(...)) → keeps only rows in the date window. map_dfr() → row-binds all filtered CSVs into one tibble (files_fl).











```{r}
dir <- "data/aspatial/rawdata" 
files <- dir_ls(dir, glob = "*.csv") #ls: list command, glob wild card
#one_file <- vroom(files[1]) #check one file 
#spec(one_file)
files_fl <- files %>%
  map_dfr(~ vroom(.x, show_col_types = FALSE) %>% #return data frames
          filter(between(registration_incorporation_date,
                        as.Date("2024-01-01"),
                        as.Date("2025-06-30"))))
```

quick review on the column types

```{r}
files_fl
spec(files_fl)
```

sanity check

```{r}
any(files_fl$registration_incorporation_date < ymd("2024-01-01") |files_fl$registration_incorporation_date > ymd("2025-06-30"))
any(duplicated(files_fl))
```

we realise that \`preimiarly ssic code has missing leading 0 probably due to nuemvic type. so we pad the 0 to the code and create new column where 2d is extracted per row.

```{r}
files_ssic <-files_fl%>%
  mutate(`ssic_5d` = str_pad(.[[22]], width = 5, side = "left", pad = "0")) %>%
  mutate(`ssic_2d` = str_sub(`ssic_5d`,1, 2))
```

it is time to rerragnet the columsn since there were 55 columsns after the ssic_2d, we select the following columns. uen, name, reg date, type, post code, and ssic 5d, ssic 2d

```{r}
files_sl <- files_ssic %>%
  select(c(1,3:4,6,8:9,17,54,55))
```

load the colasifiaction file, and skip top 4 columns to locate the correct the collumn. select the 2d only and its depscrition.

```{r}
ssic_lookup <- read_excel("data/aspatial/ssic2020-classification-structure.xlsx", skip = 4)%>%
  select(ssic_code = `SSIC 2020`, ssic_title = `SSIC 2020 Title`) %>%
  filter(!is.na(ssic_code)) %>%
  filter(str_length(ssic_code) == 2)
```

left join with main data.

```{r}
files_sl_jn <- files_sl%>%
  left_join(ssic_lookup, by = c("ssic_2d" = "ssic_code"))
```

check the number of registeration per business group.

```{r}
files_sl_ct<- files_sl_jn %>%
  group_by(`ssic_2d`,`ssic_title`) %>% 
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n))


top_10 <-head(files_sl_ct, 10)
top_10
```

plot a bar chart for the top 10 busienss type

```{r, fig.width=14}
top_10 %>%
  mutate(ssic_type = paste(ssic_2d,str_sub(ssic_title, 1,25), sep=":")) %>%
  ggplot(aes(x = reorder(ssic_2d, n), y = n, fill = ssic_type)) +
  geom_col() +
  #coord_filp() +
  labs(title =" Top 10 Business Registeration (1/1/2024- 30/6/2025)",
       x = "SSIC 2 Digits",
       y = "Number of Registeration",
       fill = "Business Type")+
  theme(plot.title = element_text(size = 16, face = "bold"))
```

```{r}
write.csv(files_sl_jn,"data/aspatial/files_sl_jn.csv")
```

```{r}
files_sl_jn <-vroom("data/aspatial/files_sl_jn.csv")
glimpse(files_sl_jn)
```

Narrow down to top 3–5 SSICs for deep-dive analysis SPPA/ST-SPPA
:::

## Geospatial Data Wrangling

first tidy up the postal code, some are missing lead "0"

```{r}
files_geo<-files_sl_jn%>%
  mutate(`address` = str_pad(postal_code, width = 6, side = "left", pad = "0"))

glimpse(files_geo)
```

```{r}
get_geocode_by_ssic <- function(df, ssic_code) {
  # 1) subset this SSIC
  sub <- df %>% filter(ssic_2d == ssic_code)

  # 2) unique, valid postcodes
  pc <- tibble(address = unique(sub$address)) 

  # 3) geocode (NOTE: column name passed as a STRING)
  geo <- geocode_onemap(pc, "address", return_geom = TRUE)

  # 4) join back
  sub %>% left_join(geo, by = "address") %>%
    st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326, remove = FALSE) %>%
    st_transform(crs=3414)
  
}


#fs_geo  <- get_geocode_by_ssic(files_geo, "64")  #FINANCIAL SERVICES
#glimspe()
#wst_geo <- get_geocode_by_ssic(files_geo, "46")  # WHOLESALE TRADE
#mcs_geo <- get_geocode_by_ssic(files_geo, "70") # MANAGEMENT CONSULTANCY ACTIVITIES
#rt_geo  <- get_geocode_by_ssic(files_geo, "47") #RETAIL TRADE
#cp_geo  <- get_geocode_by_ssic(files_geo, "62") #	COMPUTER PROGRAMMING
```

### Creating Owin Object

Similarly, the owin object can be created using the function as.owin() for polygon data. After the conversion, the class() and plot() functions can be used to verify that the object is of the correct class and that the data retains its original shape.

```{r}
sg_owin <-as.owin(mpsz_cl)
```

```{r}
class(sg_owin)
```

```{r}
plot(sg_owin)
```



### Converting sf Point Data Frames to PPP

Here we use `as.ppp()` of spatstat package to covert the point data `bsns_sf` to ppp file, confirm the change using `class()` and have a quick overview of the data statistics via `summary()`.when converting the sf to ppp, we include a nested rjitter() function to push away the overlapped data points.

```{r}
biz56140_ppp <- as.ppp(biz_56140_sf) %>%
  rjitter(retry = TRUE,
          nsim = 1,
          drop = TRUE)
class(biz56140_ppp)
```

```{r}
summary(biz56140_ppp)
```
```{r}
plot(unmark(biz56140_ppp), main = "biz56140")
```

```{r}
any(duplicated(biz56140_ppp))
```



### Combining Point Events object and Owin Object

The code chunk below combines ppp and owin into one ppp file which means it updates the window of childcare_ppp to sg_owin and keeps the points that fall inside.

```{r}
biz56140_PPP = biz56140_ppp[sg_owin]
class(biz56140_PPP)
```
```{r}
summary(biz56140_PPP)
```

```{r}
plot(unmark(biz56140_PPP), main = "biz56140_PPP")
```
## Clark-Evans Test for Nearest Neighbour Anaylysis (NNA)


Clark-Evans Test for Nearest Neighbour Anaylysis (NNA)

```{r}
clarkevans.test(biz56140_PPP,
                correction = "none",
                clipregion = "sg_owin",
                alternative= c("clustered")
                )
```


 Performing the Clark-Evans Test with CSR (Monte Carlo?)
 
```{r}
clarkevans.test(biz56140_PPP,
                correction ="none",
                clipregion ="sg_owin",
                alternative= "less",
                method = "MonteCarlo",
                nsim = 99)
```
 
 Kernel Density Estimation Method
 Working with Automatic Bandwidth Selection Method
 
 
```{r}
biz56140_diggle <- density (
  biz56140_PPP,
  sigma = bw.diggle,
  edge = TRUE,
  kernel = "gaussian")
```
 
 
 
```{r}
plot(biz56140_diggle)
```
 
```{r}
summary(biz56140_diggle)

```
 
 
```{r}
bw <- bw.diggle(biz56140_PPP)
bw
```
```{r}
biz56140_ppp_km <- rescale.ppp(
  biz56140_PPP, 1000, "km")
summary(biz56140_ppp_km)
```
 
```{r}
kde_biz56140_ppp_km <- density(biz56140_ppp_km,
                              sigma = bw.diggle,
                              edge = TRUE,
                              kernel = "gaussian")
summary(kde_biz56140_ppp_km)
```
 
 
```{r, fig.height=6}
plot(kde_biz56140_ppp_km)
```
 

## Compute spatial and spatio-temporal KDE of the selected business types.

## second order spatial and spatio-temporal analysis on the selected business types.

## Describe and discuss the analysis results obtained from first and second orders of SPPA & ST SPPA.

## short report of not more than 150 words describing the spatio-temporal patterns revealed by the geovisualisation and statistical graphics.

## discuss how these findings can be used to support urban land use planning and management. (Not more than 500 words)

Check local policy during 2024–2025:

e.g., Singapore introduced incentives for fintech or retail startups?

New sustainability/green finance schemes?

Post-COVID recovery grants?

Why: If a sudden spike in registrations in certain SSIC groups coincides with a government initiative,
